---
title: "crossLagR: Simulation and Basic Model Syntax"
author: "Chris Weber"
date: "2025-02-13"
output: pdf_document
---
y
The $\texttt{crossLagR}$ package includes several functions to simulate data.

$\texttt{simRICLPM()}$ simulates data from a random intercept cross-lagged panel model.

$\texttt{simCLPM()}$ simulates data from a cross-lagged panel model.

$\texttt{reshape\_long\_sim\_cr()}$ Takes a wide data frame, pivots to wide format, includes lags.


```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(crossLagR)
library(dplyr)

df = simRICLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                variance.between.x = 0.5,
                variance.between.y = 0.5,
                cov.between = 0,
                sample.nobs = 10e2)
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long
head(dat_long)
```

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(crossLagR)
df = simCLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                sample.nobs = 10e2)
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long
head(dat_long)
```

## Models

$\texttt{estCLPM()}$ Constructs the lavaan code to estimate a CLPM. $\texttt{estRICLPM()}$ Constructs the lavaan code to estimate a RI-CLPM. $\texttt{estCTSEM()}$ Estimates a continuous time structural equation model. $\texttt{estHLM()}$ Estimates the HLM.

### Cross-Lagged Panel Model

Let's fit several models under a random intercept cross lagged panel data generating process.

```{r}
library(crossLagR)
library(lavaan)
# Simulate Data Under RI CLPM
df = simRICLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                variance.between.x = 0.5,
                variance.between.y = 0.5,
                cov.between = 0,
                sample.nobs = 10e2)
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long

fit1 <- lavaan(estimateCLPM(waves = 5),
              data = dat,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml")

summary(fit1)
```

### Random Intercept Cross-Lagged Panel Model

```{r}
library(crossLagR)
library(lavaan)

fit2 <- lavaan(estimateRICLPM(waves = 5),
              data = dat,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml")

summary(fit2)
```

The CLPM clearly does not fit the data well, and provides rather inaccurate estimates.

### Hiearchical Linear Model, Random Intercepts and Slopes

```{r, echo = FALSE, message = FALSE, warning = FALSE}
model = estimateHLM(data = dat_long, residual_correlation = FALSE)
summary(model)
```

### Continuous Time Model

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dat_long %>%
  select(id, wave, x, y) -> ct_sem_data
# ctsem package Follows this convention:
names(ct_sem_data) = c("id", "time", "x", "y")
model = estimateCTSEM(data = ct_sem_data)
# [1] Returns a list; Extract cstem model
# [2] Return a discrete time CL and AR model
summary(model[[1]])$parmatrices
print(model[[2]] %>% t())
```
