---
title: "ISCAP panel: party identification ↔ ideology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ISCAP panel: party identification ↔ ideology}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

This vignette walks through a minimal cross-lagged analysis of the ISCAP panel
[@upenn_iscap_2022]: do shifts in party identification predict shifts in
ideological self-placement, or the reverse? See the
[user guide](https://crweber9874.github.io/crossLagR/book/06_iscap.html)
for the full treatment.

## Setup

```{r setup}
library(crossLagR)
library(lavaan)
library(dplyr)
library(tidyr)

data(upenn)
```

The ISCAP panel includes three waves we'll use here (2012, 2016, 2018). The
relevant columns are `party_identification_*` (7-point, 7 = Strong Republican)
and `ideology_*` (7-point, 7 = Extremely Conservative).

## Wide data

```{r}
wide <- upenn |>
  transmute(
    x1 = party_identification_2012,
    x2 = party_identification_2016,
    x3 = party_identification_2018,
    y1 = ideology_2012,
    y2 = ideology_2016,
    y3 = ideology_2018
  ) |>
  drop_na()

n_obs  <- nrow(wide)
n_miss <- sum(is.na(upenn[, c("party_identification_2012",
                              "party_identification_2016",
                              "party_identification_2018",
                              "ideology_2012",
                              "ideology_2016",
                              "ideology_2018")]))

cat("Complete cases used:", n_obs, " (", n_miss, "missing cells in source)\n")
```

## CLPM

`estimateCLPM()` returns lavaan model syntax with the conventional `beta_*`
(autoregressive) and `omega_*` (cross-lag) labels.

```{r}
clpm_fit <- lavaan(estimateCLPM(waves = 3), data = wide)
summary(clpm_fit, standardized = TRUE)
```

## RI-CLPM

`estimateRICLPM()` adds random intercepts $I_x, I_y$ that absorb stable
between-person differences. The AR and CL coefficients are now estimated
on the within-person deviations.

```{r}
riclpm_fit <- lavaan(estimateRICLPM(waves = 3), data = wide)
summary(riclpm_fit, standardized = TRUE)
```

## Reading the contrast

A pattern that recurs in the book:

- The CLPM AR coefficients are inflated by stable between-person trait
  variance and the CL coefficients tend to be conservatively biased.
- The RI-CLPM AR coefficients are smaller (within-person carry-over only).
  CL coefficients show whether **change** in one variable forecasts
  **change** in the other.

For the ISCAP data the cross-lags are modest in either specification,
with party identification appearing to lead ideology slightly more than
the reverse. See [the book chapter](https://crweber9874.github.io/crossLagR/book/06_iscap.html)
for the full discussion including a system-legitimacy block, latent-change
specifications, and Bayesian fits via `brms`.

## References
