---
title: "LISS panel: personality ↔ political orientation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LISS panel: personality ↔ political orientation}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  eval     = FALSE
)
```

This vignette is an abbreviated partial replication of @bakker_reconsidering_2021,
using the LISS panel data ship with `crossLagR`. We ask whether the
"closed-personality" composite (low openness + high conscientiousness) leads
right-leaning political orientation (left–right self-placement), or vice versa.
See the [user guide](https://crweber9874.github.io/crossLagR/book/07_liss.html)
for the full chapter, which also covers each Big-Five trait, all five
political items, and several alternative specifications.

## Setup

```{r setup}
library(crossLagR)
library(lavaan)
library(dplyr)
library(tidyr)

data(liss_long)
```

`liss_long` is in long form with one row per (respondent × wave). For the
CLPM we need three waves of two observed variables in wide form. Bakker et al.
use the May 2008, 2013, and 2019 personality waves. Here we use the same
three waves and pair personality with the left–right item from the
preceding December.

## Construct the composites

The closed-personality composite is the mean of reverse-scored openness items
plus conscientiousness items. We compute a simple per-wave mean ignoring the
factor model (the book chapter fits a measurement model instead).

```{r}
open_items <- c("rich_vocab", "dffclt_undrstnd", "vivid_imagin", "no_interest",
                "excellent_ideas", "poor_imagin", "quick_undrstnd",
                "dffclt_words", "reflecting", "full_of_ideas")

cons_items <- c("always_prepared", "leave_belongings", "pay_attention",
                "make_mess", "chores_done", "forget_to_put", "like_order",
                "shirk_duties", "follow_sched", "exacting")

# Items reverse-scored in the LISS Big-Five battery (5-pt scale).
open_rev <- c("dffclt_undrstnd", "no_interest", "poor_imagin",
              "dffclt_words")
cons_rev <- c("leave_belongings", "make_mess", "forget_to_put",
              "shirk_duties")

reverse5 <- function(x) 6 - x

liss <- liss_long |>
  mutate(across(all_of(open_rev), reverse5),
         across(all_of(cons_rev), reverse5)) |>
  rowwise() |>
  mutate(
    openness_high      = mean(c_across(all_of(open_items)), na.rm = TRUE),
    conscientiousness  = mean(c_across(all_of(cons_items)), na.rm = TRUE),
    closed_personality = mean(c(6 - openness_high, conscientiousness),
                              na.rm = TRUE)
  ) |>
  ungroup()
```

## Reshape to wide (3 waves: 2008, 2013, 2019)

```{r}
target_waves <- c(8, 13, 19)

wide <- liss |>
  filter(wave %in% target_waves) |>
  mutate(wave_label = match(wave, target_waves)) |>
  select(case_id, wave_label, closed_personality, left_right) |>
  pivot_wider(names_from  = wave_label,
              values_from = c(closed_personality, left_right),
              names_glue  = "{.value}{wave_label}") |>
  transmute(
    x1 = closed_personality1, x2 = closed_personality2, x3 = closed_personality3,
    y1 = left_right1,         y2 = left_right2,         y3 = left_right3
  ) |>
  drop_na()

cat("Complete cases:", nrow(wide), "\n")
```

## CLPM

```{r}
clpm_fit <- lavaan(estimateCLPM(waves = 3), data = wide)
summary(clpm_fit, standardized = TRUE)
```

## RI-CLPM

```{r}
riclpm_fit <- lavaan(estimateRICLPM(waves = 3), data = wide)
summary(riclpm_fit, standardized = TRUE)
```

## Reading the contrast

Following @bakker_reconsidering_2021, the CLPM cross-lags suggest a
modest reciprocal coupling between closed personality and right-leaning
self-placement. The RI-CLPM, which absorbs stable between-person trait
differences into the random intercepts $I_x$ and $I_y$, typically reports
much smaller within-person cross-lags — consistent with the idea that the
two constructs are correlated *traits* more than wave-to-wave drivers of one
another.

For the per-trait breakdown (all Big-Five facets × five political items),
alternative specifications, and a discussion of the substantive
implications, see [the book chapter](https://crweber9874.github.io/crossLagR/book/07_liss.html).

## References
