---
title: "State and Trait Level Effects"
header-includes:
    - \usepackage{setspace}\doublespacing
author: "Chris Weber"
date: "2025-01-21"
indent: true
output:
  pdf_document: default
---

# Introduction 

The CLPM follows this structure.

$$\begin{bmatrix}
 y_{it}\\
 x_{it}
\end{bmatrix}
=
\begin{bmatrix}
\theta_{1y} & \theta_{1x}  \\
\theta_{2y} & \theta_{2x}  
\end{bmatrix}
\begin{bmatrix}
y_{it-1} \\ x_{t-1}  
\end{bmatrix}
-   
\begin{bmatrix}
e_{y,it}\\
e_{x,it}
\end{bmatrix}
$$

The diagonal elements in the $\theta$ matrix represent the autoregressive parameters; the off-diagonal elements are the cross-lagged parameters.

To model **trait** and **state** effects, the **Random Intercept Cross Lagged** model follows the form, but includes a time invariant "trait" level parameter ($\mu_i$).

$$\begin{bmatrix}
 y_{it}\\
 x_{it}
\end{bmatrix}
=
\underbrace{
\begin{bmatrix}
\theta_{1y} & \theta_{1x}  \\
\theta_{2y} & \theta_{2x}  
\end{bmatrix}
\begin{bmatrix}
y_{it-1} \\ x_{t-1}  
\end{bmatrix}
+
\begin{bmatrix}
e_{y,it}\\
e_{x,it}
\end{bmatrix}
}_{State}
+
\underbrace{
\begin{bmatrix}
\mu_{y,i}\\
\mu_{x,i}
\end{bmatrix}
}_{Trait}
$$

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
monteCarloOLS(
   trials = 1000,
   waves = 6,
   model_type = "clpm",
   stability.p = 0.45,
   stability.q = 0.45,
   cross.p = 0.0,
   cross.q = 0.0,
   variance.q = 0.3,
   variance.p = 0.3,
   cov.between = 0,
   sample_size = 2500) -> results

results %>% summarize(xlagx = mean(xlag_x),
                      xlagy = mean(xlag_y),
                      ylagx = mean(ylag_x),
                      ylagy = mean(ylag_y)) %>% t()

```

The OLS estimates are nearly identical to the true values. Here's the way to simulate the RI-CLPM.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
monteCarloOLS(
   trials = 100,
   waves = 6,
   model_type = "riclpm",
   stability.p = 0.45,
   stability.q = 0.45,
   cross.p = 0.0,
   cross.q = 0.0,
   variance.q = 0.3,
   variance.p = 0.3,
   cov.between = 0.3,
   variance.between.x = 0.2,
   variance.between.y = 1,
   sample_size = 2500) -> results

results %>% summarize(xlagx = mean(xlag_x),
                      xlagy = mean(xlag_y),
                      ylagx = mean(ylag_x),
                      ylagy = mean(ylag_y)) %>% t()

```

Things generally fall apart when there are trait effects.


```{r}
between  = seq(0.5, 1, by = .1)
for(x in between){
            results <- rbind(results, 
                             monteCarloOLS(
                                               trials = 500,
                                               waves = 6,
                                               model_type = "riclpm",
                                               stability.p = 0.3,
                                               stability.q = 0.3,
                                               cross.p = 0,
                                               cross.q = 0,
                                               variance.q = 0.5,
                                               variance.p = 0.5,
                                               variance.between.x = x,
                                               variance.between.y = 0.5,
                                               cov.between = 0.35,
                                               sample_size = 2500)
                       )
  }
results
```

I've specified a fixed random intercept variance of 0.5, varied the random intercept variance of the $y$, a covariance btween errors, a 0.3 AR parameter, and a fixed, "within" variance of 0.5. 


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(latex2exp)
library(ggridges)
results %>%
  ggplot(aes(x = xlag_x, fill = "grey", y = as.factor(round(variance.between.x, 2)))) +
  geom_density_ridges2(
    alpha = 0.25,
    scale = 1,
    show.legend = FALSE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "darkgrey",
    point_alpha = 0.01,
    fill = "darkgrey",
    color = NA  # Remove the outline

  ) +
  # scale_x_continuous("", limits = c(-0.25, 0.25)) +
  scale_y_discrete("Random Intercept Variance (X)") +
  scale_x_continuous(TeX(" b_2 ")) +
  theme_minimal()  +
  # Draw a vertical line at 0 that is dashed, use abline
  geom_vline(xintercept = .4, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.4, y = 1, label = "True Value ", hjust = 0) +
  labs(title = "Autoregressive Parameter Estimates\nVarying Random Intercept Variance (X)", subtitle = TeX("$x_{t, i}=a_0 + b_1 y_{t-1, i} + b_2 x_{t-1, i} + e_{2,i}$"))+
  # Style the plot
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 12)
  )

```


