---
title: "State and Trait Level Effects"
header-includes:
    - \usepackage{setspace}\onehalfspacing
author: "Chris Weber"
date: "2025-01-21"
indent: true
output:
  pdf_document: default
---

# Introduction

The CLPM follows this structure.

$$\begin{bmatrix}
 y_{it}\\
 x_{it}
\end{bmatrix}
=
\begin{bmatrix}
\theta_{1y} & \theta_{1x}  \\
\theta_{2y} & \theta_{2x}  
\end{bmatrix}
\begin{bmatrix}
y_{it-1} \\ x_{t-1}  
\end{bmatrix}
-   
\begin{bmatrix}
e_{y,it}\\
e_{x,it}
\end{bmatrix}
$$

The diagonal elements in the $\theta$ matrix represent the autoregressive parameters; the off-diagonal elements are the cross-lagged parameters.

To model **trait** and **state** effects, the **Random Intercept Cross Lagged** model follows the form, but includes a time invariant "trait" level parameter ($\mu_i$).

$$\begin{bmatrix}
 y_{it}\\
 x_{it}
\end{bmatrix}
=
\underbrace{
\begin{bmatrix}
\theta_{1y} & \theta_{1x}  \\
\theta_{2y} & \theta_{2x}  
\end{bmatrix}
\begin{bmatrix}
y_{it-1} \\ x_{t-1}  
\end{bmatrix}
+
\begin{bmatrix}
e_{y,it}\\
e_{x,it}
\end{bmatrix}
}_{State}
+
\underbrace{
\begin{bmatrix}
\mu_{y,i}\\
\mu_{x,i}
\end{bmatrix}
}_{Trait}
$$

The parameters used to generate the data are as follows:

There are 10 waves. The autoregressive parameters are 0.10, and there is no cross-lagged effect; these parameters are set to 0. The variances of the latent variabels are 1, and the sample size is 2500. Here's how that looks in the function $\texttt{monteCarloOLS()}$.

This function simulates the data using these parameters, then estimates the CLPM in OLS. Currently, because we're dealing with lags, the first observation is missing. I've left it as missing, meaning that it is dropped from the analysis, though typically we can set this at some value. It repeats this process 100 times.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(crossLagR)
library(chattr)

# Sys.setenv("OPENAI_API_KEY" = "")
# The key will not be shown in the repository, just your local .renv environment.

library(ggplot2)
monteCarloOLS(
   trials = 100,
   waves = 10,
   dgp = "clpm",
   stability.p = 0.1,
   stability.q = 0.1,
   cross.p = 0.0,
   cross.q = 0.0,
   variance.q = 1,
   variance.p = 1,
   cov.between = 0,
   sample_size = 2500) -> results

results %>% summarize(x_autoregression = mean(xlag_x),
                      x_cross_lag = mean(xlag_y),
                      y_autoregression = mean(ylag_y),
                      y_cross_lag = mean(ylag_x)) %>% t()

```

The means are summarized above. We could experiment with different parameter values, by creating a data frame corresponding to the parameters the user can choose to vary.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
# Construct the parameter data set by using expand.grid to generate all permutations
# Vary AR parameters from 0 to 1 byincrements of 0.5
monteCarloOLS(
   trials = 1,
   waves = 5,
   dgp = "clpm",
   stability.p = 0,
   stability.q = 0.50,
   cross.p = 0.0,
   cross.q = 0.0,
   variance.q = 1,
   variance.p = 1,
   cov.between = 0,
   sample_size = 5000) -> results

expand.grid(
  stability.p = seq(0, 1, by = .2),
  stability.q = 0.5,
  cross.p = 0,
  cross.q = 0,
  variance.q = 1,
  variance.p = 1,
  cov.between = 0,
  sample_size = 5000) -> mc_params


for(i in 1:nrow(mc_params)){
  # print(paste("🚀 Running Monte Carlo Simulation 🚀"))
  # print(paste("Within Person Stability X: ", mc_params$stability.p[i], " | Within Person Stability Y: ", mc_params$stability.q[i]))
  # print(paste("Cross Lagged Effect X: ", mc_params$cross.p[i], " | Cross Lagged Effect Y: ", mc_params$cross.q[i]))
  # print(paste("Variance X: ", mc_params$variance.p[i], " | Variance Y: ", mc_params$variance.q[i]))
  # print(paste("Covariance Between X and Y: ", mc_params$cov.between[i]))
  # print(paste("Sample Size: ", mc_params$sample_size[i]))
  results <- rbind(results, 
                   monteCarloOLS(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "clpm",
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross.p = mc_params$cross.p[i],
                                   cross.q = mc_params$cross.q[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.between = mc_params$cov.between[i],
                                   sample_size = mc_params$sample_size[i])
                )

}

resultsOLS = results
save(resultsOLS, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS.rda")
```

To summarize the data in joyplots, use the $\texttt{ggridges}$ package.

```{r}
library(ggplot2)
library(ggridges)
library(latex2exp)
library(crossLagR)
load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS.rda")

vline_data <- data.frame(
  variable = c("X_autoregression", "Y_autoregression", "X_cross_lag", "Y_cross_lag"),
  intercept = c(NA, 0.5, 0, 0)  # Replace with your desired intercept values
)

custom_labels <- c(
  X_autoregression = "X (AR)",
  Y_autoregression = "Y (AR)",
  X_cross_lag = "X (CL)",
  Y_cross_lag = "Y (CL)"
)

results %>%
  filter(stability.p != 0.1) %>%
  rename(X_autoregression = xlag_x,
         Y_cross_lag =  ylag_x ,
         X_cross_lag = xlag_y,
         Y_autoregression = ylag_y) %>%
    pivot_longer(
        cols = c(X_autoregression, Y_autoregression, X_cross_lag, Y_cross_lag),
        names_to = "variable",
        values_to = "value"
    ) %>%
  ggplot(aes(x = value, 
             fill = "grey", 
             y = as.factor(stability.p)))+
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = custom_labels)) + 
  geom_density_ridges2(
    alpha = 0.20,
    scale = 1,
    show.legend = FALSE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = c(0.5, 0.975),
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "darkgrey",
    point_alpha = 0.01,
    fill = "darkgrey",
    color = NA  # Remove the outline
  ) +
  theme_minimal() + 
  # title
  
  labs(title = paste("Cross Lagged Panel Model Estimates"),
  subtitle = TeX("$x_{t, i}=\\theta_{12} y_{t-1, i} +  \\theta_{11} x_{t-1, i} + e_{1,i}$\n
                  $y_{t, i}=\\theta_{22} y_{t-1, i} +  \\theta_{21} x_{t-1, i} + e_{2,i}$
                 " ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(data = vline_data, aes(xintercept = intercept), linetype = "dashed", color = "black") +
  
  scale_x_continuous("Parameter Estimate (1,000 Trials)") +
  scale_y_discrete("")  

# Save graph
ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/ols_plot.pdf", width = 8, height = 6)
```

Things generally fall apart when there are trait effects.



```{r}
library(tidyr)
library(dplyr)
devtools::load_all()
monteCarloOLS(
  waves = 5,
  variance.between.x = 0.5,
  variance.between.y = 0.5,
  stability.p = 0.2,
  stability.q = 0.2,
  cross.p = 0,
  cross.q = 0,
  variance.q = 0.5,
  variance.p = 0.5,
  cov.between = 0,
  sample_size = 5000) -> results

# Construct the parameter data set by using expand.grid to generate all permutations
# Vary AR parameters from 0 to 1 byincrements of 0.5
expand.grid(
  variance.between.x = seq(0.5, 1.5, by = .2),
  variance.between.y = seq(0.5, 1.5, by = .2),
  stability.p = 0.2,
  stability.q = 0.2,
  cross.p = 0,
  cross.q = 0,
  variance.q = 0.5,
  variance.p = 0.5,
  cov.between = 0,
  sample_size = 5000) -> mc_params

# Generally don't do this -- append lists is inefficient.
for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloOLS(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross.p = mc_params$cross.p[i],
                                   cross.q = mc_params$cross.q[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.between = mc_params$cov.between[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
resultsOLS2 = results
save(resultsOLS2, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS2.rda")

```

And here is the plot. The plot shows the effects for the equation $x_{t, i}=a_0 + b_1 y_{t-1, i} + b_2 x_{t-1, i} + e_{2,i}$, where $b_1$ is the autoregressive parameter, and $b_2$ is the cross-lagged parameter. The individual plots show the consequences when the trait-level, between unit variance of $y$ is varied. The axis-labels similarly show the consequences when the trait-level, between unit variance of $x$ is varied. The model generally fails in picking up the true parameter estimates for the CL and AR values.

```{r}
library(ggridges)
library(latex2exp)
library(dplyr)
library(ggplot2)
library(tidyr)
devtools::load_all()

load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS2.rda")
resultsOLS2 %>%
  pivot_longer(
    cols = c(xlag_x, ylag_x, xlag_y, ylag_y),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) %>%
  mutate(variable = factor(variable, levels = c("xlag", "ylag"))) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "lag_type",
    values_to = "lag_value"
  ) %>%
  filter(lag_type == "x") %>%
  ggplot(aes(x = lag_value, 
             fill = variable, 
             grouping = variable,
             color = variable,
             y = as.factor(variance.between.x)))+
  facet_wrap(~paste("Var(Y) = ", variance.between.y), scales = "free", ncol = 2) +
  geom_density_ridges2(
    alpha = 1,
    scale = 3,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 2,
    point_size = 0.1,
    point_color = "grey",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) + 
  theme_minimal() + 
  # title
  labs(title = "CLPM Parameter Estimates (x)", 
  subtitle = TeX("$\\textbf{x_{t, i}=\\theta_{12} y_{t-1, i} +  \\theta_{11} x_{t-1, i} + e_{1,i}}$\n
                  $y_{t, i}=\\theta_{22} y_{t-1, i} +  \\theta_{21} x_{t-1, i} + e_{2,i}$")) + 
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Parameter Estimate", limits = c(0,1)) +
  scale_y_discrete("Trait Variance (X)") +
  # color
scale_fill_manual(values = c("xlag" = "darkgrey", "ylag" = "lightgrey"),
                  labels = c("xlag" = "Autoregressive Parameter", "ylag" = "Cross Lagged Parameter")) +
  # legend position
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        legend.key.width = unit(0.2, "cm"))


ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/ols2_plot.pdf", width = 8, height = 6)
```


```{r}
library(ggridges)
library(latex2exp)

load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS2.rda")
resultsOLS2 %>%
  pivot_longer(
    cols = c(xlag_x, ylag_x, xlag_y, ylag_y),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) %>%
  mutate(variable = factor(variable, levels = c("xlag", "ylag"))) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "lag_type",
    values_to = "lag_value"
  ) %>%
  filter(lag_type == "y") %>%
  ggplot(aes(x = lag_value, 
             fill = variable, 
             grouping = variable,
             color = variable,
             y = as.factor(variance.between.x)))+
  facet_wrap(~paste("Var(Y) = ", variance.between.y), scales = "free", ncol = 2) +
  geom_density_ridges2(
    alpha = 1,
    scale = 3,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 2,
    point_size = 0.1,
    point_color = "grey",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) + 
  theme_minimal() + 
  # title
  labs(title = "CLPM Parameter Estimates (y)", 
  subtitle = TeX("$x_{t, i}=\\theta_{12} y_{t-1, i} +  \\theta_{11} x_{t-1, i} + e_{1,i}$\n
                  $\\textbf{y_{t, i}=\\theta_{22} y_{t-1, i} +  \\theta_{21} x_{t-1, i} + e_{2,i}}$")) + 
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Parameter Estimate", limits = c(0,1)) +
  scale_y_discrete("Trait Variance (X)") +
  # color
scale_fill_manual(values = c("xlag" = "darkgrey", "ylag" = "lightgrey"),
                  labels = c("xlag" = "Autoregressive Parameter", "ylag" = "Cross Lagged Parameter")) +
  # legend position
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        legend.key.width = unit(0.2, "cm"))


ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/ols3_plot.pdf", width = 8, height = 6)
```
# Create an animated plot


```{r}
library(gganimate)
library(ggridges)
library(latex2exp)
library(dplyr)

load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS2.rda")
resultsOLS2 %>%
  pivot_longer(
    cols = c(xlag_x, ylag_x, xlag_y, ylag_y),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) %>%
  mutate(variable = factor(variable, levels = c("xlag", "ylag"))) %>%
  pivot_longer(
    cols = c(x, y),
    names_to =  "lag_type",
    values_to = "lag_value"
  ) %>%
  filter(lag_type == "y") %>%
  ggplot(aes(x = lag_value, 
             fill = variable, 
             grouping = variable,
             color = variable,
             y = as.factor(variance.between.x)))+
  facet_wrap(~paste("Var(Y) = ", variance.between.y), scales = "free", ncol = 2) +
  geom_density_ridges2(
    alpha = 1,
    scale = 3,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 2,
    point_size = 0.1,
    point_color = "grey",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) + 
  theme_minimal() + 
  # title
  labs(title = "CLPM Parameter Estimates (y)", 
  subtitle = TeX("$x_{t, i}=\\theta_{12} y_{t-1, i} +  \\theta_{11} x_{t-1, i} + e_{1,i}$\n
                  $\\textbf{y_{t, i}=\\theta_{22} y_{t-1, i} +  \\theta_{21} x_{t-1, i} + e_{2,i}}$")) + 
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Parameter Estimate", limits = c(0,1)) +
  scale_y_discrete("Trait Variance (X)") +
  # color
scale_fill_manual(values = c("xlag" = "darkgrey",
                             "ylag" = "lightgrey"),
                  labels = c("xlag" = "Autoregressive Parameter", 
                             "ylag" = "Cross Lagged Parameter")) +
  # legend position
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        legend.key.width = unit(0.2, "cm"))
```




```{r}
library(ggridges)
library(latex2exp)

load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS2.rda")
resultsOLS2 %>%
  pivot_longer(
    cols = c(xlag_x, ylag_x, xlag_y, ylag_y),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) %>%
  mutate(variable = factor(variable, levels = c("xlag", "ylag"))) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "lag_type",
    values_to = "lag_value"
  ) %>%
  filter(lag_type == "y") %>%
  ggplot(aes(x = lag_value, 
             fill = "grey", 
             y = as.factor(variance.between.x)))+
  facet_wrap(~paste("Var(Y) = ", variance.between.y), scales = "free", ncol = 3) +
  geom_density_ridges2(
    alpha = 0.25,
    scale = 1,
    show.legend = FALSE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 2,
    point_size = 1,
    point_color = "darkgrey",
    point_alpha = 0.01,
    fill = " darkgrey",
    color = NA  # Remove the outline
  ) + 
  theme_minimal() + 
  # title
  labs(title = "CLPM Parameter Estimates (y)", 
  subtitle = TeX("$x_{t, i}=\\theta_{12} y_{t-1, i} +  \\theta_{11} x_{t-1, i} + e_{1,i}$\n
                  $\\textbf{y_{t, i}=\\theta_{22} y_{t-1, i} +  \\theta_{21} x_{t-1, i} + e_{2,i}}$")) + 
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.5, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Parameter Estimate") +
  scale_y_discrete("Trait Variance (X)") 

ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/ols3_plot.pdf", width = 8, height = 6)
```



```{r}
# Pivot the xlag and ylag variables

```

## Summary

The OLS model only picks up the correct parameter estimates -- i.e., the values used to simulate the data, when there is no unit level variance. We estimate the following equation.

$$x_{t, i}=a_0 + b_1 y_{t-1, i} + b_2 x_{t-1, i} + e_{2,i}$$

Yet, we assume the DGP looks like this

$$x_{t, i}=a_0 + b_1 y_{t-1, i} + b_2 x_{t-1, i} + \mu_i + e_{2,i}$$

Where $var(\mu_i) \sim (\bar{\mu_i}, \sigma_{\mu_i})$

The extent to which $\sigma_{\mu_i}$ is greater than zero -- the trait level variance -- the CLPM, because it is unable to accomodate both trait and state level effects, fails to estimate the correct parameter values, for both CL and AR paramters.

## The RI-CLPM

The RI-CLPM is a model that includes both trait and state level effects. The $\texttt{monteCarloRICLPM()}$ function can simulate data under this model.

```{r}
rm(list = ls())
library(tidyr)
library(dplyr)
devtools::load_all()
monteCarloOLS(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.8,
   cov.pq = 0,
   sample_size = 5000) -> results


# Construct the parameter data set by using expand.grid to generate all permutations
# Vary AR parameters from 0 to 1 byincrements of 0.5
expand.grid(
  stability.p = 0.2,
  stability.q = 0.2,
  variance.between.x = seq(0.8, 1.6, by = .2),
  variance.between.y = 0.5,
  cross.p = 0,
  cross.q = 0,
  variance.q = 1,
  variance.p = 1,
  cov.between = 0,
  sample_size = 5000) -> mc_params

# Generally don't do this -- append lists is inefficient.
for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloOLS(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross.p = mc_params$cross.p[i],
                                   cross.q = mc_params$cross.q[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.between = mc_params$cov.between[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
resultsMCOLS = results
save(resultsMCOLS, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLS2.rda")

```

```{r}
rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
library(lavaan)
monteCarloRICLPM(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.6,
   cov.pq = 0,
   sample_size = 5000) -> results

# Construct the parameter data set by using expand.grid to generate all permutations
# Vary AR parameters from 0 to 1 byincrements of 0.5
expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = 0.6,
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params

#
for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloRICLPM(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
resultsRICLPM = results
save(resultsRICLPM, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsRICLPM.rda")
```


```{r}
rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
library(lavaan)
monteCarloFI(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.6,
   cov.pq = 0,
   sample_size = 5000) -> results

# Construct the parameter data set by using expand.grid to generate all permutations
# Vary AR parameters from 0 to 1 byincrements of 0.5
expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = 0.6,
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params

#
for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloFI(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
resultsOLSFI = results
save(resultsOLSFI, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLSFI.rda")
```


```{r}
library(lavaan)
monteCarloCLPM(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   variance.between.x = 0.5,
   variance.between.y = 0.6,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.5,
   variance.p = 0.5,
   cov.pq = 0,
   sample_size = 5000) -> results

# Vary AR parameters from 0 to 1 byincrements of 0.5
expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = 0.6,
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params

#
for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloCLPM(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
resultsCLPM = results
save(resultsCLPM, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsCLPM.RData")

```



```{r, echo = FALSE, message = FALSE, warning = FALSE}
rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
library(lavaan)
monteCarloCTSEM(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.8,
   cov.pq = 0,
   sample_size = 5000) -> results


expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = 0.6,
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params

for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloCTSEM(
                                   trials = 25,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}

resultsCTSEM = results
save(resultsCTSEM, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsCTSEM.RData")

```



```{r}
rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
library(lavaan)
monteCarloLChange(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.6,
   cov.pq = 0,
   sample_size = 5000) -> results


expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = 0.6,
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params


for(i in 1:nrow(mc_params)){
  print(paste("🚀 Running Monte Carlo Simulation 🚀"))
  print(paste("Within Person Stability X: ", mc_params$variance.between.x[i], " | Within Person Stability Y: ", mc_params$variance.between.y[i]))
  results <- rbind(results, 
                   monteCarloLChange(
                                   trials = 100,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}

resultsLCHANGE= results
save(resultsLCHANGE, file = "~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsLCHANGE.rda")
```

```{r}
rm(list = ls())
library(dplyr)
library(tidyr)
library(dplyr)
library(crossLagR)

load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsCLPM.RData")
resultsCLPM <- resultsCLPM %>% mutate(model = "CLPM")
load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsRICLPM.rda")
resultsRICLPM <- resultsRICLPM %>% mutate(model = "RICLPM")
load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsCTSEM.RData")
resultsCTSEM <- resultsCTSEM %>% mutate(model = "CTSEM")
# load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsLCHANGE.rdat")
# resultsLCHANGE <- resultsLCHANGE %>% mutate(model = "LCHANGE")
load("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/resultsOLSFI.rda")
resultsOLSFI <- resultsOLSFI %>% mutate(model = "OLSFI")



bind_rows(resultsCLPM, 
          resultsRICLPM, 
          resultsCTSEM, 
          resultsOLSFI) -> resultsSIM

```

```{r}

```

```{r}
library(ggplot2)
library(ggridges)
library(latex2exp)
  
ylag_x <- ggplot(
  # Data 
  resultsSIM %>% filter(variance.between.x!=0.5) %>% 
         #rename model Model
         mutate(Model = as.factor(model)), 
  # Structure  
    aes(x = ylag_x, 
        y = as.factor(variance.between.x),
        grouping = as.factor(Model),
        color = Model,
        fill = Model,
            )) + 
  geom_density_ridges( alpha = 0.25,
    scale = 1,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "black",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) +
  theme_minimal() + 
  # title
  labs(title = paste("Cross Lagged Effects (Y)"),
  subtitle = TeX("$x_{t, i}= \\theta_{11}  x_{t-1, i} +  \\theta_{12} y_{t-1, i} + \\mu_{x,i} + e_{1,i}$\n
                  $y_{t, i}= \\theta_{22}  y_{t-1, i} +  \\theta_{21}  y_{x-1, i} + \\mu_{y,i} + e_{1,i}$" ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  # geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
 # annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous(TeX("$var(\\mu_{x,i})$ "), limits = c(-1,1)) +
  scale_y_discrete("") 

ylag_y <- ggplot(
  # Data 
  resultsSIM %>% filter(variance.between.x!=0.5) %>% 
         #rename model Model
         mutate(Model = as.factor(model)), 
  # Structure  
    aes(x = ylag_y, 
        y = as.factor(variance.between.x),
        grouping = as.factor(Model),
        color = Model,
        fill = Model,
            )) + 
  geom_density_ridges( alpha = 0.25,
    scale = 1,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "black",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) +
  theme_minimal() + 
  # title
  labs(title = paste("Autoregressive Effect (Y)"),
  subtitle = TeX("$x_{t, i}= \\theta_{11} x_{t-1, i} + \\theta_{12} y_{t-1, i} + \\mu_{x,i} + e_{1,i}$\n
                  $y_{t, i}= \\theta_{22} y_{t-1, i} + \\theta_{21} x_{t-1, i} + \\mu_{y,i} + e_{2,i}$" ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  #geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  #annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous(TeX("$var(\\mu_{x,i})$ "), limits = c(-1,1)) +
  scale_y_discrete("Random Intercept Variance (X)")

xlag_y <- ggplot(
  # Data 
  resultsSIM %>% filter(variance.between.x!=0.5) %>% 
         #rename model Model
         mutate(Model = as.factor(model)), 
  # Structure  
    aes(x = xlag_y, 
        y = as.factor(variance.between.x),
        grouping = as.factor(Model),
        color = Model,
        fill = Model,
            )) +
  geom_density_ridges( alpha = 0.25,
    scale = 1,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "black",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) +
  theme_minimal() +
  # title
  labs(title = paste("Cross Lagged Effect (X)"),
  subtitle = TeX("$x_{t, i}= \\theta_{11} x_{t-1, i} + \\theta_{12} y_{t-1, i} + \\mu_{x,i} + e_{1,i}$\n
                  $y_{t, i}= \\theta_{22} y_{t-1, i} + \\theta_{21} x_{t-1, i} + \\mu_{y,i} + e_{2,i}$" ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  #annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous(TeX("$var(\\mu_{x,i})$ "), limits = c(-1,1)) +
  scale_y_discrete("Random Intercept Variance (X)")

xlag_x <- ggplot(
  # Data 
  resultsSIM %>% filter(variance.between.x!=0.5) %>% 
         #rename model Model
         mutate(Model = as.factor(model)), 
  # Structure  
    aes(x = xlag_x, 
        y = as.factor(variance.between.x),
        grouping = as.factor(Model),
        color = Model,
        fill = Model,
            )) +
  geom_density_ridges( alpha = 0.25,
    scale = 1,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "black",
    point_alpha = 0.01,
    color = NA  # Remove the outline
  ) +
  theme_minimal() +
  # title
  labs(title = paste("Autoregressive Effect (X)"),
  subtitle = TeX("$x_{t, i}= \\theta_{11} x_{t-1, i} + \\theta_{12} y_{t-1, i} + \\mu_{x,i} + e_{1,i}$\n
                  $y_{t, i}= \\theta_{22} y_{t-1, i} + \\theta_{21} x_{t-1, i} + \\mu_{y,i} + e_{2,i}$" ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  #annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  scale_x_continuous(TeX("$var(\\mu_{x,i})$ "), limits = c(-1,1)) +
  scale_y_discrete("Random Intercept Variance (X)")

# write each to a separate pdf
ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/ylag_x.pdf", ylag_x, width = 8, height = 6)
ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/ylag_y.pdf", ylag_y, width = 8, height = 6)
ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/xlag_x.pdf", xlag_x, width = 8, height = 6)
ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/xlag_y.pdf", xlag_y, width = 8, height = 6)
```




```{r}
library(ggplot2)
library(ggridges)
resultsSIM %>%
  pivot_longer(
    cols = c(xlag_x, ylag_x, xlag_y, ylag_y),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) %>%
  mutate(variable = factor(variable, levels = c("xlag", "ylag"))) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "lag_type",
    values_to = "values"
  ) %>%
  filter(variance.between.y == 0.6) %>%
  filter(variance.between.x != 0.5) %>%
  filter(lag_type == "x") %>%
  ggplot(aes(x = values, 
             fill = "grey", 
             y = as.factor(variance.between.x),
             grouping = model,
             fill = model))+
            
  facet_wrap(~ as.factor(model)) + 
  geom_density_ridges2(
    alpha = 0.40,
    scale = 1,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 2,
    point_size = 1,
    point_alpha = 0.01,
    fill = " darkgrey",
    color = NA  # Remove the outline
  ) +
  theme_minimal() + 
  # title
  labs(title = paste("Monte Carlo Simulations (X)"),
  subtitle = TeX("$\\textbf{x_{t, i}= \\theta_{11} x_{t-1, i} + \\theta_{12} y_{t-1, i} + \\mu_{x,i} + e_{1,i}}$\n
                  $y_{t, i}= \\theta_{22} y_{t-1, i} + \\theta_{21} x_{t-1, i} + \\mu_{y,i} + e_{2,i}$" ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Parameter Estimate") +
  scale_y_discrete("Trait Variance (X)") 

ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/mc_plot.pdf", width = 8, height = 6)

```


```{r}
library(ggplot2)
library(ggridges)
resultsSIM %>%
  pivot_longer(
    cols = c(xlag_x, ylag_x, xlag_y, ylag_y),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) %>%
  mutate(variable = factor(variable, levels = c("xlag", "ylag"))) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "lag_type",
    values_to = "values"
  ) %>%
  filter(variance.between.y == 0.6) %>%
  filter(variance.between.x != 0.5) %>%
  filter(lag_type == "y") %>%
  ggplot(aes(x = values, 
             fill = "grey", 
             y = as.factor(variance.between.x),
             grouping = model,
             fill = model))+
            
  facet_wrap(~ as.factor(model)) + 
  geom_density_ridges2(
    alpha = 0.40,
    scale = 1,
    show.legend = TRUE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 2,
    point_size = 1,
    point_alpha = 0.01,
    fill = " darkgrey",
    color = NA  # Remove the outline
  ) +
  theme_minimal() + 
  # title
  labs(title = paste("Monte Carlo Simulations (Y)"),
  subtitle = TeX("$x_{t, i}= \\theta_{11} x_{t-1, i} + \\theta_{12} y_{t-1, i} + \\mu_{x,i} + e_{1,i}$\n
                  $\\textbf{y_{t, i}= \\theta_{22} y_{t-1, i} + \\theta_{21} x_{t-1, i} + \\mu_{y,i} + e_{2,i}}$" ))+
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Parameter Estimate") +
  scale_y_discrete("Trait Variance (y)") 

ggsave("~/Dropbox/github_repos/crossLag_p/crossLagR/tmpdat/mc_plot2.pdf", width = 8, height = 6)

```




The models captures these estimates perfectly. We could try the same thing with the continuous time structural equation model. This one takes a little bit -- which could be sped up by compiling once.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
library(lavaan)
monteCarloCTSEM(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.8,
   cov.pq = 0,
   sample_size = 5000) -> results


expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = seq(0.8, 1.6, by = .2),
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params

for(i in 1:nrow(mc_params)){
  results <- rbind(results, 
                   monteCarloCTSEM(
                                   trials = 25,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
```

The random intercept variances vary from -- 0.8 to 1.6, by increments of 0.2. The AR parameters (i.e., stability) are fixed at 0.2, and the cross lagged effects are fixed at 0. The "within subject" variances are fixed at 0.4.

```{r}

rm(list = ls())
devtools::load_all()
library(tidyr)
library(dplyr)
library(lavaan)
monteCarloLChange(
   trials = 1,
   waves = 5,
   dgp = "riclpm",
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   variance.between.x = 0.8,
   variance.between.y = 0.8,
   cov.pq = 0,
   sample_size = 5000) -> results


expand.grid(
   variance.between.x = seq(0.8, 1.6, by = .2),
   variance.between.y = seq(0.8, 1.6, by = .2),
   stability.p = 0.2,
   stability.q = 0.2,
   cross_lag_x = 0.0,
   cross_lag_y = 0.0,
   variance.q = 0.4,
   variance.p = 0.4,
   cov.pq = 0,
   sample_size = 5000) -> mc_params


for(i in 1:nrow(mc_params)){
  print(paste("🚀 Running Monte Carlo Simulation 🚀"))
  print(paste("Within Person Stability X: ", mc_params$variance.between.x[i], " | Within Person Stability Y: ", mc_params$variance.between.y[i]))
  results <- rbind(results, 
                   monteCarloLChange(
                                   trials = 25,
                                   waves = 5,
                                   dgp = "riclpm",
                                   variance.between.x = mc_params$variance.between.x[i],
                                   variance.between.y = mc_params$variance.between.y[i],
                                   stability.p = mc_params$stability.p[i],
                                   stability.q = mc_params$stability.q[i],
                                   cross_lag_x = mc_params$cross_lag_x[i],
                                   cross_lag_y = mc_params$cross_lag_y[i],
                                   variance.q = mc_params$variance.q[i],
                                   variance.p = mc_params$variance.p[i],
                                   cov.pq = mc_params$cov.pq[i],
                                   sample_size = mc_params$sample_size[i])
                )

}
```




```{r}
library(ggplot2)
library(ggridges)
results %>%
  ggplot(aes(x = xlag_y, 
             fill = "grey", 
             y = as.factor(variance.between.x)))+
  facet_wrap(~paste("Var(Y) = ", variance.between.y), scales = "free", ncol = 2) +
  geom_density_ridges2(
    alpha = 0.25,
    scale = 1,
    show.legend = FALSE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "black",
    point_alpha = 0.01,
    fill = " black",
    color = NA  # Remove the outline
  ) +
   geom_density_ridges2(
     aes(x = xlag_x, 
             fill = "grey", 
             y = as.factor(variance.between.x)),
    alpha = 0.25,
    scale = 1,
    show.legend = FALSE,
    quantile_lines = TRUE,
    rel_min_height = 0.01,
    quantiles = 0.5,
    calc_ecdf = FALSE,
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0),
    point_shape = 3,
    point_size = 1,
    point_color = "darkblue",
    point_alpha = 0.01,
    fill = "darkblue",
    color = NA  # Remove the outline
  ) +
  theme_minimal() + 
  # title
  labs(title = "Autoregressive (Blue)\nand Cross Lagged (Grey) Parameter Estimates") + 
  # Style the plot
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11) ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  annotate("text", x = 0.2, y = 1, label = "True AR ", hjust = 0, size = 3) +
  annotate("text", x = 0, y = 1, label = "True CL ", hjust = 0, size = 3) +
  scale_x_continuous("Autoregressive (AR) and Cross Lagged (CL) Parameters") +
  scale_y_discrete("Random Intercept Variance (Y)") 
```

The CTSEM also picks up the correct parameter values.

## The Latent Change Model

```{r}
library(chattr)
library(dplyr)
library(lavaan)
devtools::load_all()
library(dplyr)
simRICLPM(waves = 10,
          variance.p = 0.8,
          variance.q = 0.8,
          stability.p = 0.2,
          stability.q = 0.2,
          cross.q = 0,
          cross.p = 0,
          cov.pq = 0.3,
          cov.between = 0,
          variance.between.x = 0.8,
          variance.between.y = 0.8,
          sample.nobs = 10000)[[2]] %>%
  mutate(id = seq(1:nrow(.))) %>%
  lavaan::lavaan(model = latentChange(waves = 5, 
                              model_type = "dual_change"
                             ),
                meanstructure = FALSE,
                estimator = "ML",
                missing = "fiml",
                fixed.x = FALSE,
                mimic="mplus",
                control=list(iter.max=1000),
                verbose=FALSE) %>%summary(standardized = TRUE)
```


