---
title: "Simulator and sensitivity internals (contributor notes)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulator and sensitivity internals (contributor notes)}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Internal contributor doc.** Excluded from the built package via
> `.Rbuildignore`. Read this on GitHub before changing anything in
> `simCLPM`, `simRICLPM`, `simFromSyntax`, `populate_unified_labels`, or
> `sensitivityCLPM`.
>
> **Authorship.** Authored by Claude (Anthropic) under the prompt:
> *"now create a simple internal document that describes how to use the
> simulator. I've generally worked out the math correctly, but include a
> detailed (though concise and straightforward) well explained walkthrough
> of why the variance is this, and the characteristics of these new
> sensitivity functions. It can be a quarto in vignettes."*
> Maintainer review by C. Weber.

This vignette explains *why* each simulator does what it does --- the
variance accounting, the parameterization choices, and the assumptions
baked into the sensitivity diagnostic. The package CLPM chapter has the
substantive motivation; this is the implementation-level companion.

## Two routes to a simulated dataset

There are two ways to manufacture data in `crossLagR`:

1. **Direct simulators** (`simCLPM`, `simRICLPM`, `simCLPMu`) hard-code a
   specific lavaan population model. Each takes substantive arguments
   (`beta_x`, `omega_xy`, `var_p`, `var_BX`, ...) and emits a `list(model,
   data)` where `model` is the lavaan syntax used and `data` is the
   simulated wide-format data frame.
2. **The generic path** (`simFromSyntax` + `populate_unified_labels`)
   takes any estimator-syntax function (e.g. `estimateRICLPM`), substitutes
   user-supplied numeric values into the unified labels (`ar_x`, `cl_xy`,
   `d_var_x`, ...), and calls `lavaan::simulateData()` on the populated
   syntax.

Use a direct simulator when you need control over a specific DGP
(particularly `simCLPMu`, the only one that injects a time-varying
confounder). Use the generic path when you want to simulate from an
arbitrary unified-label estimator's own implied population.

## Direct simulators

### `simCLPM`

```{r}
sim <- simCLPM(waves    = 4,
               beta_x   = 0.4, beta_y = 0.4,
               omega_xy = 0.2, omega_yx = 0.2,
               var_x    = 1,   var_y    = 1,
               cov_xy   = 0.3,
               sample_size = 500, seed = 1)
```

The lavaan syntax (`sim$model`) declares perfect indicators
(`p =~ 1*y; q =~ 1*x`), so the latent dynamics live on `p` and `q`. The
wave-1 means default to 0 and the residual variances are constrained
across waves 2..T. Wave-1 variance can be set independently with
`var_y1`, `var_x1`, `cov_xy1`. Useful when you need the panel to look
stationary across all waves or want to inject a different initial
condition.

### `simRICLPM`

The RI-CLPM augments the CLPM with a between-person random intercept on
each construct:

\[
y_{it} = \mu_y + I_{y,i} + p_{it}, \qquad
x_{it} = \mu_x + I_{x,i} + q_{it}
\]

where $I_{y,i} \sim N(\mu_{BY}, \sigma^2_{BY})$ and $I_{x,i} \sim
N(\mu_{BX}, \sigma^2_{BX})$ are the trait factors, and $p_{it}, q_{it}$
are the within-person deviations on which AR/CL dynamics operate.

The signature mirrors that decomposition: `beta_x` / `omega_xy` are
within-person dynamics on $p$ and $q$; `var_p` / `var_q` are dynamic
residual variances; `var_BX` / `var_BY` / `cov_BXBY` parameterize the
trait factors.

\textbf{Identifiability constraints baked in.} The trait factors are
allowed to covary with each other (`cov_BXBY`) but are constrained to be
uncorrelated with the wave-1 within-person factors. This is the standard
RI-CLPM identification (Hamaker, Kuiper \& Grasman, 2015) and is what
makes the trait/within decomposition unique.

### `simCLPMu`

Same dynamic equation as `simCLPM` but each wave includes an additive
shock from a separate AR(1) confounder $u_t$:

\[
p_t = \text{stability}_p\, p_{t-1} + \text{cross}_q\, q_{t-1} + \gamma_p u_t,
\]

with $u_t = \rho\, u_{t-1} + \varepsilon_t$. Used to construct DGPs for
"how badly does X estimator do when there's an unmodeled confounder"
studies. The confounder is unobserved in `sim$data` (the function
returns `u1..uT` columns too, so you *can* model it; just drop them when
simulating realistic data).

## The generic path: `simFromSyntax`

`simFromSyntax(estimator, waves, sample_size, ar_x, ...)` is a thin
two-step pipeline:

```{r}
fn         <- crossLagR::estimateCLPM       # or any unified estimator
syntax     <- fn(waves = 4)                 # lavaan with labels ar_x*, etc.
populated  <- populate_unified_labels(      # substitute "ar_x*" -> "0.3*"
                syntax, ar_x = 0.3, ar_y = 0.3,
                cl_xy = 0.1, cl_yx = 0.1,
                d_var_x = 0.5, d_var_y = 0.5,
                d_cov_xy = 0)
dat        <- lavaan::simulateData(populated, sample.nobs = 500,
                                   meanstructure = TRUE)
```

`populate_unified_labels` uses word-boundary regex (`\\bar_x\\*`) so
`ar_xx` (if any estimator ever names a parameter that way) isn't a
partial match. Labels not supplied are left at lavaan's `simulateData`
defaults, which is why higher-level parameters (growth-factor variances
for ALT/LGM/LCMSR, trait variances for TSO, latent FE variances for BB)
end up as lavaan defaults rather than user-controlled --- a known
limitation we should fix if anyone ever needs control over them.

## Variance decomposition in `sensitivityCLPM`

The sensitivity function takes a fitted CLPM and asks: "If the true DGP
had the within-person dynamics I just estimated *plus* between-person
trait variance giving ICC $\kappa$, how badly would a CLPM mis-estimate
the AR/CL parameters?"

### Why the variance mapping is `var_BX = icc / (1 - icc) * d_var_x`

ICC is the fraction of total variance attributable to the trait:

\[
\text{ICC}_x \equiv \frac{\sigma^2_{BX}}{\sigma^2_{BX} + \sigma^2_{WX}}.
\]

Solving for $\sigma^2_{BX}$ given a target ICC $\kappa$ and a fixed
within-person variance $\sigma^2_{WX}$:

\[
\sigma^2_{BX} = \frac{\kappa}{1 - \kappa}\, \sigma^2_{WX}.
\]

The implementation uses the CLPM's estimated residual variance `d_var_x`
as the proxy for $\sigma^2_{WX}$. This is **not exactly correct**: the
total within-person variance under a stationary AR(1) is
$\sigma^2_{WX,\text{stationary}} = d\_var\_x / (1 - ar_x^2 - \text{cross
terms})$, not `d_var_x` itself. We use `d_var_x` directly because:

- For modest AR/CL values it is within a few percent of the steady-state
  within-person variance.
- The chapter's bias formula
  $b^{\text{CLPM}}_{xx} \approx \kappa + (1 - \kappa)\, b^{\text{true}}_{xx}$
  is itself a univariate approximation; matching the exact stationary
  variance would not make the diagnostic meaningfully more accurate.
- Users care about the qualitative shape (CLPM diverges from truth as
  ICC grows), not a calibrated bias point estimate.

If you ever need a calibrated sensitivity (e.g., to publish a bias
correction), replace `d_var_x` with the stationary variance and document
the change.

### What the function actually does

```r
sensitivityCLPM(fit, icc_grid, n_sims, sample_size, waves)
```

Pseudocode:

```
extract orig = (ar_x, ar_y, cl_xy, cl_yx, d_var_x, d_var_y, d_cov_xy) from fit
for icc in icc_grid:
  var_BX = icc / (1 - icc) * d_var_x
  var_BY = icc / (1 - icc) * d_var_y
  for s in 1..n_sims:
    dat = simRICLPM(
            waves, sample_size,
            beta_x   = orig$ar_x,  beta_y   = orig$ar_y,
            omega_xy = orig$cl_xy, omega_yx = orig$cl_yx,
            var_p    = d_var_x,    var_q    = d_var_y,
            cov_pq   = d_cov_xy,
            var_BX,  var_BY, cov_BXBY = 0
          )
    record CLPM and RI-CLPM AR/CL estimates fit to dat
```

### Why both CLPM and RI-CLPM are refit

The RI-CLPM is the *oracle* in this sensitivity: it knows the true
structure of the DGP, so its estimates should track the within-person
truth as ICC varies. The CLPM is the *target of the diagnosis*: it omits
the trait, so its estimates should drift away from the truth as ICC
grows. The gap between the two curves at any ICC is the visible bias.

The "truth" reference line in `plot.crossLagR_sensitivity` is the
original CLPM fit's estimate --- treated as the within-person truth by
construction. The RI-CLPM curve should sit on this line; the CLPM curve
should rise above it as ICC increases (for AR) or shift in either
direction (for CL, depending on the trait covariance --- currently fixed
at 0).

### Things to know before changing this

- `cov_BXBY = 0` is fixed. Allowing trait covariance would let the CL
  estimates be contaminated by `sigma_IxIy / (sigma^2_Iy + sigma^2_y*)`
  (see the chapter's univariate CL formula) in addition to the
  trait-variance term. If you expose this argument, add a regression
  test for the CL bias direction.
- `n_sims = 30` is enough to see the *shape* of the sensitivity curve;
  it is not enough to compute a sharp bias point estimate. The print
  method reports the SD per cell so users can judge MC noise.
- Convergence failures are silently turned into `NA` rows. This is fine
  for the plot (the ribbon just gets wider) but means downstream users
  who aggregate `sens$estimates` should `na.rm = TRUE`.

## Interpreting the sensitivity plot

The plot has four facets (`ar_x`, `ar_y`, `cl_xy`, `cl_yx`), each with
two curves (CLPM red, RI-CLPM blue) and a dashed horizontal line at the
baseline.

- **RI-CLPM tracks the baseline.** If it doesn't, either the DGP is so
  extreme that RI-CLPM is also struggling (high ICC, small N), or there
  is a bug in how `simRICLPM` is being parameterized.
- **CLPM curve rises with ICC for AR parameters.** This is the Hamaker
  bias: the AR coefficient absorbs the ICC. At ICC = 0 they should
  coincide.
- **CLPM curve for CL parameters is approximately flat (with
  `cov_BXBY = 0`).** The chapter's formula attributes CL bias to the
  trait *covariance*, which is zero here. If you let it drift, you'd see
  CL bias too.
- **Confidence ribbons (10th-90th pct across `n_sims`)** widen at high
  ICC because the fits are harder.

## Caveats users should know

1. **It is a what-if, not a calibration.** The CLPM estimates you pass in
   are already (potentially) biased; treating them as the within-person
   truth means the "truth reference" is itself uncertain. To get the
   actually-implied within-person dynamics, fit an RI-CLPM directly with
   `estimateRICLPM`.
2. **Use a small grid for exploration, a dense grid for publication.**
   `n_sims = 30` and 4-6 ICC points is enough to see the shape. Bump to
   `n_sims = 200` and 10+ ICC points if you're putting this in a paper.
3. **High-ICC cells are noisy.** ICC > 0.8 produces RI-CLPMs that
   frequently fail to converge. The function trims `icc_grid` to
   `[0, 1)` but does not warn at 0.9; consider lowering by hand.
4. **It assumes both fits use the same syntax wrappers.** If you pass a
   `fit` that wasn't built from `estimateCLPM` (e.g., your own custom
   lavaan syntax with non-standard labels), parameter extraction
   silently fails and you'll get `NA` estimates. The function checks for
   this and errors out before simulating.

## Adding a new sensitivity dimension

If you ever want to sweep a second axis (e.g., AR magnitude alongside
ICC), the simplest extension is:

```r
sensitivityCLPMGrid <- function(fit, icc_grid, ar_inflate_grid, n_sims, ...) {
  # for each combination of (icc, inflate):
  #   set orig$ar_x  <- orig$ar_x  * inflate
  #   set orig$ar_y  <- orig$ar_y  * inflate
  #   call the inner simulation loop from sensitivityCLPM
}
```

Keep the existing `sensitivityCLPM` as the one-axis sweep; expose the
grid version as a separate function so the plot method can specialize on
the one-axis case.
