# Installation and Package Syntax

## Installation

`crossLagR` can be found on GitHub and is not yet on CRAN. Install with `devtools` (or `remotes`):

```r
# install.packages("devtools")
devtools::install_github("crweber9874/crossLagR")
```

The package is generally just a wrapper around `lavaan` for SEM estimation, `MASS` for multivariate normal draws in some simulators, and the usual tidyverse tools (`dplyr`, `tidyr`, `tibble`, `purrr`, `stringr`) for data wrangling procedures. `DiagrammeR` is used to render the path diagrams reproduced throughout this guide. All dependencies are listed in the DESCRIPTION file and will be installed automatically with the above command.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(crossLagR)
library(lavaan)
library(dplyr)
```

## Package Conventions

Functions fall into three families, identifiable by their prefix:

- `estimate*()` — generate `lavaan` model syntax. Each returns a character string that can be fed directly into `lavaan::lavaan()` or `lavaan::sem()`. `estimateRICLPM(waves = 4)`, for example, returns lavaan syntax for a 4-wave RI-CLPM.
- `simFromSyntax()` — simulate data from any `estimate*()` model by populating its syntax with true parameter values, then drawing from it with `lavaan::simulateData()`. Returns a wide-format data frame with columns `x1, …, xT, y1, …, yT`.
- `monteCarlo*()` — repeat a simulate-then-fit cycle and return a tibble of results.

It's not clear to me whether thi is the optimal way to write a package, but it works for now.All `estimate*()` functions in the unified family share a common labeling convention so that estimates from different models can be lined up side-by-side. The obvious issue with this is that the interpretation of these parameters can vary across models, so caution is warranted. The convention is to generally follow the framework introduced by @usami2019:

| Label | Meaning | Used in |
|:---|:---|:---|
| `ar_x`, `ar_y` | Autoregressive parameters | every estimator |
| `cl_xy`, `cl_yx` | Cross-lagged parameters | every estimator |
| `I_x`, `I_y` | Trait / intercept factor | RI-CLPM, B&B, ALT, LCM-SR |
| `S_x`, `S_y` | Slope factor | ALT, LCM-SR, LGM |
| `A_x`, `A_y` | Accumulating change factor | LCSM |
| `u_var_x`, `u_var_y` | Unique factor (measurement-error) variance | LCSM |
| `d_var_x`, `d_var_y` | Dynamic residual variance | RI-CLPM, ALT, LCM-SR |
| `d_cov_xy` | Dynamic residual covariance | RI-CLPM, ALT, LCM-SR |

: Unified parameter labels in `crossLagR`. The *substantive* interpretation depends on the model — see later chapters — but the labels themselves are stable across estimators. For example, an AR parameter in one model may not mean the same as an AR parameter in another, but they will both be labeled `ar_x` or `ar_y` in the output.

## Common Arguments

Every `estimate*()` accepts at least:

- `waves` — number of time points. The lower bound for identification depends on the model (2 for CLPM, 3 for most others, 4 for `estimate_change_to_change`).
- `constrain_beta` — equality constraint across waves on AR parameters. Default `TRUE`.
- `constrain_omega` — equality constraint across waves on CL parameters. Default `TRUE`.

Bivariate/multivariate models add:

- `variable_type = c("univariate", "bivariate")` — for `estimateLChange()` and `estimateLGM()`.
- `include_z = TRUE` — for `estimateRICLPM()`; adds a third process.
- `constrain_residual_variances`, `constrain_residual_covariances` — equality constraints on innovation variances and within-wave covariances.


::: {.callout-important title="Data convention: column names must be x1, x2, ..., y1, y2, ..."}

```r
my_data |>
  tidyr::pivot_wider(
    id_cols     = case_id,
    names_from  = wave,
    values_from = c(personality, ideology)
  ) |>
  dplyr::rename_with(~ sub("^personality_", "x", .x)) |>
  dplyr::rename_with(~ sub("^ideology_",    "y", .x))
```

:::

## Estimator Quick Reference

```{r echo=TRUE, eval=FALSE}
# CLPM — baseline cross-lagged panel
estimateCLPM(waves = 4)

# RI-CLPM — random intercept + within-person AR/CL
estimateRICLPM(waves = 4)

# Latent Change — proportional change with optional accumulating factor
estimateLChange(waves = 4, variable_type = "bivariate",
                estimate_constant_change = TRUE)

# Linear Growth — intercept + slope, no dynamics
estimateLGM(waves = 4, variable_type = "bivariate")

# ALT — growth + AR/CL on observed scores (compounding)
estimateALT(waves = 4)

# LCM-SR / GCLM — growth + AR/CL on structured residuals
estimateLCMSR(waves = 4)

# Bollen & Brand — latent fixed effects dynamic panel
estimateBollen_and_Brand(waves = 4)

# Trait-State-Occasion — three-component variance decomposition
estimateTSO(waves = 4)
```

Each call returns a character string. To fit, pipe it into `lavaan`:

```{r echo=TRUE, eval=FALSE}
fit <- lavaan(estimateRICLPM(waves = 4), data = my_panel)
summary(fit, standardized = TRUE)
```

## Simulation Quick Reference

`simFromSyntax()` takes the same arguments as `estimate*()` plus true parameter values and sample size. It returns a wide-format data frame with columns `x1, …, xT, y1, …, yT`. For example:


```{r echo=TRUE, eval=FALSE}
# CLPM with 4 waves
dat <- simFromSyntax("CLPM", waves = 4, sample_size = 1000,
                     ar_x = 0.5, ar_y = 0.5,
                     cl_xy = 0.2, cl_yx = 0.1)

# RI-CLPM with between-person trait variance
dat <- simFromSyntax("RICLPM", waves = 4, sample_size = 1000,
                     ar_x = 0.3, ar_y = 0.3,
                     cl_xy = 0.2, cl_yx = 0.1,
                     var_between_x = 1, var_between_y = 1)

# Latent change
dat <- simFromSyntax("LCHANGE", waves = 5, sample_size = 1000,
                     ar_x = -0.5, ar_y = -0.5,
                     cl_xy = 0.3, cl_yx = 0.3,
                     estimator_args = list(variable_type = "bivariate"))

# Wide data frame: x1..xT, y1..yT
str(dat)
```

The wide-format output is what every `estimate*()` expects when fed to `lavaan`.

## Quick Example

```{r echo=TRUE, message=FALSE, warning=FALSE}
set.seed(2026)

# Simulate from an RI-CLPM
dat <- simFromSyntax("RICLPM", waves = 5, sample_size = 1000,
                     ar_x = 0.3, ar_y = 0.3,
                     cl_xy = 0.2, cl_yx = 0.1,
                     var_between_x = 1, var_between_y = 1)

# Fit the RI-CLPM
fit <- lavaan(estimateRICLPM(waves = 5), data = dat)

# Pull just the AR and CL parameters
parameterEstimates(fit) |>
  dplyr::filter(label %in% c("ar_x", "ar_y", "cl_xy", "cl_yx")) |>
  dplyr::select(label, est, se, pvalue) |>
  dplyr::mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## Help and Documentation

Every exported function has a roxygen-generated help page. The usual:

```r
?estimateRICLPM
?simFromSyntax
help(package = "crossLagR")
```

Source code is on GitHub at <https://github.com/crweber9874/crossLagR>; issues and feature requests welcome.
