# Revisiting the Cross-Lagged Panel Model

Social scientists often turn to panel data to diagnose causal effects when experiments are impractical [@chiu_causal_2025]. Whereas dozens of options exist for analyzing longer time series, researchers wishing to analyze cross-sectionally dominant panel structures -- in which there are more units than time points -- face a more limited set of choices. National surveys like found in the American National Election Studies (ANES) may consist of just a handful of waves, with many respondents interviewed only once or twice.

We consider of the popular methods for analyzing panel data in political science. We start by focusing on a common approach to estimate dynamic processes in panel data (particularly public opinion panel data that are "cross section" dominant, with far more respondents than panel waves: the cross-lagged panel model (CLPM). The CLPM -- introduced by @stanley1963 and extended on by @kenny1975 -- has been widely used to study reciprocal relationships between variables over time. These methods first appeared in the political science literature in the late 1970s [@abramowitz_impact_1978; @campbell_style_1979; @iyengar_testing_1978; @jennings_effect_1977; @markus_political_1979; @markus_dynamic_1979,@meier_issue_1979]. The availability of two large panel surveys with rich political content – Jennings and Niemi's @jennings_political_1974 socialization study and the ANES 1972-1974-1976 panel – likely promoted the CLPM's adoption, as most of the cross-lagged analyses published in this era analyze these two datasets. Another factor leading to the widespread adoption of the CLPM was the availability of Jöreskog's LISREL software, which allowed researchers to quickly and easily solve systems of equations using maximum likelihood [@markus_analyzing_1979; @mccullough_effects_1978; @shingles_causal_1976].

Much of the CLPM's appeal lies in its intuitive structure, flexibility, and ease of implementation. In the half century after its introduction, the CLPM has also been refined and extended, due to important limitations in its original formulation. For instance, the model can  account for unobserved heterogeneity across individuals -- stable characteristics of individuals that  confound the "dynamic" estimates in the CLPM [@usami2015; @usami2021; @klopack_and_wickrama2020; @kim_and_steiner2021; @ludtke2022; @mcardle2009]. Because the CLPM does not account for stable unit effects, the cross-lagged estimates will tend to be biased (upwards), as they conflate within-person dynamics with between-person differences. Though not directly referenced as a critique of the CLPM, @hsiao2022 shows that including a lagged dependent variable will produce a biased estimate of the lagged effect, because the variable is correlated with the error term. The correlation arises from the failure to account for "between unit" heterogeneity (pp. 64 -67). As a consequence, CLPM estimates may be biased, especially when the variables of interest exhibit high stability over time [@hamaker_critique_2015; @usami2015; @ludtke2022].

This problem is elaborated on in @hsiao2022. Dynamic models -- models that include lagged dependent variables -- often result in correlations between the observed variables and the error terms, leading to biased estimates. @hsiao2022 proposes and instrumental variable approach using lagged values of the dependent variable as instruments. When stable unit effects are ignored, the lagged dependent variable becomes correlated with the error term, violating the exogeneity assumption. This correlation leads to biased estimates of the lagged effect.

Another popular extension to the CLPM is the random intercept cross-lagged panel model (RI-CLPM) proposed in @hamaker_critique_2015, which separates within-person dynamics from between-person "trait" effects [@usami2015; @ludtke2022]. Another approach is the Latent Change Score Model (LSCM), which focuses more directly on estimating different change processes; the LCSM  also accounts for stable unit effects, by accounting for unit averaged change [@usami2015; @mcardle2009; @kim_and_steiner2021]. Another alternative is the latent growth model (LGM), which models individual trajectories over time [@usami2015; @mcardle2003].

Ultimately,preference for one model over another is strongly tied to data access and availability. While the CLPM and a restrictive version of the LCSM can be estimated with as few as two waves of data, the RI-CLPM, the Latent Growth Model, and Latent Change Score Model require more panel waves to properly identify the models. Data with three or more waves of data afford more leverage to estimate dynamic relationships. In this project, we elaborate on these models, beginning with the CLPM and change score approach to modeling dynamic relationships.

### The CLPM Diagram

The CLPM can be represented as a Directed Acyclic Graph (DAG) @fig-clpm [DAG; @pearl_causality_2001].
```{r}
#| label: fig-clpm
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| out-width: 80%
#| out-height: 80%
#| fig-cap: 'The Cross-Lagged Panel Model. The DAG displays cross-lagged and autoregressive paths between two variables over three time points.'

library(DiagrammeR)

grViz("
digraph CLPM_simple {
  graph [layout = neato, bgcolor = white, splines = true]
  node [shape = plaintext, fontsize = 24]

  Xt1 [pos = '-3.0,1.2!', label = <X<sub>1</sub>>]
  Xt2 [pos = '0.0,1.2!', label = <X<sub>2</sub>>]
  Xt3 [pos = '3.0,1.2!', label = <X<sub>3</sub>>]
  Yt1 [pos = '-3.0,-1.2!', label = <Y<sub>1</sub>>]
  Yt2 [pos = '0.0,-1.2!', label = <Y<sub>2</sub>>]
  Yt3 [pos = '3.0,-1.2!', label = <Y<sub>3</sub>>]

  Xt1 -> Xt2 [arrowsize = 0.7]
  Xt2 -> Xt3 [arrowsize = 0.7]
  Yt1 -> Yt2 [arrowsize = 0.7]
  Yt2 -> Yt3 [arrowsize = 0.7]

  Xt1 -> Yt2 [arrowsize = 0.7]
  Yt1 -> Xt2 [arrowsize = 0.7]
  Xt2 -> Yt3 [arrowsize = 0.7]
  Yt2 -> Xt3 [arrowsize = 0.7]
}
")
```

The CLPM -- a dynamic panel model -- leverages the temporal ordering of variables to estimate how one variable leads to change in another variable. Does the lagged effect of $y$ on $x$ exert a stronger influence than the lagged effect of $x$ on $y$? Is one variable *causally dominant* (for a critique, see [@rogosa_critique_1980]? Alternatively, contrasting the the cross-lagged effects is also critical to establishing *Granger Causality*
[@granger_investigating_1969].

A crucial limitation of the CLPM is that it does not account for stable trait-like differences between individuals. Consider a simple example, the relationship between a personality characteristic, *openness versus closed psychological orientation*, and political attitudes, a *right-wing political orientation* [@bakker_reconsidering_2021 @luttig_reconsidering_2021]. @bakker_reconsidering_2021 for instance, examines whether a psychological disposition towards openness predicts a right-wing political orientation, or whether a right-wing political orientation predicts a psychological disposition towards openness.

Political orientations are generally stable [@jennings_continuity_1975; @goren_party_2005; @ansolabehere_strength_2008]. It is also well regarded that personality traits are stable over time, and partially heritable [@costa_personality_1988; @roberts_patterns_2006; @hatemi_political_2015]. 

The CLPM  does not account for this time-invariant stability. It implicitly assumes that all variation in the observed variables is time-varying, and that there are no stable, time-invariant components. 

Following @hamaker_critique_2015 and @usami2021, the observed variables might be further decomposed into a time-invariant trait component and a time-varying component.

$$
\begin{eqnarray}
x_{it} = \mu_{xt} + I_{x,i} + x^*_{x,it} \\
y_{it} = \mu_{yt} + I_{y,i} + y^*_{y,it}
\end{eqnarray}
$$

where $\mu_{xt}$ and $\mu_{yt}$ are time-specific means, and $x^*_{it}$ and $y^*_{it}$ are person-specific deviations from those time-specific means [@usami2019; @usami2015; @usami2021]. 

 @fig-clpm-factors-minimal provides a visual representation of a modified version of the CLPM with time-invariant trait components.

```{r}
#| label: fig-clpm-factors-minimal
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| out-width: 80%
#| out-height: 80%
#| fig-cap: 'The Cross-Lagged Panel Model with Stable, Time-Invariant Unit Factors. The DAG displays cross-lagged and autoregressive paths between two variables over three time points. Intercepts, $I$ vary across respondents but not time.'

library(DiagrammeR)

grViz("
digraph CLPM_simple_factors {
  graph [layout = neato, bgcolor = white, splines = true]
  node [shape = plaintext, fontsize = 24]

  Xt1 [pos = '-3.0,1.2!', label = <X<sub>1</sub>>]
  Xt2 [pos = '0.0,1.2!', label = <X<sub>2</sub>>]
  Xt3 [pos = '3.0,1.2!', label = <X<sub>3</sub>>]
  Yt1 [pos = '-3.0,-1.2!', label = <Y<sub>1</sub>>]
  Yt2 [pos = '0.0,-1.2!', label = <Y<sub>2</sub>>]
  Yt3 [pos = '3.0,-1.2!', label = <Y<sub>3</sub>>]

  Ix [pos = '0,2.4!', label = <I<sub>x</sub>>]
  Iy [pos = '0,-2.4!', label = <I<sub>y</sub>>]

  Xt1 -> Xt2 [arrowsize = 0.7, label = <b<sub>xx</sub>>, fontsize = 14]
  Xt2 -> Xt3 [arrowsize = 0.7, label = <b<sub>xx</sub>>, fontsize = 14]
  Yt1 -> Yt2 [arrowsize = 0.7, label = <b<sub>yy</sub>>, fontsize = 14]
  Yt2 -> Yt3 [arrowsize = 0.7, label = <b<sub>yy</sub>>, fontsize = 14]

  Xt1 -> Yt2 [arrowsize = 0.7, label = <b<sub>xy</sub>>, fontsize = 14]
  Yt1 -> Xt2 [arrowsize = 0.7, label = <b<sub>yx</sub>>, fontsize = 14]
  Xt2 -> Yt3 [arrowsize = 0.7, label = <b<sub>xy</sub>>, fontsize = 14]
  Yt2 -> Xt3 [arrowsize = 0.7, label = <b<sub>yx</sub>>, fontsize = 14]

  Ix -> Xt1 [arrowsize = 0.6]
  Ix -> Xt2 [arrowsize = 0.6]
  Ix -> Xt3 [arrowsize = 0.6]
  Iy -> Yt1 [arrowsize = 0.6]
  Iy -> Yt2 [arrowsize = 0.6]
  Iy -> Yt3 [arrowsize = 0.6]

  Ix -> Iy [dir = both, style = dashed, color = gray40, constraint = false,
            arrowsize = 0.5, tailport = sw, headport = nw, minlen = 12,
            label = <σ<sub>I<sub>x</sub>I<sub>y</sub></sub>>, fontsize = 12, labeldistance = 2.0]
}
")
```

It is useful to ask the question: What if one fits @fig-clpm but @fig-clpm-factors-minimal is the true population model. Assume the following model is estimated.

$$
\begin{align}
x_{it} &=  b^{CLPM}_{xx}\,x_{i,t-1} + b^{CLPM}_{yx}\,y_{i,t-1} + e_{x,it} \\
y_{it} &=  b^{CLPM}_{yy}\,y_{i,t-1} + b^{CLPM}_{xy}\,x_{i,t-1} + e_{y,it}
\end{align}
$$

The conventional interpretation of the CLPM coefficients are that $\beta_{xx}$, captures the autoregressive effect of $x$ on itself, and $\beta_{yy}$ does the same for $y$. On the other hand, $\beta_{yx}$ captures the cross-lagged effect of $y$ on $x$, and $\beta_{xy}$ captures the cross-lagged effect of $x$ on $y$. 

If @fig-clpm-factors-minimal captures the true data-generating process, the above intepretation of the CLPM parameters is incorrect.The degree of bias in the parameters should be proportional to the size of the trait component. In @fig-clpm-factors-minimal, both $I_x$ and $I_y$ represent the time invariant trait components of the observed measures. The variance of $x$ and $y$ can be decomposed into a between-person component (the trait) and a within-person component (the deviation from the trait).

$$
\begin{aligned}
  {Var}(x_{it}) &= \text{Var}(I_{x,i}) + \text{Var}(x^*_{x,it})\\
  {Var}(y_{it}) &= \text{Var}(I_{y,i}) + \text{Var}(y^*_{y,it})
\end{aligned}
$$


$$
\begin{align}
x^*_{it} &=  b^{True}_{xx}\,x^*_{i,t-1} + b^{True}_{yx}\,y^*_{i,t-1} + e_{x,it} \\
y^*_{it} &=  b^{True}_{yy}\,y^*_{i,t-1} + b^{True}_{xy}\,x^*_{i,t-1} + e_{y,it}
\end{align}
$$


Importantly, the percentage of the total variance attributable to the between-person component is the **intraclass correlation**: $\text{ICC}_x \equiv \sigma^2_{I_x}/(\sigma^2_{I_x} + \sigma^2_{x^*})$, with $\text{ICC}_y$ defined analogously.

The problem with fitting the CLPM to data generated by @fig-clpm-factors-minimal is that a non-zero covariance forms between the lagged regressors and the error terms. This is because the lagged regressor contains the trait component. The intuition is easiest to see through univariate estimates, estimating the cross-lagged and autoregressive effects individually — what one would estimate from $x_t$ on $x_{t-1}$ alone, or $x_t$ on $y_{t-1}$ alone.\footnote{Also assume stationarity and assuming $\text{Cov}(x^*_t, y^*_t) = 0$}.

$$
b_{xx}^{\text{univariate}}
\equiv \frac{\text{Cov}(x_t, x_{t-1})}{\text{Var}(x_{t-1})}
= \frac{\sigma^2_{I_x} + b^{\text{true}}_{xx}\,\sigma^2_{x^*}}{\sigma^2_{I_x} + \sigma^2_{x^*}}
= \text{ICC}_x + (1-\text{ICC}_x)\,b^{\text{true}}_{xx},
$$

$$
b_{yx}^{\text{univariate}}
\equiv \frac{\text{Cov}(x_t, y_{t-1})}{\text{Var}(y_{t-1})}
= \frac{\sigma_{I_xI_y} + b^{\text{true}}_{yx}\,\sigma^2_{y^*}}{\sigma^2_{I_y} + \sigma^2_{y^*}}
= \frac{\sigma_{I_xI_y}}{\sigma^2_{I_y} + \sigma^2_{y^*}} + (1-\text{ICC}_y)\,b^{\text{true}}_{yx}.
$$

If the true autoregressive effect is zero ($b^{\text{true}}_{xx}=0$), then $b_{xx}^{\text{univariate}} = \text{ICC}_x$ — the CLPM picks up a spurious AR equal to the ICC. If the true cross-lagged effect is zero ($b^{\text{true}}_{yx}=0$), then $b_{yx}^{\text{univariate}} = \sigma_{I_xI_y}/(\sigma^2_{I_y}+\sigma^2_{y^*})$ — a spurious cross-lag driven entirely by the trait covariance. If the trait covariance is also zero ($\sigma_{I_xI_y}=0$), then $b_{yx}^{\text{univariate}} = (1-\text{ICC}_y)\,b^{\text{true}}_{yx}$, i.e. attenuated by the share of variance that is *between*-person.

The *multivariate* CLPM includes both lags simultaneously, so its coefficients are partial slopes rather than these simple ratios. But the general intuition still holds. When $\sigma^2_{I_x} > 0$, the AR is inflated by the trait variance, and the CL is contaminated by the trait covariance. When $\sigma^2_{I_x} = 0$, there is no bias in the AR, but the CL is still attenuated by the between-person variance in $y$.

## Demonstration

The bias is proportional to the size of the trait component -- how much of the total variance is defined by the between-person trait. This can be demonstrated with the `crossLagR` package and a simple simulation. We simulate data from a four-wave panel with the following truth values.

| Quantity | Value |
|---|---:|
| waves | 4 |
| n_obs | 3000 |
| $\sigma_{x^*}$ | 0.5 |
| $\sigma_{I_x}$ | Varies |
| $\sigma_{I_y}$ | Varies |
| $b^{True}_{xx}$ (true) | 0.30 |
| $b^{True}_{yy}$ (true) | 0.30 |
| $b^{True}_{yx}$ (true) | 0.00 |
| $b^{True}_{xy}$ (true) | 0.00 |
| $\rho_I$ | 0.5 |


```{r}
#| label: bias-demo-setup
#| message: false
#| warning: false
#| cache: false
#| echo: false

suppressPackageStartupMessages({
  library(crossLagR); library(lavaan); library(tidyverse)
})

devtools::load_all()
waves       <- 4
n_obs       <- 3000
within_var  <- 0.5
true_vals   <- c(ar_x = 0.30, ar_y = 0.30, cl_xy = 0.00, cl_yx = 0.00)
rho_I       <- 0.5

clpm_syn   <- estimateCLPM(waves   = waves)
riclpm_syn <- estimateRICLPM(waves = waves)

# Build an RI-CLPM population at a given between-to-within variance ratio.
# bw_ratio = 0 reduces to a pure CLPM (no trait variance).
make_riclpm_pop <- function(bw_ratio) {
  pt <- lavaanify(riclpm_syn)
  pt$ustart[pt$label == "ar_x"]     <- true_vals["ar_x"]
  pt$ustart[pt$label == "ar_y"]     <- true_vals["ar_y"]
  pt$ustart[pt$label == "cl_xy"]    <- true_vals["cl_xy"]
  pt$ustart[pt$label == "cl_yx"]    <- true_vals["cl_yx"]
  pt$ustart[pt$label == "d_var_x"]  <- within_var
  pt$ustart[pt$label == "d_var_y"]  <- within_var
  pt$ustart[pt$label == "d_cov_xy"] <- 0
  # Percent between within
  bw_var <- bw_ratio * within_var
  set_cov <- function(pt, lhs, rhs, val) {
    idx <- (pt$lhs == lhs & pt$op == "~~" & pt$rhs == rhs & pt$free > 0) |
           (pt$lhs == rhs & pt$op == "~~" & pt$rhs == lhs & pt$free > 0)
    pt$ustart[idx] <- val; pt
  }
  pt <- set_cov(pt, "I_x", "I_x", bw_var)
  pt <- set_cov(pt, "I_y", "I_y", bw_var)
  pt <- set_cov(pt, "I_x", "I_y", rho_I * bw_var)
  pt
}

## Simple function to fit lavaan model and extract parameters. 
fit_one <- function(syntax, dat, est_name) {
  fit <- tryCatch(
    suppressWarnings(lavaan::lavaan(syntax, data = dat, meanstructure = FALSE)),
    error = function(e) NULL
  )
  labels <- c("ar_x", "ar_y", "cl_xy", "cl_yx")
  if (is.null(fit) || !lavInspect(fit, "converged")) {
    return(tibble(estimator = est_name, parameter = labels, estimate = NA_real_))
  }
  pt <- parameterEstimates(fit)
  tibble(
    estimator = est_name,
    parameter = labels,
    estimate  = vapply(labels, function(l) {
      v <- pt$est[pt$label == l]; if (length(v)) v[1] else NA_real_
    }, numeric(1))
  )
}
```


```{r}
#| label: fig-bias-by-bw-ratio
#| message: false
#| warning: false
#| cache: true
#| fig-cap: "Monte Carlo mean bias (estimate − truth) for the four AR/CL parameters as the between-to-within variance ratio increases. Bands are 5th-95th percentiles across replications."

bw_grid <- c(0, 0.25, 0.5, 1, 2, 4, 9)
n_rep   <- 100

set.seed(2026)
sweep_df <- map_dfr(bw_grid, function(bw) {
  pop <- make_riclpm_pop(bw)
  map_dfr(seq_len(n_rep), function(rep_id) {
    dat <- simulateData(pop, sample.nobs = n_obs)
    bind_rows(
      fit_one(clpm_syn,   dat, "CLPM (misspecified)"),
      fit_one(riclpm_syn, dat, "RI-CLPM (true DGP)")
    ) |>
      mutate(
        bw_ratio = bw,
        rep_id   = rep_id,
        n_obs    = nrow(dat),
        n_miss   = sum(is.na(dat))
      )
  })
}) |>
  mutate(truth = true_vals[parameter],
         bias  = estimate - truth,
         parameter = factor(parameter,
                            levels = c("ar_x", "ar_y", "cl_xy", "cl_yx")))

plot_df <- sweep_df |>
  group_by(estimator, bw_ratio, parameter) |>
  summarise(
    mean_bias = mean(bias, na.rm = TRUE),
    p05       = quantile(bias, probs = 0.05, na.rm = TRUE),
    p95       = quantile(bias, probs = 0.95, na.rm = TRUE),
    .groups   = "drop"
  )

ggplot(plot_df, aes(bw_ratio, mean_bias, color = estimator, fill = estimator)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.4) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ parameter, scales = "free_y") +
  labs(x = "Between-to-within variance ratio",
       y = "Monte Carlo mean bias  (estimate − truth)",
       color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

```