# Bias in the Cross-Lagged Panel Model {#sec-bias}

The previous chapter introduced the cross-lagged panel model and the substantive question it purports to address -- conditioning on variables in the past predicts future values. This chapter turns to its central problem, effectively ignored by the CLPM: when individuals carry stable, trait-like differences across waves, the standard CLPM conflates those differences with within-person dynamics. It may appear as though a relationship exists between past and present values, but the relationship is instead an artifact of unmoving differences between individuals (as opposed to within individuals). The chapter derives the population bias in closed form, demonstrates it with a short simulation, and sketches out a sensitivity analysis framework.

A crucial limitation of the CLPM is that it does not account for stable trait-like differences between individuals. Consider a simple example, the relationship between a personality characteristic, *openness versus closed psychological orientation*, and political attitudes, a *right-wing political orientation* [@bakker_reconsidering_2021 @luttig_reconsidering_2021]. @bakker_reconsidering_2021 for instance, examines whether a psychological disposition towards openness predicts a right-wing political orientation, or whether a right-wing political orientation predicts a psychological disposition towards openness.

Political orientations are generally stable [@jennings_continuity_1975; @goren_party_2005; @ansolabehere_strength_2008]. It is also well regarded that personality traits are stable over time, and partially heritable [@costa_personality_1988; @roberts_patterns_2006; @hatemi_political_2015]. 

The CLPM  does not account for this time-invariant stability. It implicitly assumes that all variation in the observed variables is time-varying, and that there are no stable, time-invariant effects. 

Following @hamaker_critique_2015 and @usami2021, the observed variables might be further decomposed into a time-invariant trait component and a time-varying component.

$$
\begin{eqnarray}
x_{it} = \mu_{x,t} + I_{x,i} + x^*_{x,it} \\
y_{it} = \mu_{y,t} + I_{y,i} + y^*_{y,it}
\end{eqnarray}
$$

where $\mu_{x,t}$ and $\mu_{y,t}$ are time-specific means, and $x^*_{it}$ and $y^*_{it}$ are person-specific deviations from wave-specific means [@usami2019; @usami2015; @usami2021]. 

 @fig-clpm-factors-minimal provides a visual representation of a modified version of the CLPM with time-invariant trait components.

```{r}
#| label: fig-clpm-factors-minimal
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| out-width: 80%
#| out-height: 80%
#| fig-cap: 'The Cross-Lagged Panel Model with Stable, Time-Invariant Unit Factors. The DAG displays cross-lagged and autoregressive paths between two variables over three time points. Intercepts, $I$ vary across respondents but not time.'

library(DiagrammeR)

grViz("
digraph CLPM_simple_factors {
  graph [layout = neato, bgcolor = white, splines = true]
  node [shape = plaintext, fontsize = 24]

  Xt1 [pos = '-3.0,1.2!', label = <X<sub>1</sub>>]
  Xt2 [pos = '0.0,1.2!', label = <X<sub>2</sub>>]
  Xt3 [pos = '3.0,1.2!', label = <X<sub>3</sub>>]
  Yt1 [pos = '-3.0,-1.2!', label = <Y<sub>1</sub>>]
  Yt2 [pos = '0.0,-1.2!', label = <Y<sub>2</sub>>]
  Yt3 [pos = '3.0,-1.2!', label = <Y<sub>3</sub>>]

  Ix [pos = '0,2.4!', label = <I<sub>x</sub>>]
  Iy [pos = '0,-2.4!', label = <I<sub>y</sub>>]

  Xt1 -> Xt2 [arrowsize = 0.7, label = <b<sub>xx</sub>>, fontsize = 14]
  Xt2 -> Xt3 [arrowsize = 0.7, label = <b<sub>xx</sub>>, fontsize = 14]
  Yt1 -> Yt2 [arrowsize = 0.7, label = <b<sub>yy</sub>>, fontsize = 14]
  Yt2 -> Yt3 [arrowsize = 0.7, label = <b<sub>yy</sub>>, fontsize = 14]

  Xt1 -> Yt2 [arrowsize = 0.7, label = <b<sub>xy</sub>>, fontsize = 14]
  Yt1 -> Xt2 [arrowsize = 0.7, label = <b<sub>yx</sub>>, fontsize = 14]
  Xt2 -> Yt3 [arrowsize = 0.7, label = <b<sub>xy</sub>>, fontsize = 14]
  Yt2 -> Xt3 [arrowsize = 0.7, label = <b<sub>yx</sub>>, fontsize = 14]

  Ix -> Xt1 [arrowsize = 0.6]
  Ix -> Xt2 [arrowsize = 0.6]
  Ix -> Xt3 [arrowsize = 0.6]
  Iy -> Yt1 [arrowsize = 0.6]
  Iy -> Yt2 [arrowsize = 0.6]
  Iy -> Yt3 [arrowsize = 0.6]

  Ix -> Iy [dir = both, style = dashed, color = gray40, constraint = false,
            arrowsize = 0.5, tailport = sw, headport = nw, minlen = 12,
            label = <σ<sub>I<sub>x</sub>I<sub>y</sub></sub>>, fontsize = 12, labeldistance = 2.0]
}
")
```

To demonstrate precisely why the CLPM will produce biased estimates, it is important to trace how stable unit effects, $I_x$ and $I_y$, bias the autoregressive and cross-lagged parameters, should a researcher proceed to fit a standard CLPM ignoring unit effects. What if one opts to fit 
 @fig-clpm but @fig-clpm-factors-minimal is the true data generating model? 
 
$$
\begin{align}
x_{it} &=  \hat{b}_{xx}\,x_{i,t-1} + \hat{b}_{yx}\,y_{i,t-1} + e_{x,it} \\
y_{it} &=  \hat{b}_{yy}\,y_{i,t-1} + \hat{b}_{xy}\,x_{i,t-1} + e_{y,it}
\end{align}
$$

The conventional interpretation of the CLPM coefficients are that $\hat{b}_{xx}$, captures the autoregressive effect of $x$ on itself, and $\hat{b}_{yy}$ does the same for $y$. On the other hand, $\beta_{yx}$ captures the cross-lagged effect of $y$ on $x$, and $\hat{b}_{xy}$ captures the cross-lagged effect of $x$ on $y$. 

### What Economists have Said for a Bit

This general problem -- stable unit effects in panel data -- have been recognized for economists for decades. The issue is that when unit effects are present, this generates a non-zero correlation between the regressors and the error terms, leading to biased and inconsistent estimates. Only in the absence of unit effects, should the standard CLPM retrieve unbiased estimates. An immediate solution might be to include the unit effects in the model -- by subtracting the unit means from each variable, or equivalently, by including a dummy for the units in the regression (what is colloquially referred to as "fixed effects"). While this approach is reasonable in the presence of unit effects, it quickly becomes problematic when including lagged realizations of the dependent variable in the regression equation. And unfortunately, this contributes to bias in short panels, a problem known as the *dynamic panel bias* or *Nickell bias* [@nickell1981]. The bias is a mechanical artifact of the estimation procedure. 

The bias can be understood through a simple scenario. To control for unit effects, a researcher subtracts the unit means from each variable (the *within transformation*). However, because of the panel structure, each unit mean ($\eta_i$) is itself a function of the unit fixed effect, plus the wave specific errors (the within effect).

$$
\bar{y}_i = \frac{1}{T}\sum_{t=1}^{T}\left(\eta_i + \sum_{s=0}^{\infty}\rho^s\varepsilon_{i,t-s}\right)
$$

But now the demeaned lagged dependent variable contains past errors that overlap with the error term, thereby producing non-zero correlation between the regressors and the error terms, and once again, biased and inconsistent estimates. Incidentally, *Dynamic panel bias* decreases as $T$ increases, and is most problematic in *short T* panels. The bias is $O(1/T)$; meaning that it decreases as $T$ increases (@nickell1981).

@anderson_hsiao_1981 provide an alternative approach to confront *unit effects* through    *first-differencing* and using lagged variables as instruments. @arellano_bond_1991 develop 
a gmm approach, which uses varying lags as instruments (aslo @blundell_bond_1988, @balatagi_2013). Other "fixed effects" estimators\footnote{Used in terms of removing the unit effects.} @bollen_general_2010 show that SEM rather than GMM can be used to estimate the dynamic panel effects -- the autoregressive and cross lagged in the CLPM.

We return to these approaches later in the book. But for now, suffice it to say that these issues have been documented in the econometrics literature for at least 45 years. Only recently has a there emerged a clear bridge connecting the econometrics literature with the SEM literature, which we build upon. @ludtke2022 provide a useful overview.


### The Causal Interpretation

Another way to demonstrate the problem inherent in the CLPM is to consider the causal interpretation of the model.\footnote{We fully acknowledge that causal inference is challenging using observational data. We agree. At the same time, the CLPM is often discussed in a causal context, problematic as that may be. So, we feel it reasonable to also demonstrate that the causal interpretation of the CLPM is also problematic.} The causal parameters of interest in the CLPM are the cross-lagged effects. @ludtke2022 provide a causal interpretation of CLPM, among other models. In particular, the CLPM might be thought of as a means of estimating a dose-response relationship -- *how much does $y_t$ change for a one unit increase of $x$ at time $t-1$*? Each lagged variable can be thought of as a continuous *treatment* and the contemporaneous variable as the *outcome*. Importantly, let $Y_t(x)$ be the potential value of $y_{it}$ had the treatment $x_{i,t-1}$ been set to level $x$. Symmetrically, let $X_t(y)$ be the potential value of $x_{it}$ under $y_{i,t-1}=y$. $\tau_{x\to y}$  and $\tau_{y\to x}$ represent the causal effects parameters of interest, **the expected change in the outcome per unit change in the lagged treatment**.


$$
\tau_{x\to y} \equiv E\!\left[\frac{\partial\, Y_t(x)}{\partial x}\right]
= E\big[Y_t(x+1)-Y_t(x)\big],
\qquad
\tau_{y\to x} \equiv E\!\left[\frac{\partial\, X_t(y)}{\partial y}\right]
= E\big[X_t(y+1)-X_t(y)\big].
$$

Assuming a linear effect, $\tau_{x\to y}$ and $\tau_{y\to x}$ are estimated using the structural cross-lagged slopes from the CLPM. But here's the rub. To identify $\tau$ one must make the assumption of **conditional ignorability**. The lagged treatment values must effectively be as-good-as-randomly-assigned, conditional on confounders. This seems implausible. If $I_x$ influences $y_{it}$ and is correlated with $I_y$ -- it serves as a confounder and opens a "back-door" path to the outcome. 

$\tau_{x\to y}$ and $\tau_{y\to x}$ can be estimated using least squares regression. However, the CLPM regression of $x_t$ on $x_{t-1}$ and $y_{t-1}$ does not condition on $I_x$ or $I_y$, so this path is left open and the autoregressive and cross lagged parameters now include both state and trait effects.


```{r}
#| label: fig-clpm-po
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| out-width: 85%
#| fig-cap: 'Cross-lagged parameters of interest [@ludtke2022]. Confounding comes from time-invariant *traits* (characteristics of units, fixed over the panel; observed $i_x,i_y$, unobserved $u_x,u_y$) and time-varying *states* (observed $I_x,I_y$, unobserved $U_x,U_y$). Both drive the lagged treatments and contemporaneous outcomes, opening back-doors the CLPM leaves unblocked.'

library(DiagrammeR)

grViz("
digraph CLPM_LUDTKE {
  graph [layout = neato, splines = false, bgcolor = white, outputorder = edgesfirst]
  node [shape = plaintext, fontsize = 18]

  Cinv [shape = box, style = 'rounded,dashed', color = gray45, fontsize = 12, margin = 0.12,
        pos = '0,2.9!',
        label = <<b>traits (time-invariant)</b><br/><font point-size='9'>characteristics of units; fixed over the panel</font><br/><font point-size='10'>obs i<sub>x</sub>, i<sub>y</sub> &middot; unobs u<sub>x</sub>, u<sub>y</sub></font>>]
  Cvar [shape = box, style = 'rounded,solid', color = gray45, fontsize = 12, margin = 0.12,
        pos = '0,-2.9!',
        label = <<b>states (time-varying)</b><br/><font point-size='9'>change from wave to wave</font><br/><font point-size='10'>obs I<sub>x</sub>, I<sub>y</sub> &middot; unobs U<sub>x</sub>, U<sub>y</sub></font>>]

  Xt1 [pos = '-2.0,1.1!',  label = <X<sub>t-1</sub>>]
  Xt  [pos = '2.0,1.1!',   label = <X<sub>t</sub>>]
  Yt1 [pos = '-2.0,-1.1!', label = <Y<sub>t-1</sub>>]
  Yt  [pos = '2.0,-1.1!',  label = <Y<sub>t</sub>>]

  tauYX [pos = '1.4,0.95!',  label = <τ<sub>y→x</sub>>, fontsize = 15]
  tauXY [pos = '1.4,-0.95!', label = <τ<sub>x→y</sub>>, fontsize = 15]

  edge [color = gray65, penwidth = 0.6, arrowsize = 0.6]
  Cinv -> Xt1
  Cinv -> Xt
  Cinv -> Yt1
  Cinv -> Yt
  Cvar -> Xt1
  Cvar -> Xt
  Cvar -> Yt1
  Cvar -> Yt

  edge [color = black, penwidth = 1.5, arrowsize = 0.8]
  Xt1 -> Yt
  Yt1 -> Xt
}
")
```

And the problem is the same: the CLPM does not account for stable, time-invariant unit effects. The CLPM is a model of the *state* effects, but the *trait* effects are left unblocked, and since they influence both the lagged treatment and the contemporaneous outcome, they confound the causal effect of interest.


### The Intraclass Correlation

I understand the problem in terms of the **Intraclass Correlation**(ICC), the measure we always report in a multilevel model to show how much the data are subject to "clustering" around the unit. The ICC can be a useful diagnostic, as it decomposes the variance of the observed variables into the stable *between* and time-varying *within* components. 

$$
\begin{aligned}
  {Var}(x_{it}) &= \text{Var}(I_{x,i}) + \text{Var}(x^*_{it})\\
  {Var}(y_{it}) &= \text{Var}(I_{y,i}) + \text{Var}(y^*_{it})
\end{aligned}
$$


The ICC is the proportion of the total variance attributable to the unit: $\text{ICC}_x \equiv \sigma^2_{I_x}/(\sigma^2_{I_x} + \sigma^2_{x^*})$, with $\text{ICC}_y$ defined analogously. The intraclass correlation plays an important role in understanding the nature of bias in the CLPM. 

This is my understanding of its usefulness for the CLPM.Recall that the CLPM makes no distinction between trait and state effects: the variances of both $x$ and $y$ are composites of stable unit effects and time-varying departures from that trait. This trait-state distinction unspecified. 

Yet, say the structural parameters -- the cross lagged and autoregressive structure -- are captured with a least squares regression model.

$$
\begin{aligned}
x^*_{it} &=  \beta_{xx}\,x^*_{i,t-1} + \beta_{yx}\,y^*_{i,t-1} + e_{x,it} \\
y^*_{it} &=  \beta_{yy}\,y^*_{i,t-1} + \beta_{xy}\,x^*_{i,t-1} + e_{y,it}
\end{aligned}
$$

From linear regression, the least squares estimates of $\beta$ will only be unbiased and consistent if there is no covariance between the error terms and the regressors. In the CLPM , this only occurs if the $var(I_x)$ and $var(I_y)$ are zero and $\text{Cov}(I_x, I_y) = 0$. Otherwise, it can be shown that the estimates will be biased in a manner that is proportional to the variance in $y$ and $x$ attributable to a stable, underlying trait. 

## Deriving the Bias in the CLPM

A key -- perhaps questionable -- assumption below is that the *trait* and *state* components are orthogonal. Say one estimates a CLPM model. And suppose *y* is *political ideology* and *x* is a *closed or rigid cognitive style*. If one or both of these variables is driven by a stable trait -- there exist nonzero variance across units -- then:

* The AR paramters are biased proportional the the variance in $x$ and $y$ attributable to the stable, underlying trait
* The CL parameters are biased when the stable traits are correlated across units. The bias is proportional to the covariance between the stable traits.

And that's pretty much it. The AR parameters are biased by the variances, CL by the covariances. 

The *multivariate* CLPM estimates two lagged effects **simultaneously**, so every coefficient is a *partial* slope, and if standardized, equivalent to the partial correlation (see @rogosa_critique_1980 for a classic critique of the partial correlation parameters in the CLPM). Start with the the classic linear regression model, and to keep things compact, stack the $x_{it}$ and $y_{it}$ variables into a vector $\mathbf z_{it} = (x_{it}, y_{it})'$. 


$$
\mathbf B = X'X^{-1}X'Y = \text{Cov}(\mathbf z_{it}, \mathbf z_{i,t-1})\,\text{Var}(\mathbf z_{i,t-1})^{-1} = \mathbf C\,\mathbf S^{-1}
$$

On the right hand side, we can construct the covariance matrices $\mathbf C$ and $\mathbf S$ as functions of the trait and state components. In particular, define the following matrices (along with $\mathbf I$, the identity matrix) to *define* the bias in the CLPM, differentiate the between and within effects.

- **Between Effects**: $\boldsymbol\Omega = \text{Var}(\mathbf I_i)$: This is the **trait (between-person) covariance matrix**. On the diagonals are trait variances $\sigma^2_{I_x}, \sigma^2_{I_y}$ and on the off-diagonal is the trait covariance $\sigma_{I_xI_y}$. Call $x$ **closed personality** and $y$ (**right-wing ideology**) maybe -- and this matrix represents the between-person (co)variability in the aspect of these ideology and personality that does not change over the course of a panel.

- **Within Effects**: $\boldsymbol\Psi = \text{Var}(\mathbf z^*_{it})$ --- the **within-person covariance matrix**, which is formed from the wave-specific deviations from the trait. This is the dynamic part of the model, allowing one to test whether a change in one variable at $t-1$ translates to a change in another at time $t$. Returning to the example of closed personality and right-wing ideology, this matrix captures the variability in the change in these variables over time -- for instance, if one shifts towards a right leaning orientation at time $t-1$, does that predict a shift towards a more closed personality at time $t$? 

- **Within Effects**: $\mathbf B^{\text{true}}$ --- this matrix is the dynamic AR and CL parameters, the **true within-person dynamics**: autoregressions on the diagonal, cross-lags off the diagonal. These are the parameters necessary to estimate the causal effects of interest, the $\tau$'s. 

### No Separation of Trait and State in the CLPM

@arellano_bond_1991, @usami2015, and @usami2021, as well as @hamaker_critique_2015 show that the CLPM is a combination of the trait and state components, and failing to separate them contributes to based estimates.

We build on @usami2021's work showing that CLPM is a restrictive variant of a larger class of models. It's restrictive in the sense it never separates $\boldsymbol\Omega$ and $\boldsymbol\Psi$. Instead, the data $x$ and $y$ is a combination of these two components.

To identify the model, we need to calculate the inverse of the variance matrix of the lagged regressors. Recall from OLS that the regression coefficients are a function of the covariance of the regressor and outcome scaled to the variance of the regressors. The covariance of of the regressor (lagged $x$ and $y$) is $\mathbf S$, the inverse of the variance of the regressors is $\mathbf S^{-1}$, and the covariance between the regressors and the outcomes, $\mathbf C$. This is equivalent to $\mathbf (X^TX^{-1})$ as in $\mathbf y = \mathbf X\boldsymbol\beta + \mathbf e = (\mathbf X^T\mathbf X)^{-1}\mathbf X^T\mathbf y$.  

Starting with the *variances*, the variance of the lagged regressors is a combination of the trait and state components.

$$
\mathbf S \equiv \text{Var}(\mathbf z_{i,t-1}) = \boldsymbol\Omega + \boldsymbol\Psi
$$

And,the *covariances* between the lagged regressors and the outcome is also a combination of the trait and state components.

$$
\mathbf C \equiv \text{Cov}(\mathbf z_{it}, \mathbf z_{i,t-1}) = \boldsymbol\Omega + \mathbf B^{\text{true}}\boldsymbol\Psi
$$

The important thing to note is the separability of trait and state effects -- they are by construction orthogonal, where the trait contributes only to $\boldsymbol\Omega$. This may be an untenable assumption in practice.

[*Say the lagged regressor is authoritarianism. If authoritarians are more rigid, unlikely to change from wave to wave, this assumption is violated.*]{style="color:red;"}

The slopes are calculated -- the AR and CL parameters -- by regressing $z_{it}$ on $z_{i,t-1}$.


$$
\hat{\mathbf B}^{\text{CLPM}} = \mathbf C\,\mathbf S^{-1}
$$


We can decompoase both the $\mathbf S$ and $\mathbf C$, into "between" (trait) and "within" (state) components. The $\mathbf C$ matrix represents the covariances between the lagged regressors and the outcomes. If there is a trait variance component in both $x$ and $y$ and these
trait components are correlated, this will produce bias in both the autoregressive and cross-lagged parameters. And both matrices can be further expressed as functions of these components, with a construction that generally follows how one demonstrates the linear regression coefficients are biased if regressors are endogenous.

Start from $\hat{\mathbf B}^{\text{CLPM}} = \mathbf C\,\mathbf S^{-1}$ but just substitute the two variance components above.  The within matrix is simply the total variance minus the between variance (we don't need to specify it differently, as long as we know the total variance and the between variance): $\boldsymbol\Psi = \mathbf S - \boldsymbol\Omega$, so that,

$$
\mathbf C = \boldsymbol\Omega + \mathbf B^{\text{true}}\boldsymbol\Psi
          = \boldsymbol\Omega + \mathbf B^{\text{true}}(\mathbf S - \boldsymbol\Omega)
          = \mathbf B^{\text{true}}\mathbf S + (\mathbf I - \mathbf B^{\text{true}})\boldsymbol\Omega .
$$

Now post-multiply by $\mathbf S^{-1}$. The first term collapses to $\mathbf B^{\text{true}}$, and everything that is left over becomes the bias: 

$$
\hat{\mathbf B}^{\text{CLPM}} = \mathbf B^{\text{true}} + \bigl(\mathbf I - \mathbf B^{\text{true}}\bigr)\,\mathbf V,
\qquad \mathbf V \equiv \boldsymbol\Omega\,\mathbf S^{-1}.
$$

The bias is carried by $\mathbf V = \boldsymbol\Omega\,\mathbf S^{-1}$. It's just the trait covariance matrix divided by the total covariance matrix...it's like a matrix version of the ICC, including between variances and covariances.

Note that if there is no trait variance ($\boldsymbol\Omega=\mathbf 0$ (there is thus no trait covariance) then $\mathbf V=\boldsymbol\Omega\mathbf S^{-1}=\mathbf 0$), the bias goes away, and the CLPM retrieves the true dynamic parameters. Likewise, *if there is trait variance, but no covariance, then the CL parameter will be correct*.\footnote{The notable exception is *dynamic panel bias*, which still may exist. Check into this...it's a frustrating addition.} When a trait component is present, the degree by which the parameters are biased depends on the size of the trait component -- the ICC, or equivalently ratio of trait to state variance -- *and* the true dynamic parameters.

- **Diagonal of $\mathbf V$**. This represents from the trait *variances*, the diagonal of $\boldsymbol\Omega$). It will inflate the *autoregressive parameters*
.
- **Off-diagonal of $\mathbf V$**. The trait **covariances**, the off-diagonal of $\boldsymbol\Omega$. Setting $\mathbf B^{\text{true}} = \mathbf 0$ leaves $\hat{\mathbf B}^{\text{CLPM}} = \mathbf V$, an entirely spurious cross-lag from trait covariance alone.

## Simulation Demonstration

So, the bias is proportional to the size of the trait component -- how much of the total variance is defined by the between-person trait. This can be demonstrated with the `crossLagR` package and a simple simulation. We simulate data from a four-wave panel with the following parameter values.

| Quantity | Value |
|---|---:|
| waves | 4 |
| n_obs | 3000 |
| $\sigma_{x^*}$ | 0.5 |
| $\sigma_{I_x}$ | Varies |
| $\sigma_{I_y}$ | Varies |
| $b^{True}_{xx}$ | 0.30 |
| $b^{True}_{yy}$ | 0.30 |
| $b^{True}_{yx}$ | 0.00 |
| $b^{True}_{xy}$ | 0.00 |
| $\rho_I$ | 0.5 |


```{r}
#| label: bias-demo-setup
#| message: false
#| warning: false
#| cache: false
#| echo: false

# AI recommends using suppressPackageStartupMessages(). This is just a standard load, suppressing messagges.
suppressPackageStartupMessages({
  library(crossLagR); library(lavaan); library(tidyverse)
})

devtools::load_all()
waves       <- 4
n_obs       <- 3000
within_var  <- 0.5
true_vals   <- c(ar_x = 0.30, ar_y = 0.30, cl_xy = 0.00, cl_yx = 0.00)
# The correlation between Ix and Iy
rho_I       <- 0.5

clpm_syn   <- estimateCLPM(waves   = waves)
riclpm_syn <- estimateRICLPM(waves = waves)

# This is a useful part of lavaan, lavaanify. 
# Basically, take the syntax of a lavaan model, return a dataframe with the parameter table, including starting values, labels, and free/fixed parameters. Then modify...
make_riclpm_pop <- function(bw_ratio) {
  pt <- lavaanify(riclpm_syn)
  pt$ustart[pt$label == "ar_x"]     <- true_vals["ar_x"]
  pt$ustart[pt$label == "ar_y"]     <- true_vals["ar_y"]
  pt$ustart[pt$label == "cl_xy"]    <- true_vals["cl_xy"]
  pt$ustart[pt$label == "cl_yx"]    <- true_vals["cl_yx"]
  pt$ustart[pt$label == "d_var_x"]  <- within_var
  pt$ustart[pt$label == "d_var_y"]  <- within_var
  pt$ustart[pt$label == "d_cov_xy"] <- 0
  # Percent between within
  bw_var <- bw_ratio * within_var
  set_cov <- function(pt, lhs, rhs, val) {
    idx <- (pt$lhs == lhs & pt$op == "~~" & pt$rhs == rhs & pt$free > 0) |
           (pt$lhs == rhs & pt$op == "~~" & pt$rhs == lhs & pt$free > 0)
    pt$ustart[idx] <- val; pt
  }
  pt <- set_cov(pt, "I_x", "I_x", bw_var)
  pt <- set_cov(pt, "I_y", "I_y", bw_var)
  pt <- set_cov(pt, "I_x", "I_y", rho_I * bw_var)
  pt
}

## Simple function to fit lavaan model and extract parameters. 
fit_one <- function(syntax, dat, est_name) {
  fit <- tryCatch(
    suppressWarnings(lavaan::lavaan(syntax, data = dat, meanstructure = FALSE)),
    error = function(e) NULL
  )
  labels <- c("ar_x", "ar_y", "cl_xy", "cl_yx")
  if (is.null(fit) || !lavInspect(fit, "converged")) {
    return(tibble(estimator = est_name, parameter = labels, estimate = NA_real_))
  }
  pt <- parameterEstimates(fit)
  tibble(
    estimator = est_name,
    parameter = labels,
    estimate  = vapply(labels, function(l) {
      v <- pt$est[pt$label == l]; if (length(v)) v[1] else NA_real_
    }, numeric(1))
  )
}
```

So we're using the package estimate functions to generate lavaan syntax, then using lavaanify to extract the parameter table, and then modifying the starting values of the parameters to create a population model with a specified between-to-within variance ratio. 


```{r}
#| label: fig-bias-by-bw-ratio
#| echo: false
#| message: false
#| warning: false
#| cache: true
#| fig-cap: "Monte Carlo mean bias (estimate − truth) for the four AR/CL parameters as the between-to-within variance ratio increases. Bands are 5th-95th percentiles across replications."

bw_grid <- c(0, 0.25, 0.5, 1, 2, 4, 9)
n_rep   <- 100

# map is a useful tidy function form purrr to iterate over a list of values and return a list of results. map_dfr is a variant that returns a data frame, row-binding. Basically we're just iterating over the bw_grid, simulating data, fitting the two models, and extracting the parameter estimates.
set.seed(2026)
sweep_df <- map_dfr(bw_grid, function(bw) {
  pop <- make_riclpm_pop(bw)
  map_dfr(seq_len(n_rep), function(rep_id) {
    dat <- simulateData(pop, sample.nobs = n_obs)
    bind_rows(
      fit_one(clpm_syn,   dat, "CLPM (misspecified)"),
      fit_one(riclpm_syn, dat, "RI-CLPM (true DGP)")
    ) |>
      mutate(
        bw_ratio = bw,
        rep_id   = rep_id,
        n_obs    = nrow(dat),
        n_miss   = sum(is.na(dat))
      )
  })
}) |>
  mutate(truth = true_vals[parameter],
         bias  = estimate - truth,
         parameter = factor(parameter,
                            levels = c("ar_x", "ar_y", "cl_xy", "cl_yx")))

plot_df <- sweep_df |>
  group_by(estimator, bw_ratio, parameter) |>
  summarise(
    mean_bias = mean(bias, na.rm = TRUE),
    p05       = quantile(bias, probs = 0.05, na.rm = TRUE),
    p95       = quantile(bias, probs = 0.95, na.rm = TRUE),
    .groups   = "drop"
  )

ggplot(plot_df, aes(bw_ratio, mean_bias, color = estimator, fill = estimator)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.4) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ parameter, scales = "free_y") +
  labs(x = "Between-to-within variance ratio",
       y = "Monte Carlo mean bias  (estimate − truth)",
       color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

```

The autoregressive bias is monotonic  -- increasing the trait variance inflates the AR estimate towards 1. The cross-lagged bias is more complicated. It is zero when there is no trait variance, it peaks and then returns to zero. Below it becomes clear that this peak is simply a function of $rho_I$, the shared variance between trait *x* and trait *y*. 

## The Bias Function

We might want to ask questions like, "*if the true cross-lag coefficient were zero, and the trait correlation were `rho_I`, how much trait variance would produce a reported effect?*" The sensititivity analysis proposes an *as-if* scenario: What would we expect the degree of bias to be given a pre-specified degree of trait-(co)variance.

First, recall the bias function, which is just a generalization of the classic omitted-variable-bias formula [@greene_2018; @wooldridge_2010].

$\hat{\mathbf B}^{\text{CLPM}} - \mathbf B^{\text{true}} = (\mathbf I - \mathbf B^{\text{true}})\,\mathbf V$

```{r}
#| warning: false
#| cache: false
clpmBias(7, 0.3, rho_i = 0.5)
```


The important thing to note is that the bias is a function of the trait covariance matrix $\boldsymbol\Omega$ and the total covariance matrix $\mathbf S$. We can just program thi into an `R` function that calculates the bias at user specified levels of trait variance, trait covariance, and true AR and CL parameters. 

`crossLagR::clpmBias()` takes input *assuming* that the true-cross lagged parameter is zero -- if this is not the case, one can pass the CL and AR parameters to the bias function in `crossLagR::clpmCLBias()`, which generalises the formula to a non-zero true cross-lag and calculates the degree of bias.

Below , we vary the cross lag parameter from 0 to 0.10, and calculate the degree of bias at a low level of trait variance (between-to-within ratio = 0.01), so there's effectively no unit clustering.

```{r}
#| label: artifact-coupling-lowtrait
#| echo: true
#| message: true
#| warning: false
#| cache: false
varying_parameters <- clpmCLBias(
  bw_ratio = 0.01,
  ar       = 0.3,
  cl       = c(0, 0.02, 0.05, 0.10),
  rho_i    = 0.5,
  expand   = TRUE,
  verbose  = TRUE
)
round(
  varying_parameters[, c("cl", "ar_pop", "cl_pop",
                    "ar_bias", "ar_bias_pct",
                    "cl_bias", "cl_bias_pct")],
  3
)
```


`crossLagR::cl_bias_peak_ratio()` calculates the between-to-within variance ratio $r^\star$ at which the cross-lagged bias peaks. From the figures above, it's clear this is a non-monotonic function. So, fixing $b^{\text{true}}$ and $\rho_I$, where does the peak actually occur?

The location of the peak is can be found by differentiating the bias function, assuming $\rho_I \neq 1$ and $r > 0$, $f'(r)=0$ is $r^\star = 1/\sqrt{1-\rho_I^2}$. In other words, (1) the peak of the function is independent of the true AR parameter, and (2) the peak is a function of the trait correlation $\rho_I$. The more shared trait variance, the largest the bias. This means we know the *maximum* bias that can be expected for a given trait correlation. In a sense we can say -- this is the worst-case scenario; the maximum amount of bias to emerge from fitting a CLPM and ignoring the trait-state distinction is $r^\star$.

Take this simple simulation below, which builds off the one from above, using the following parameter table:

| Quantity | Value |
|---|---:|
| waves | 4 |
| n_obs | 3000 |
| $\sigma_{x^*}$ | 0.5 |
| $\sigma_{I_x}$ | Varies |
| $\sigma_{I_y}$ | Varies |
| $b^{True}_{xx}$ | 0.30 |
| $b^{True}_{yy}$ | 0.30 |
| $b^{True}_{yx}$ | 0.00 |
| $b^{True}_{xy}$ | 0.00 |
| $\rho_I$ | 0.5 |

Instead, vary the parameter values for the cross-lag, fix the AR parameter at values above, assuming $\rho


```{r}
#| label: artifact-coupling-peak
#| echo: true
#| message: true
#| warning: false
#| cache: false
# rho_I and true_vals come from `bias-demo-setup` above; r_star is the
# trait amount at which CL bias peaks for this rho_I.
artifact_peak <- clpmCLBias(
  bw_ratio = cl_bias_peak_ratio(rho_i = rho_I),
  ar       = true_vals[["ar_x"]],
  cl       = c(0, 0.02, 0.05, 0.10),
  rho_i    = rho_I,
  expand   = TRUE,
  verbose  = TRUE
)
round(
  artifact_peak[, c("cl", "ar_pop", "cl_pop",
                         "ar_bias", "ar_bias_pct",
                         "cl_bias", "cl_bias_pct")],
  3
)
```


There are a few additional functions, `cl_bias_peak_ratio()`. This one is pretty simple, also used a lot, and it's just $r^\star = 1/\sqrt{1-\rho_I^2}$. So, let's simply calculate that point. Then we'll do two things: (1) calculate the amount of bias at that point, and (2) calculate a tipping point. Understanding the former is important for the latter. 

To calculate the *maximum* amount of bias, this is again just the peak of teh bias function, and is exclusively a function of the trait correlation $\rho_I$, which we square to create an $R^2$ variance in one explained by the other, standardized covariance. 


Take the following example. Let's assume, from the parameters above, that we want to explore bias in the cross lagged parameter. Here's how to do that -- I think -- 


(1) Assume a true value of the AR parameter (from table above).
(2) Assume **no** true cross-lagged effect.
(3) Calculate $r^\star$, the point at which bias is maximized, given the trait correlation $\rho_I$.
(4) Calculate the bias at that point, which is the maximum bias possible for that $\rho_I$.

```{r}
#| label: peak-checker
#| echo: false
#| message: false
#| warning: false

r_star      <- cl_bias_peak_ratio(rho_i = rho_I)
peak_bias   <- clpmBias(bw_ratio = r_star, ar = true_vals[["ar_x"]],
                        rho_i = rho_I)
```


This asks an interesting question: *say we ignore the trait-state distinction, and fit the CLPM do the data. Though we've ignored the distinction, does it really matter? What's the degree of bias produced by the trait-state confounding? How large is the bias in relation the the reported cross-lagged effect?* 


```{r}
#| label: fig-cl-bias-peak
#| echo: false
#| message: false
#| warning: false
#| out-width: 75%
#| fig-cap: "Cross-lagged bias."

# Internal package helper (loaded by devtools::load_all() above);
# the true AR and trait correlation come straight from the setup.
plot_cl_bias_peak(ar = true_vals[["ar_x"]], rho_i = rho_I)
```




## Functions Used in This Chapter

| Function | What it does |
|:---|:---|
| `clpmBias()` | Population AR and CL bias of the CLPM under an RI-CLPM truth, assuming the true cross-lag is zero. |
| `clpmCLBias()` | Generalizes `clpmBias()` to a non-zero true cross-lag: returns the population CLPM AR/CL estimate and bias, optionally under indicator measurement error. |
| `clpmArtifactTipping()` | Inverts `clpmCLBias()`): given an *observed* CLPM cross-lag, solves for the trait-variance ratio at which that estimate would be reproduced as pure artifact. |
| `cl_bias_peak_ratio()` | Closed-form variance ratio $r^\star = 1/\sqrt{1-\rho_I^2}$ at which the cross-lagged bias peaks. |
| `plot_cl_bias_peak()` | Plots CL bias against the between-to-within variance ratio, marking the peak at $r^\star$. |
| `plot_cl_bias_surface()` | Plots the CL bias surface across the variance ratio and the true autoregressive parameter, with a ridge line at $r^\star$. |

