# Latent Change Score Models

Whereas the previous chapter established dynamic lagged and cross lagged effects in *level* form (the CLPM and its bias under unit heterogeneity), this chapter turns to the **change** representation. By "differencing" the observed scores, this effectively absorbs all time-invariant heterogeneity, including stable between-person differences.

The Latent Change Score Model (LCSM) parameterizes dynamics differently as *differences* across waves [@mcardle2009; @kim_and_steiner2021]. The logic is to remove out any stable respondent effects by simply calculating the wave to wave difference. The LCSM uses a measurement equation that separates latent true scores from unique factors, and a dynamic equation that maps lagged true scores onto latent change scores via a proportional self-feedback parameter ($\beta$, i.e. `ar_x`) and a coupling parameter ($\omega$, i.e. `cl_xy`). Finally, a second-order *constant change* factor can be specified which accumulates over waves, though requires 4+ waves for identification. This term captures stable, person-specific trajectories that are not accounted for by differencing.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(crossLagR)
library(dplyr)
library(ggplot2)
library(lavaan)
```


## The LCSM

@fig-lcm-simple sketches the bivariate LCSM with a constant-change factor, in the same DAG style introduced in the previous chapter. Each observed score is decomposed as $X_t = X_{t-1} + \Delta x_t$; each change score is regressed on prior levels via proportional change ($\beta$) and coupling ($\omega$), and accumulates a person-specific constant ($\Delta_x$, $\Delta_y$). The detailed measurement-model version with multiple indicators is reproduced in the appendix as @fig-lcm-full.

```{r}
#| label: fig-lcm-simple
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
#| fig-cap: "Bivariate Latent Change Score Model. β = proportional change; ω = coupling. Δ_x, Δ_y are person-specific constant-change factors loading on each Δx_t and Δy_t."
library(DiagrammeR)
grViz("
digraph LCSM_simple {
  graph [layout = neato, bgcolor = white, splines = line]
  node [shape = plaintext, fontsize = 24]
  edge [arrowsize = 0.6]

  Xt1 [pos = '-3.0,1.8!', label = <X<sub>1</sub>>]
  Xt2 [pos = '-1.0,1.8!', label = <X<sub>2</sub>>]
  Xt3 [pos = '1.0,1.8!',  label = <X<sub>3</sub>>]
  Xt4 [pos = '3.0,1.8!',  label = <X<sub>4</sub>>]

  dX2 [pos = '-1.0,0.6!', label = <Δx<sub>2</sub>>]
  dX3 [pos = '1.0,0.6!',  label = <Δx<sub>3</sub>>]
  dX4 [pos = '3.0,0.6!',  label = <Δx<sub>4</sub>>]

  dY2 [pos = '-1.0,-0.6!', label = <Δy<sub>2</sub>>]
  dY3 [pos = '1.0,-0.6!',  label = <Δy<sub>3</sub>>]
  dY4 [pos = '3.0,-0.6!',  label = <Δy<sub>4</sub>>]

  Yt1 [pos = '-3.0,-1.8!', label = <Y<sub>1</sub>>]
  Yt2 [pos = '-1.0,-1.8!', label = <Y<sub>2</sub>>]
  Yt3 [pos = '1.0,-1.8!',  label = <Y<sub>3</sub>>]
  Yt4 [pos = '3.0,-1.8!',  label = <Y<sub>4</sub>>]

  Ax [pos = '-4.5,1.2!',  label = <Δ<sub>x</sub>>]
  Ay [pos = '-4.5,-1.2!', label = <Δ<sub>y</sub>>]

  # Identity: X_t = X_{t-1} + Δx_t
  Xt1 -> Xt2 [label = '1', fontsize = 14]
  Xt2 -> Xt3 [label = '1', fontsize = 14]
  Xt3 -> Xt4 [label = '1', fontsize = 14]
  Yt1 -> Yt2 [label = '1', fontsize = 14]
  Yt2 -> Yt3 [label = '1', fontsize = 14]
  Yt3 -> Yt4 [label = '1', fontsize = 14]
  dX2 -> Xt2 [label = '1', fontsize = 14]
  dX3 -> Xt3 [label = '1', fontsize = 14]
  dX4 -> Xt4 [label = '1', fontsize = 14]
  dY2 -> Yt2 [label = '1', fontsize = 14]
  dY3 -> Yt3 [label = '1', fontsize = 14]
  dY4 -> Yt4 [label = '1', fontsize = 14]

  # Proportional change (β)
  Xt1 -> dX2 [label = <β>, fontsize = 16]
  Xt2 -> dX3 [label = <β>, fontsize = 16]
  Xt3 -> dX4 [label = <β>, fontsize = 16]
  Yt1 -> dY2 [label = <β>, fontsize = 16]
  Yt2 -> dY3 [label = <β>, fontsize = 16]
  Yt3 -> dY4 [label = <β>, fontsize = 16]

  # Coupling (ω)
  Yt1 -> dX2 [label = <ω>, fontsize = 16]
  Yt2 -> dX3 [label = <ω>, fontsize = 16]
  Yt3 -> dX4 [label = <ω>, fontsize = 16]
  Xt1 -> dY2 [label = <ω>, fontsize = 16]
  Xt2 -> dY3 [label = <ω>, fontsize = 16]
  Xt3 -> dY4 [label = <ω>, fontsize = 16]

  # Constant-change accumulating factors (left side)
  Ax -> dX2 [arrowsize = 0.5, label = '1', fontsize = 12]
  Ax -> dX3 [arrowsize = 0.5, label = '1', fontsize = 12]
  Ax -> dX4 [arrowsize = 0.5, label = '1', fontsize = 12]
  Ay -> dY2 [arrowsize = 0.5, label = '1', fontsize = 12]
  Ay -> dY3 [arrowsize = 0.5, label = '1', fontsize = 12]
  Ay -> dY4 [arrowsize = 0.5, label = '1', fontsize = 12]
}
")
```



## `estimateLChange()`: Generated `lavaan` Syntax

`estimateLChange()` returns a `lavaan` model string with the unified labels: `ar_x`, `ar_y` are proportional self-feedback (the effective autoregressive parameter is $1+AR()$); `cl_xy`, `cl_yx` are coupling (cross-lagged) parameters; `A_x`, `A_y` are accumulating constant-change factors; `u_var_x`, `u_var_y` are unique-factor (measurement-error) variances. The function returns syntax that can be passed directly to `lavaan::lavaan()`.

### Univariate LCSM with constant change

The simplest specification: one variable, four waves, constant-change factor switched to **on**.

```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateLChange(waves                    = 4,
                variable_type            = "univariate",
                estimate_constant_change = TRUE) |>
  cat()
```

### Bivariate LCSM with constant change

This adds the second process and the coupling parameters `cl_xy`, `cl_yx`. With `constrain_beta = TRUE` and `constrain_omega = TRUE` the proportional and coupling parameters are held equal across waves.

```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateLChange(waves                    = 4,
                variable_type            = "bivariate",
                constrain_beta           = TRUE,
                constrain_omega          = TRUE,
                estimate_constant_change = TRUE) |>
  cat()
```

### Bivariate LCSM without constant change

Dropping the accumulating factor leaves only proportional change and coupling — the LCSM analogue of the CLPM in change form.

```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateLChange(waves                    = 4,
                variable_type            = "bivariate",
                estimate_constant_change = FALSE) |>
  cat()
```

### Bivariate LCSM with change-to-change effects

Requires $\geq 4$ waves; adds parameters `phi_x`, `phi_y` (change AR) and `phi_cl_x`, `phi_cl_y` (change cross-lag). 

Unfortunately, there is not a "level" analogue to this specification. The change-to-change parameters capture change on change itself. In particular, $\phi_x$ is the effect of prior change in $x$ on subsequent change in $x$, while $\phi_{cl_x}$ is the effect of prior change in $y$ on current/subsequent change in $x$.

```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateLChange(waves                     = 4,
                constrain_beta           = TRUE,
                constrain_omega          = TRUE,
                estimate_constant_change = TRUE,
                variable_type             = "bivariate",
                estimate_change_to_change = TRUE) |>
  cat()
```

## `simLChange()`: Simulating from the LCSM

`simLChange()` simulates data under user specified true values. Arguments mirror the estimator syntax: `beta_x`/`beta_y` (proportional change), `omega_x`/`omega_y` (coupling), `constant_mean_x`/`constant_mean_y` and `latent_variance_x`/`latent_variance_y` (constant-change factor mean and variance), and `estimate_constant_change` to toggle accumulating change.

```{r echo=TRUE, message=FALSE, warning=FALSE}
sim <- simLChange(
  waves                    = 5,
  variable_type            = "bivariate",
  beta_x                   = -0.7,
  beta_y                   = -0.7,
  omega_x                  = 0.5,
  omega_y                  = 0.5,
  estimate_constant_change = FALSE,
  sample.nobs              = 500
)
str(sim$data, list.len = 6)
```

## Algebraic Equivalence: LCSM = CLPM in Change Form

Without a constant-change factor and without measurement error, the LCSM dynamic equation $\Delta x_t = \beta_x x_{t-1} + \omega_x y_{t-1} + u$ rearranges to $x_t = (1+\beta_x) x_{t-1} + \omega_x y_{t-1} + u$ — exactly the CLPM regression on the level scores. The CLPM should recover $1+\beta$ for autoregressives and $\omega$ for cross-lags. I do a lot of this throughout the document -- it's useful to see the equivalence across models

1. **Measurement error.** `simLChange()` defaults `indicator_variance_x = indicator_variance_y = 0.5`, putting noise on the observed scores. The CLPM has no latent measurement model, so its coefficients are attenuated by errors-in-variables. The LCSM separates true scores from indicators and is immune.
2. **Constant change.** When the accumulating factor is on, each unit carries a stable trajectory; the CLPM treats this as wave-specific variation and biases AR up / CL down.

The simulated data -- under the latent change specification -- leads to estimates for the CLPM which mirror the generating parameters. 


```{r echo = TRUE, message = FALSE, warning = FALSE}
clean <- simLChange(
  waves                    = 5,
  variable_type            = "bivariate",
  beta_x                   = -0.2,
  beta_y                   = -0.2,
  omega_x                  = 0.20,
  omega_y                  = 0.55,
  indicator_variance_x     = 0.001,
  indicator_variance_y     = 0.001,
  estimate_constant_change = FALSE,
  sample.nobs              = 1000
)
cat("DGP: LCSM, no constant change, near-zero measurement error.\n",
    "Expect AR_x ≈ AR_y ≈ 0.80, CL_yx ≈ 0.20, CL_xy ≈ 0.55.\n", sep = "")
summary(lavaan(estimateCLPM(waves = 5), data = clean$data),
        standardized = TRUE)
```

## Constant Change and Unit Effects

Now turn the constant-change factor on, keeping measurement error off. This creates an upward bias for the AR parameter and a downward bias for the CL parameter in the CLPM.

```{r echo = TRUE, message = FALSE, warning = FALSE}
lcsm_data <- simLChange(
  waves                    = 5,
  variable_type            = "bivariate",
  beta_x                   = -0.5,
  beta_y                   = -0.5,
  omega_x                  = 0.20,
  omega_y                  = 0.55,
  indicator_variance_x     = 0.01,
  indicator_variance_y     = 0.01,
  estimate_constant_change = TRUE,
  latent_variance_x        = 1,
  latent_variance_y        = 1,
  sample.nobs              = 1000
)
cat("DGP: LCSM with constant change (near-zero measurement error).\n",
    "Estimator: CLPM (does not absorb unit effects).\n", sep = "")
summary(lavaan(estimateCLPM(waves = 5), data = lcsm_data$data),
        standardized = TRUE)
```

The corresponding LCSM fit on the same data recovers the generating parameters:

```{r echo = TRUE, message = FALSE, warning = FALSE}
summary(lavaan(estimateLChange(waves = 5, variable_type = "bivariate", estimate_constant_change = TRUE),
               data = lcsm_data$data),
        standardized = FALSE)
```

## Summary
The LCSM is a powerful framework for modeling dynamics in change form. By differencing observed scores, it effectively absorbs all time-invariant heterogeneity, including stable between-person differences. The `estimateLChange()` function generates `lavaan` syntax for a variety of LCSM specifications, while `simLChange()` simulates data.

There is a clear algebraic equivalence between the LCSM without constant change and the CLPM in change form. However, the presence of measurement error and/or constant change leads to bias in the CLPM estimates, while the LCSM can recover the true parameters. 


## Appendix: Detailed LCSM DAG

@fig-lcm-full shows the bivariate LCSM with its full measurement model: latent true scores $q_t$ (for $X$) and $p_t$ (for $Y$), constant-change factors $\Delta_x, \Delta_y$, change scores $\Delta x_t, \Delta y_t$, and observed indicators across five waves.

```{r}
#| label: fig-lcm-full
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Bivariate Latent Change Score Model with constant-change factors. Full measurement model across five waves."
library(DiagrammeR)
grViz("
digraph LCM {
  graph [layout = neato, rankdir = TB, bgcolor = white]
  node [shape = box, style = filled, fillcolor = white, fontsize = 11]

  A_X [pos = '-2.5,1!', shape = circle, fillcolor = white, label = <Δ<sub>x</sub>>,  width = 0.5, height = 0.5]
  I [pos = '-2.5,0!', shape = triangle, fillcolor = white, label = 'I',  width = 0.5, height = 0.5]
  X1 [pos = '-2,2.2!', label = 'x1', width = 0.6, height = 0.3]
  X2 [pos = '-1,2.2!', label = 'x2', width = 0.6, height = 0.3]
  X3 [pos = '0,2.2!', label = 'x3', width = 0.6, height = 0.3]
  X4 [pos = '1,2.2!', label = 'x4', width = 0.6, height = 0.3]
  X5 [pos = '2,2.2!', label = 'x5', width = 0.6, height = 0.3]

  wX1 [pos = '-2,1.35!', shape = circle, fillcolor = white, label = 'q1']
  wX2 [pos = '-1,1.35!', shape = circle, fillcolor = white, label = 'q2']
  wX3 [pos = '0,1.35!', shape = circle, fillcolor = white, label = 'q3']
  wX4 [pos = '1,1.35!', shape = circle, fillcolor = white, label = 'q4']
  wX5 [pos = '2,1.35!', shape = circle, fillcolor = white, label = 'q5']

  wY1 [pos = '-2,-1.35!', shape = circle, fillcolor = white, label = 'p1']
  wY2 [pos = '-1,-1.35!', shape = circle, fillcolor = white, label = 'p2']
  wY3 [pos = '0,-1.35!', shape = circle, fillcolor = white, label = 'p3']
  wY4 [pos = '1,-1.35!', shape = circle, fillcolor = white, label = 'p4']
  wY5 [pos = '2,-1.35!', shape = circle, fillcolor = white, label = 'p5']

  dX2 [pos = '-1,0.5!', shape = circle, fillcolor = white, label = 'Δx2']
  dX3 [pos = '0,0.5!', shape = circle, fillcolor = white, label = 'Δx3']
  dX4 [pos = '1,0.5!', shape = circle, fillcolor = white, label = 'Δx4']
  dX5 [pos = '2,0.5!', shape = circle, fillcolor = white, label = 'Δx5']

  dY2 [pos = '-1,-0.5!', shape = circle, fillcolor = white, label = 'Δy2']
  dY3 [pos = '0,-0.5!',  shape = circle, fillcolor = white, label = 'Δy3']
  dY4 [pos = '1,-0.5!',  shape = circle, fillcolor = white, label = 'Δy4']
  dY5 [pos = '2,-0.5!',  shape = circle, fillcolor = white, label = 'Δy5']

  Y1 [pos = '-2,-2.2!', label = 'y1', width = 0.6, height = 0.3]
  Y2 [pos = '-1,-2.2!', label = 'y2', width = 0.6, height = 0.3]
  Y3 [pos = '0,-2.2!', label = 'y3', width = 0.6, height = 0.3]
  Y4 [pos = '1,-2.2!', label = 'y4', width = 0.6, height = 0.3]
  Y5 [pos = '2,-2.2!', label = 'y5', width = 0.6, height = 0.3]
  A_Y [pos = '-2.5,-1!', shape = circle, fillcolor = white, label = <Δ<sub>y</sub>>, width = 0.5, height = 0.5]

  A_X -> dX2 [arrowsize = 0.5, tailport = e, headport = n]
  A_X -> dX3 [arrowsize = 0.5, tailport = e, headport = n]
  A_X -> dX4 [arrowsize = 0.5, tailport = e, headport = n]
  A_X -> dX5 [arrowsize = 0.5, tailport = e, headport = n]
  A_Y -> dY2 [arrowsize = 0.5, tailport = e, headport = s]
  A_Y -> dY3 [arrowsize = 0.5, tailport = e, headport = s]
  A_Y -> dY4 [arrowsize = 0.5, tailport = e, headport = s]
  A_Y -> dY5 [arrowsize = 0.5, tailport = e, headport = s]

  I -> A_X [arrowsize = 0.5, tailport = n, headport = s]
  I -> A_Y [arrowsize = 0.5, tailport = s, headport = n]
  I -> wX1 [arrowsize = 0.5, tailport = n, headport = s]
  I -> wY1 [arrowsize = 0.5, tailport = s, headport = n]

  wX1 -> X1 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX2 -> X2 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX3 -> X3 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX4 -> X4 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX5 -> X5 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY1 -> Y1 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY2 -> Y2 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY3 -> Y3 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY4 -> Y4 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY5 -> Y5 [label = '1', arrowsize = 0.5, fontsize = 11]

  wY1 -> dY2 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wY2 -> dY3 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wY3 -> dY4 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wY4 -> dY5 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wX1 -> dX2 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wX2 -> dX3 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wX3 -> dX4 [label = 'β', arrowsize = 0.5, fontsize = 10]
  wX4 -> dX5 [label = 'β', arrowsize = 0.5, fontsize = 10]

  wX1 -> dY2 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wX2 -> dY3 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wX3 -> dY4 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wX4 -> dY5 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wY1 -> dX2 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wY2 -> dX3 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wY3 -> dX4 [label = 'ω', arrowsize = 0.5, fontsize = 10]
  wY4 -> dX5 [label = 'ω', arrowsize = 0.5, fontsize = 10]

  wX1 -> wX2 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX2 -> wX3 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX3 -> wX4 [label = '1', arrowsize = 0.5, fontsize = 11]
  wX4 -> wX5 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY1 -> wY2 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY2 -> wY3 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY3 -> wY4 [label = '1', arrowsize = 0.5, fontsize = 11]
  wY4 -> wY5 [label = '1', arrowsize = 0.5, fontsize = 11]

  dX2 -> wX2 [arrowsize = 0.5]
  dX3 -> wX3 [arrowsize = 0.5]
  dX4 -> wX4 [arrowsize = 0.5]
  dX5 -> wX5 [arrowsize = 0.5]
  dY2 -> wY2 [arrowsize = 0.5]
  dY3 -> wY3 [arrowsize = 0.5]
  dY4 -> wY4 [arrowsize = 0.5]
  dY5 -> wY5 [arrowsize = 0.5]
}
")
```

