# Extensions and Alternatives

The CLPM (Chapter 1), the LCSM (Chapter 2), and the RI-CLPM (Chapter 3) are special cases of a broader unified model, which @usami2019 organizes into two equations: a **decomposition** component. This is just the way that we can split the observed scores into stable and time-varying components. For instance, the RI-CLPM separates the stable trait from the time-varying deviations around that stable trait. The **dynamic equation** is typically what one is interested in with a panel dataset, a change process. In the RI-CLPM this is represented by the autoregressive and cross-lagged paths on the time-varying deviations. In the LCM, the dynamic equation is the proportional change specified by the latent change scores, what proportion of y changes for every unit change in x. @usami2019 shows that many models share common decomposition and dynamic equations, and these two components can be combined and modified to yield different families of models. This can be a useful exercise -- there are a number of heavily used structural equation models that can be estimated with panel data, and it isn't always clear how the parameters of models related to one another. One commonly estimated model is the **Latent Growth Model** (LGM), which capture each unit's trajectory over time with a latent intercept and slope factor. The RI-CLPM is a special case of the LGM, where the stable factors are random intercepts and there are no slope factors and dynamic paths are specified on the time-varying deviations from those random intercepts. The `crossLagR` includes syntax and simulation functions for all the models in this chapter.


## The Unified Framework

@usami2019 proposes couching these various models in an overarching framework. The notion is to start by acknowledging that observed scores are a combination of stable and time-varying components.  This chapter implements this framework to develop several related SEM estimators for panel data, including the Latent Growth Model (LCM), Autoregressive Latent Trajectory (ALT) model, and the General Cross-Lagged Model (GCLM; also known as the Latent Growth Model with Structured Residuals). Under this framework, these models can also be compared to the CLPM, RI-CLPM, and LCSM, which are special cases of the more general framework.

$$
\begin{aligned}
x_{it} &= \mu_{xt} + \underbrace{F_{x,i}^\top \lambda_{xt}}_{\text{stable}} + x^*_{it}, \\
y_{it} &= \mu_{yt} + \underbrace{F_{y,i}^\top \lambda_{yt}}_{\text{stable}} + y^*_{it}.
\end{aligned}
$$

Here $F_{x,i}, F_{y,i}$ captures an array of stable factors -- these wouldn't specify all at once, but the framework simply allows us to capture an array of things that show up in panel models, that all fall in this bucket: stable components. These are "random intercepts" in the RI-CLPM. They are "growth factors" in the ALT and LCM-SR; they represent the unit specific slope and intercept that capture each unit's trajectory. These are "latent fixed effects" in Bollen & Brand. Importantly, *the loadings $\lambda_{xt}, \lambda_{yt}$ define how these stable factors contribute to the observed scores at each wave*. 

The stable component is important in **decomposition**. It allows us to partition the observed scores into stable and time-varying components. The time-varying within-person component is $x^*_{it}, y^*_{it}$.
$$
\begin{aligned}
x^*_{it} &= \beta_x\,x^*_{i,t-1} + \omega_{yx}\,y^*_{i,t-1} + u_{x,it}, \\
y^*_{it} &= \beta_y\,y^*_{i,t-1} + \omega_{xy}\,x^*_{i,t-1} + u_{y,it}.
\end{aligned}
$$

where $\beta_x, \beta_y$ are the autoregressive coefficients (lavaan labels `ar_x`, `ar_y`) and $\omega_{xy}, \omega_{yx}$ are the cross-lagged coefficients (lavaan labels `cl_xy`, `cl_yx`).[^pq-note]

[^pq-note]: In the `estimateRICLPM()` lavaan syntax, the within-person deviations $x^*_{it}$ and $y^*_{it}$ are written as latent variables `p` and `q`, respectively (i.e., `p1, p2, ...` correspond to $x^*_{i,1}, x^*_{i,2}, \ldots$ and `q1, q2, ...` to $y^*_{i,1}, y^*_{i,2}, \ldots$). The dynamic equation above is therefore implemented as `p ~ ar_x*p + cl_yx*q` and `q ~ ar_y*q + cl_xy*p`.

Each estimator in `crossLagR` largely depend on the stable factor component.  Let's say that any model is defined by three broad categories of parameters: the stable factors $F$, their loadings $\lambda_t$, and the dynamic equation (what goes on in the time-varying component). 

1. **What stable factors $F$?** In the CLPM, there are no stable factors -- all variation is wave-specific. In the RI-CLPM, there are two stable factors -- the random intercepts for x and y. In the ALT and LCM-SR, there are four stable factors -- the intercept and slope for x and y. In Bollen & Brand, there are two stable factors -- the latent fixed effects for x and y.

2. **The trait factors are *expressed* by the loadings $\lambda_t$** This is particularly important in growth models, where the loadings determine how the stable factors contribute to the observed scores at each wave, for instance. 

3. **Dynamic Effects** In the RI-CLPM the dynamic equations are estimated on $x^*$ or directly. In the Autoregressive Latent Trajectory (ALT) model, the AR and CL paths are specified directly the observed scores, which creates a compounding effect over time. And General Cross Cross-Lagged Model which is a growth model, where the residuals are then regressed in a cross lag structure. So, unlike the RI-CLPM the GCLM autoregressive and cross-lagged paths are deviations from a unit's growth trajectory, not deviations from a unit's trait. 

The `crossLagR` package uses identical parameter labels across models, *but the interpretation of these parameters differs by model.* @tbl-unified-labels-short summarizes the unified parameter labels across models, their varying substantive interpretations.

| Unified Label | Parameter | CLPM | RI-CLPM | Bollen & Brand | Latent Change | ALT | LCM-SR |
|:---:|:---|:---|:---|:---|:---|:---|:---|
| `ar_x` | Autoregressive (X) | Rank-order | Within-person carry-over | AR with fixed effects | Proportional change | AR on scores (compounding) | AR on residuals from growth |
| `ar_y` | Autoregressive (Y) | Rank-order | Within-person carry-over | AR with fixed effects | Proportional change | AR on scores (compounding) | AR on residuals from growth |
| `cl_xy` | Cross-lag (X $\to$ Y) | Granger-causal | Within-person cross-lag | Cross-lag with FE | Coupling (X level $\to$ $\Delta$Y) | CL on scores | CL on residuals from growth |
| `cl_yx` | Cross-lag (Y $\to$ X) | Granger-causal | Within-person cross-lag | Cross-lag with FE | Coupling (Y level $\to$ $\Delta$X) | CL on scores | CL on residuals from growth |
| `I_x`, `I_y` | Trait / intercept factors | --- | Random intercepts | Latent fixed effects | --- | Growth intercept (accumulating) | Growth intercept (direct only) |
| `S_x`, `S_y` | Slope factors | --- | --- | --- | --- | Growth slope (accumulating) | Growth slope (direct only) |
| `A_x`, `A_y` | Accumulating change factors | --- | --- | --- | Constant-change factor | --- | --- |

: Unified parameter labels across estimators. All models share `ar_x`, `ar_y`, `cl_xy`, `cl_yx` labels; the substantive interpretation differs. {#tbl-unified-labels}

@tbl-unified-labels-short briefly summarizes this varying interpretation. The CLPM, for instance, does not have a stable trait component, so the AR paths capture rank-order stability (if an individual scores higher than their own average at time 1, they also score higher than their own average at time 2?). The model generally will preserve the rank-order of individuals over time, in $x$ and $y$. 

| Label | By model, what it means...|
| `ar_x`, `ar_y` | CLPM: rank-order stability • RI-CLPM: within-person carry-over • B&B: AR + fixed effects • LCSM: proportional change • ALT: AR on scores (compounds with growth) • LCM-SR: AR on growth residuals |
| `cl_xy`, `cl_yx` | CLPM: Granger-causal • RI-CLPM: within-person cross-lag • B&B: CL + fixed effects • LCSM: level $\to$ $\Delta$ coupling • ALT: CL on scores • LCM-SR: CL on growth residuals |
| `I_x`, `I_y` | RI-CLPM intercepts • B&B fixed-effect latents • ALT/LCM-SR growth intercepts |
| `S_x`, `S_y` | Growth slopes (ALT, LCM-SR only) |
| `A_x`, `A_y` | Accumulating-change factor (LCSM only) |

: Parameter interpretations by model. {#tbl-unified-labels-short}


```{r echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(crossLagR)
  library(dplyr)
  library(lavaan)
  library(DiagrammeR)
})
```

## The Latent Growth Model

The LGM is a different style of model, in that each unit has an intercept $I$ (initial level) and slope $S$ (linear rate of change).In the bivariate form, $(I_x, S_x)$ and $(I_y, S_y)$ covary across variables but no lagged paths connect $x$ to $y$. This is the natural baseline for the ALT and LCM-SR, both of which add a dynamic equation on top of an LGM-style stable component.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-lgm
#| fig-cap: "Bivariate Latent Growth Model. Growth factors (I, S) load directly on observed scores with fixed loadings (intercept = 1; slope = 0, 1, 2, 3); no autoregressive or cross-lagged paths."
grViz("
digraph LGM {
  graph [layout = neato, bgcolor = white]
  node [shape = box, style = filled, fillcolor = white, fontsize = 11]

  I_X [pos = '-4.5,2.6!', shape = circle, label = <I<sub>x</sub>>, width = 0.5, height = 0.5]
  S_X [pos = '-3.0,3.6!', shape = circle, label = <S<sub>x</sub>>, width = 0.5, height = 0.5]
  X1 [pos = '-1.5,1.8!', label = 'x1']
  X2 [pos = '-0.5,1.8!', label = 'x2']
  X3 [pos = '0.5,1.8!',  label = 'x3']
  X4 [pos = '1.5,1.8!',  label = 'x4']
  Y1 [pos = '-1.5,-1.8!', label = 'y1']
  Y2 [pos = '-0.5,-1.8!', label = 'y2']
  Y3 [pos = '0.5,-1.8!',  label = 'y3']
  Y4 [pos = '1.5,-1.8!',  label = 'y4']
  I_Y [pos = '-4.5,-2.6!', shape = circle, label = <I<sub>y</sub>>, width = 0.5, height = 0.5]
  S_Y [pos = '-3.0,-3.6!', shape = circle, label = <S<sub>y</sub>>, width = 0.5, height = 0.5]

  I_X -> X1 [label = '1', arrowsize = 0.5, fontsize = 10]
  I_X -> X2 [label = '1', arrowsize = 0.5, fontsize = 10]
  I_X -> X3 [label = '1', arrowsize = 0.5, fontsize = 10]
  I_X -> X4 [label = '1', arrowsize = 0.5, fontsize = 10]
  S_X -> X1 [label = '0', arrowsize = 0.5, fontsize = 10]
  S_X -> X2 [label = '1', arrowsize = 0.5, fontsize = 10]
  S_X -> X3 [label = '2', arrowsize = 0.5, fontsize = 10]
  S_X -> X4 [label = '3', arrowsize = 0.5, fontsize = 10]
  I_Y -> Y1 [label = '1', arrowsize = 0.5, fontsize = 10]
  I_Y -> Y2 [label = '1', arrowsize = 0.5, fontsize = 10]
  I_Y -> Y3 [label = '1', arrowsize = 0.5, fontsize = 10]
  I_Y -> Y4 [label = '1', arrowsize = 0.5, fontsize = 10]
  S_Y -> Y1 [label = '0', arrowsize = 0.5, fontsize = 10]
  S_Y -> Y2 [label = '1', arrowsize = 0.5, fontsize = 10]
  S_Y -> Y3 [label = '2', arrowsize = 0.5, fontsize = 10]
  S_Y -> Y4 [label = '3', arrowsize = 0.5, fontsize = 10]

  I_X -> I_Y [dir = both, style = dashed, arrowsize = 0.4]
  S_X -> S_Y [dir = both, style = dashed, arrowsize = 0.4]
  I_X -> S_X [dir = both, style = dashed, arrowsize = 0.4]
  I_Y -> S_Y [dir = both, style = dashed, arrowsize = 0.4]
}
")
```

::: {.callout-note title="crossLagR syntax: `estimateLGM()`"}
```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateLGM(waves = 4, variable_type = "bivariate") |> cat()
```
:::

## ALT: Autoregressive Latent Trajectory

The ALT is a combination of the LGM and the CLPM. Each wave score (x and y) is a combination of a growth trajectory (as in the LGM) and a dynamic process (as in the CLPM). The problem with the ALT is that the trajectories and dynamics are not cleanly separated, and the growth factors compound with the AR and CL paths (@usami2021, @usami2022). The AR and CL parameters may thus be different from the CLPM or RI-CLPM.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-alt
#| fig-cap: "Bivariate ALT. Growth factors (I, S) load on observed scores; AR (β) and CL (ω) operate directly on the scores, so growth and dynamics compound."
grViz("
digraph ALT {
  graph [layout = neato, bgcolor = white, splines = line]
  node [shape = box, style = filled, fillcolor = white, fontsize = 11]

  I_X [pos = '-3.5,2.4!', shape = circle, label = <I<sub>x</sub>>, width = 0.5, height = 0.5]
  S_X [pos = '-3.5,1.4!', shape = circle, label = <S<sub>x</sub>>, width = 0.5, height = 0.5]
  X1 [pos = '-1.5,2.0!', label = 'x1']
  X2 [pos = '0.0,2.0!',  label = 'x2']
  X3 [pos = '1.5,2.0!',  label = 'x3']
  X4 [pos = '3.0,2.0!',  label = 'x4']
  Y1 [pos = '-1.5,-2.0!', label = 'y1']
  Y2 [pos = '0.0,-2.0!',  label = 'y2']
  Y3 [pos = '1.5,-2.0!',  label = 'y3']
  Y4 [pos = '3.0,-2.0!',  label = 'y4']
  I_Y [pos = '-3.5,-1.4!', shape = circle, label = <I<sub>y</sub>>, width = 0.5, height = 0.5]
  S_Y [pos = '-3.5,-2.4!', shape = circle, label = <S<sub>y</sub>>, width = 0.5, height = 0.5]

  I_X -> X1 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_X -> X2 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_X -> X3 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_X -> X4 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_X -> X1 [label = '0', arrowsize = 0.4, fontsize = 9]
  S_X -> X2 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_X -> X3 [label = '2', arrowsize = 0.4, fontsize = 9]
  S_X -> X4 [label = '3', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y1 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y2 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y3 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y4 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y1 [label = '0', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y2 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y3 [label = '2', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y4 [label = '3', arrowsize = 0.4, fontsize = 9]

  X1 -> X2 [label = <β>, arrowsize = 0.5, fontsize = 12]
  X2 -> X3 [label = <β>, arrowsize = 0.5, fontsize = 12]
  X3 -> X4 [label = <β>, arrowsize = 0.5, fontsize = 12]
  Y1 -> Y2 [label = <β>, arrowsize = 0.5, fontsize = 12]
  Y2 -> Y3 [label = <β>, arrowsize = 0.5, fontsize = 12]
  Y3 -> Y4 [label = <β>, arrowsize = 0.5, fontsize = 12]
  X1 -> Y2 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  X2 -> Y3 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  X3 -> Y4 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  Y1 -> X2 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  Y2 -> X3 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  Y3 -> X4 [label = <ω>, arrowsize = 0.5, fontsize = 12]

  I_X -> I_Y [dir = both, style = dashed, arrowsize = 0.4]
  S_X -> S_Y [dir = both, style = dashed, arrowsize = 0.4]
}
")
```

::: {.callout-note title="crossLagR syntax: `estimateALT()`"}
```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateALT(waves           = 4,
            constrain_beta  = TRUE,
            constrain_omega = TRUE) |>
  cat()
```
:::

## GCLM

The General Cross-Lagged Model (GCLM; also known as the Latent Growth Model with Structured Residuals) is similar to the ALT, in that it combines a growth model with a dynamic equation. However, in the GCLM, the growth factors have direct effects on the observed scores, and the AR and CL paths are specified on the residuals from that growth model. The AR and CL parameters capture the dynamic process in the deviations from each person's trajectory, rather than deviations from a stable trait (as in the RI-CLPM).

```{r echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-gclm
#| fig-cap: "Bivariate LCM-SR / GCLM. Growth factors (I, S) have direct effects only on observed scores; AR (β) and CL (ω) act on the structured residuals (q, p) — deviations from each person's trajectory."
grViz("
digraph GCLM {
  graph [layout = neato, bgcolor = white, splines = line]
  node [shape = box, style = filled, fillcolor = white, fontsize = 11]

  I_X [pos = '-3.5,3.0!', shape = circle, label = <I<sub>x</sub>>, width = 0.5, height = 0.5]
  S_X [pos = '-3.5,2.0!', shape = circle, label = <S<sub>x</sub>>, width = 0.5, height = 0.5]
  X1 [pos = '-1.5,2.6!', label = 'x1']
  X2 [pos = '0.0,2.6!',  label = 'x2']
  X3 [pos = '1.5,2.6!',  label = 'x3']
  X4 [pos = '3.0,2.6!',  label = 'x4']

  qX1 [pos = '-1.5,1.0!', shape = circle, label = 'q1', width = 0.4, height = 0.4]
  qX2 [pos = '0.0,1.0!',  shape = circle, label = 'q2', width = 0.4, height = 0.4]
  qX3 [pos = '1.5,1.0!',  shape = circle, label = 'q3', width = 0.4, height = 0.4]
  qX4 [pos = '3.0,1.0!',  shape = circle, label = 'q4', width = 0.4, height = 0.4]

  pY1 [pos = '-1.5,-1.0!', shape = circle, label = 'p1', width = 0.4, height = 0.4]
  pY2 [pos = '0.0,-1.0!',  shape = circle, label = 'p2', width = 0.4, height = 0.4]
  pY3 [pos = '1.5,-1.0!',  shape = circle, label = 'p3', width = 0.4, height = 0.4]
  pY4 [pos = '3.0,-1.0!',  shape = circle, label = 'p4', width = 0.4, height = 0.4]

  Y1 [pos = '-1.5,-2.6!', label = 'y1']
  Y2 [pos = '0.0,-2.6!',  label = 'y2']
  Y3 [pos = '1.5,-2.6!',  label = 'y3']
  Y4 [pos = '3.0,-2.6!',  label = 'y4']
  I_Y [pos = '-3.5,-2.0!', shape = circle, label = <I<sub>y</sub>>, width = 0.5, height = 0.5]
  S_Y [pos = '-3.5,-3.0!', shape = circle, label = <S<sub>y</sub>>, width = 0.5, height = 0.5]

  I_X -> X1 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_X -> X2 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_X -> X3 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_X -> X4 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_X -> X1 [label = '0', arrowsize = 0.4, fontsize = 9]
  S_X -> X2 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_X -> X3 [label = '2', arrowsize = 0.4, fontsize = 9]
  S_X -> X4 [label = '3', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y1 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y2 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y3 [label = '1', arrowsize = 0.4, fontsize = 9]
  I_Y -> Y4 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y1 [label = '0', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y2 [label = '1', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y3 [label = '2', arrowsize = 0.4, fontsize = 9]
  S_Y -> Y4 [label = '3', arrowsize = 0.4, fontsize = 9]

  qX1 -> X1 [label = '1', arrowsize = 0.4, fontsize = 9]
  qX2 -> X2 [label = '1', arrowsize = 0.4, fontsize = 9]
  qX3 -> X3 [label = '1', arrowsize = 0.4, fontsize = 9]
  qX4 -> X4 [label = '1', arrowsize = 0.4, fontsize = 9]
  pY1 -> Y1 [label = '1', arrowsize = 0.4, fontsize = 9]
  pY2 -> Y2 [label = '1', arrowsize = 0.4, fontsize = 9]
  pY3 -> Y3 [label = '1', arrowsize = 0.4, fontsize = 9]
  pY4 -> Y4 [label = '1', arrowsize = 0.4, fontsize = 9]

  qX1 -> qX2 [label = <β>, arrowsize = 0.5, fontsize = 12]
  qX2 -> qX3 [label = <β>, arrowsize = 0.5, fontsize = 12]
  qX3 -> qX4 [label = <β>, arrowsize = 0.5, fontsize = 12]
  pY1 -> pY2 [label = <β>, arrowsize = 0.5, fontsize = 12]
  pY2 -> pY3 [label = <β>, arrowsize = 0.5, fontsize = 12]
  pY3 -> pY4 [label = <β>, arrowsize = 0.5, fontsize = 12]
  qX1 -> pY2 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  qX2 -> pY3 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  qX3 -> pY4 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  pY1 -> qX2 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  pY2 -> qX3 [label = <ω>, arrowsize = 0.5, fontsize = 12]
  pY3 -> qX4 [label = <ω>, arrowsize = 0.5, fontsize = 12]

  I_X -> I_Y [dir = both, style = dashed, arrowsize = 0.4]
  S_X -> S_Y [dir = both, style = dashed, arrowsize = 0.4]
}
")
```

::: {.callout-note title="crossLagR syntax: `estimateLCMSR()`"}
```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateLCMSR(waves           = 4,
              constrain_beta  = TRUE,
              constrain_omega = TRUE) |>
  cat()
```
:::

## Bollen & Brand: Dynamic Panel with Latent Fixed Effects

@bollen_general_2010 propose a dynamic panel model in which a latent factor absorbs the unit-specific fixed effect — it forms an SEM analogue of Arellano-Bond [see also @dishop_tutorial_2018, the practical tutorial implementation that `estimateBollen_and_Brand()` follows].

The decomposition equation is important. In the RI-CLPM, the stable factor is a random intercept. Each unit is drawn from a distribution of units. The model captures unobserved heterogeneity across units. But a critical identifying assumption in the RI-CLPM is that the stable factors is uncorrelated with the time varying components. If this is violated, the RI-CLPM estimates are biased. 

A useful alternative — and one with the same spirit as the well-known Arellano-Bond GMM estimator, though here estimated by maximum likelihood [@bollen_general_2010; @dishop_tutorial_2018] — is to specify the stable factor as a latent unit-effect that loads on every wave of the *observed* CLPM equation, while **freely estimating its covariance with the lagged predictors**. It's this part that is critical: Letting that covariance be non-zero is the relaxes the orthogonality assumption that the RI-CLPM imposes, and it is exactly the source of the dynamic-panel bias identified by Nickell (1981) clearly elaborated upon in influential texbooks ([@hsiao2022]) And @ludtke2022 show empirically that this matters — the Bollen-Brand specification and the RI-CLPM can yield different cross-lag estimates on the same data because of the departures in how the dynamic equation is specified.

The `crossLagR` function `estimateBollen_and_Brand()` implements the specification. Including all the covariances.

::: {.callout-note title="crossLagR syntax: `estimateBollen_and_Brand()`"}
```{r echo=TRUE, message=FALSE, warning=FALSE}
estimateBollen_and_Brand(waves = 4) |> cat()
```
:::

## Comparing on One Dataset

Fitting several estimators to one dataset is mechanical because the lavaan labels are unified: `ar_x`, `ar_y`, `cl_xy`, `cl_yx` mean *the AR/CL parameters of that particular model* and can be pulled directly from `parameterEstimates()`. The problem with this -- of course, is that the estimates are not directly comparable across models, because they have different interpretations. 

Let's simulate data from a RI-CLPM DGP, and fit all the models to it. So, RI-CLPM is will fit well, and the truth table is:

| Parameter | True value | What it means in the DGP |
|:---|:---:|:---|
| `ar_x` | 0.30 | Within-person carry-over of $x^*$ |
| `ar_y` | 0.30 | Within-person carry-over of $y^*$ |
| `cl_xy` | 0.20 | Within-person cross-lag, X $\to$ Y |
| `cl_yx` | 0.10 | Within-person cross-lag, Y $\to$ X |
| `var(I_x)` | 1.00 | Trait variance for X |
| `var(I_y)` | 1.00 | Trait variance for Y |
| `var(p)`, `var(q)` | 1.00 | Within-person dynamic-residual variance |




```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(2026)
dat <- simRICLPM(waves       = 5,
                 beta_x      = 0.3,
                 beta_y      = 0.3,
                 omega_xy    = 0.2,
                 omega_yx    = 0.1,
                 var_BX      = 1.0,
                 var_BY      = 1.0,
                 var_p       = 1.0,
                 var_q       = 1.0,
                 sample.nobs = 1000,
                 estimate_means = TRUE)$data

pick <- function(fit, lab) {
  if (is.null(fit) || !lavInspect(fit, "converged")) return(NA_real_)
  pe <- parameterEstimates(fit)
  v  <- pe$est[pe$label == lab]
  if (length(v)) v[1] else NA_real_
}

fit_one <- function(name, syntax) {
  fit <- tryCatch(lavaan(syntax, data = dat,
                         warn = FALSE, verbose = FALSE),
                  error = function(e) NULL)
  tibble(model = name,
         ar_x  = pick(fit, "ar_x"),
         ar_y  = pick(fit, "ar_y"),
         cl_xy = pick(fit, "cl_xy"),
         cl_yx = pick(fit, "cl_yx"))
}

bind_rows(
  fit_one("CLPM",       estimateCLPM(waves = 5)),
  fit_one("RI-CLPM",    estimateRICLPM(waves = 5)),
  fit_one("LCSM",       estimateLChange(waves = 5, variable_type = "bivariate")),
  fit_one("ALT",        estimateALT(waves = 5)),
  fit_one("LCM-SR",     estimateLCMSR(waves = 5)),
  fit_one("Bollen-Brand", estimateBollen_and_Brand(waves = 5,
                                                   x_effect = "lagged",
                                                   x_autoregression = TRUE,
                                                   y_effect_on_x = TRUE,
                                                   constrain_coefficients = TRUE))
) |>
  mutate(across(where(is.numeric), ~ round(., 3)))
```

It's useful to see the model similarities, but it's also important to interpret the estimates with caution. The AR coefficient means *carry-over on observed scores* in the CLPM, *carry-over on within-person deviations* in the RI-CLPM, *proportional change* in the LCSM (effective AR is $1 + \beta_x$), *compounding carry-over* in the ALT, *carry-over in residuals from growth* in the LCM-SR, and *carry-over net of a latent fixed effect* in Bollen & Brand. 

## Summary

@usami2019 introduce a useful framework for understanding the relationships between different panel data models. The framework follows from from two key equations, the **decomposition equation** of observed scores into stable and time-varying components, and the **dynamic equation** that captures how the time-varying components evolve over time. All models in this chapter can be expressed in the same framework, the differences between them are in how the stable component is defined, and how the dynamic equation is specified -- e.g,, on the observed scores, or deviations from a stable trait or growth trajectory. 