# Simulations

The previous chapters established that the CLPM, RI-CLPM, LCSM, ALT, LCM-SR, and Bollen & Brand all share the same four AR/CL labels but differ in how they decompose observed variance. The substantive question is **empirical**: when each model is fit to data that was *not* generated under its own assumptions, what does it recover?

This chapter answers that question with a focused simulation experiment.

**Simulation — Heterogeneity.** Data are generated from an RI-CLPM and the between-to-within variance ratio is varied from `0.25` (mostly within-person noise) to `9` (mostly stable trait). All six estimators are fit to each replication.

All AR/CL parameters in the DGP are held at the same target values, so the only thing changing across conditions is the *amount* of between-person heterogeneity.

| Parameter | Label | True Value |
|:---|:---:|:---:|
| AR(X) | `ar_x` | 0.30 |
| AR(Y) | `ar_y` | 0.20 |
| CL(X $\to$ Y) | `cl_xy` | 0.15 |
| CL(Y $\to$ X) | `cl_yx` | 0.10 |

: True AR/CL parameters shared across both simulations. Within-person dynamic residual variance is fixed at 0.5. {#tbl-true-params}

```{r}
#| label: setup
#| message: false
#| warning: false
#| cache: false
#| echo: false
#|
suppressPackageStartupMessages({
  library(crossLagR)
  library(lavaan)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggridges)
  library(gt)
  library(here)
})
devtools::load_all(quiet = TRUE)

waves      <- 4
n_obs      <- 1000
within_var <- 0.5
trials     <- 100

true_vals <- c(ar_x = 0.30, ar_y = 0.20, cl_xy = 0.15, cl_yx = 0.10)
bw_ratios <- c(0.25, 0.5, 1, 2, 4, 9)   # between-to-within variance ratios: This is what's varied

# recycle ggplot theme. 
theme_grand <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.2, color = "grey85"),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text       = element_text(face = "bold"),
    legend.title     = element_text(face = "bold"),
    legend.position  = "right",
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(color = "grey40", size = 10)
  )
```

## Population Models

Each DGP is a population `lavaan` parameter table built by piping the estimator's syntax through `lavaanify()`, then overwriting the variance components.

```{r}
#| label: dgp-helpers
#| message: false
#| warning: false
#| echo: false
#| 
# Set a parameter in a lavaan parameter table by label.
set_param <- function(pt, label, val) {
  pt$ustart[pt$label == label] <- val
  pt
}

# Set a free covariance entry (lhs ~~ rhs) by name pair (symmetric).
set_cov <- function(pt, lhs, rhs, val) {
  rows <- pt$op == "~~" & pt$free > 0 &
          ((pt$lhs == lhs & pt$rhs == rhs) | (pt$lhs == rhs & pt$rhs == lhs))
  pt$ustart[rows] <- val
  pt
}

# RI-CLPM population: AR/CL fixed at truth; trait variance scales with bw_ratio.
make_riclpm_pop <- function(bw_ratio) {
  pt <- lavaanify(estimateRICLPM(waves = waves))
  for (nm in names(true_vals)) pt <- set_param(pt, nm, true_vals[[nm]])
  pt <- set_param(pt, "d_var_x", within_var)
  pt <- set_param(pt, "d_var_y", within_var)
  pt <- set_param(pt, "d_cov_xy", 0.05)

  bw_var <- bw_ratio * within_var
  pt <- set_cov(pt, "I_x", "I_x", bw_var)
  pt <- set_cov(pt, "I_y", "I_y", bw_var)
  pt <- set_cov(pt, "I_x", "I_y", 0.3 * bw_var)
  pt
}
```

## Estimators and Fit Helper

```{r}
#| label: estimators
#| message: false
#| warning: false
#| echo: false 
#| 
estimators <- list(
  CLPM             = estimateCLPM(waves = waves),
  `RI-CLPM`        = estimateRICLPM(waves = waves),
  `Bollen & Brand` = estimateBollen_and_Brand(
    waves = waves, x_effect = "lagged",
    x_autoregression = TRUE, y_effect_on_x = TRUE,
    constrain_coefficients = TRUE),
  `Latent Change`  = estimateLChange(
    waves = waves, variable_type = "bivariate",
    estimate_constant_change = FALSE),
  ALT              = estimateALT(waves = waves),
  `LCM-SR`         = estimateLCMSR(waves = waves)
)

ARCL_LABELS      <- c("ar_x", "ar_y", "cl_xy", "cl_yx")
NEEDS_MEANSTRUCT <- c("ALT", "LCM-SR")

# Fit one estimator; return one row per AR/CL parameter + diagnostic flags.
fit_one <- function(name, dat) {
  warns <- character(0)
  fit <- tryCatch(
    withCallingHandlers(
      lavaan::lavaan(estimators[[name]], data = dat,
                     meanstructure = name %in% NEEDS_MEANSTRUCT),
      warning = function(w) {
        warns <<- c(warns, conditionMessage(w)); invokeRestart("muffleWarning")
      }),
    error = function(e) NULL)

  if (is.null(fit) || !lavInspect(fit, "converged")) {
    return(tibble(estimator = name, parameter = ARCL_LABELS,
                  estimate = NA_real_, converged = FALSE,
                  n_warnings = length(warns), heywood = NA))
  }

  pe       <- parameterEstimates(fit)
  estimate <- vapply(ARCL_LABELS, function(l) {
    v <- pe$est[pe$label == l]; if (length(v)) v[1] else NA_real_
  }, numeric(1))
  heywood  <- any(pe$est[pe$op == "~~" & pe$lhs == pe$rhs] < 0, na.rm = TRUE)

  tibble(estimator = name, parameter = ARCL_LABELS, estimate = estimate,
         converged = TRUE, n_warnings = length(warns), heywood = heywood)
}

fit_all <- function(dat) bind_rows(lapply(names(estimators), fit_one, dat = dat))
```

## Simulation 1: Varying Between-Person Heterogeneity

Data are generated from the RI-CLPM at each `bw_ratio` $\in \{0.25, 0.5, 1, 2, 4, 9\}$. Each cell is replicated `r trials` times. All six estimators are fit to each replication.

```{r}
#| label: sim1-run
#| message: false
#| warning: false
#| cache: false
#| echo: false

sim1_file <- here::here("data", "sim1_riclpm_bw.rds")

if (!file.exists(sim1_file)) {
  set.seed(2026)
  sim1 <- vector("list", length(bw_ratios) * trials)
  k <- 1L
  for (bw in bw_ratios) {
    pop <- make_riclpm_pop(bw)
    for (j in seq_len(trials)) {
      dat <- tryCatch(simulateData(pop, sample.nobs = n_obs), error = function(e) NULL)
      if (is.null(dat)) next
      res <- fit_all(dat)
      res$bw_ratio <- bw
      res$trial    <- j
      res$true     <- true_vals[res$parameter]
      sim1[[k]] <- res
      k <- k + 1L
    }
  }
  sim1 <- bind_rows(sim1)
  if (!dir.exists(dirname(sim1_file))) dir.create(dirname(sim1_file), recursive = TRUE)
  saveRDS(sim1, sim1_file)
} else {
  sim1 <- readRDS(sim1_file)
}

sim1 <- sim1 |>
  mutate(
    bw_ratio_f = factor(bw_ratio, levels = sort(unique(bw_ratio))),
    parameter  = factor(parameter, levels = c("ar_x", "ar_y", "cl_xy", "cl_yx")),
    estimator  = factor(estimator,
                        levels = c("CLPM", "Latent Change", "RI-CLPM",
                                   "Bollen & Brand", "ALT", "LCM-SR"))
  )
```

### Distributions of Estimates

@fig-sim1-ridges shows the sampling distribution of each estimator's recovery of the four AR/CL parameters. Vertical lines mark the true value; each ridge is colored by its mean absolute deviation from truth — the more green the closer to the truth, red ridges drift far from the true value. One important caveat: The LCGM autoregressive parameters drift far from true parameter values, but the AR parameter should be interpreted as $1+\beta$ where $\beta$ is the true slope generated from the RI-CLPM. 

```{r}
#| label: fig-sim1-ridges
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Sampling distribution of AR/CL estimates under the RI-CLPM DGP, by between:within variance ratio. Columns are parameters, rows are estimators. The vertical dashed line marks the true value. Fill encodes mean |estimate − truth| for that cell: green = on target, red = far from truth."
#| fig-height: 11
#| fig-width: 13

truth_df <- tibble(parameter = factor(names(true_vals), levels = c("ar_x", "ar_y", "cl_xy", "cl_yx")),
                   truth     = true_vals)

sim1 |>
  filter(converged, !is.na(estimate)) |>
  group_by(estimator, parameter, bw_ratio_f) |>
  mutate(mean_abs_dev = mean(abs(estimate - true), na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = estimate, y = bw_ratio_f, fill = mean_abs_dev)) +
  geom_density_ridges(scale = 1.4, rel_min_height = 0.01,
                      color = "white", linewidth = 0.2, alpha = 0.9) +
  geom_vline(data = truth_df, aes(xintercept = truth),
             color = "grey20", linetype = "dashed", linewidth = 0.4) +
  facet_grid(estimator ~ parameter, scales = "free_x", switch = "y") +
  scale_fill_gradient(low = "#2ecc71", high = "#e74c3c",
                      name = "Mean\n|estimate − truth|") +
  labs(x = "Estimate", y = "Between : within variance ratio",
       title = "Simulation 1: Varying Trait Variance under an RI-CLPM DGP",
       subtitle = paste0(trials, " replications per cell · n = ", n_obs, " · ", waves, " waves")) +
  theme_grand +
  theme(strip.placement = "outside",
        panel.spacing   = unit(0.4, "lines"))
```

### Bias and Precision

@fig-sim1-bias collapses the distributions into mean bias with 95% MC intervals. It's also pretty revealing. The CLPM breaks down with a moderate amount of trait variance. Also the spikes that appear for some are because only a handful of models actually converge, so these are means of only a handful of actual estimates. 

```{r}
#| label: fig-sim1-bias
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Bias of AR/CL estimates as a function of between:within variance ratio (RI-CLPM DGP). Points are means across replications; vertical segments are 95% Monte Carlo intervals. The dashed line at zero is unbiased."
#| fig-height: 7
#| fig-width: 13

sim1_bias <- sim1 |>
  filter(converged, !is.na(estimate)) |>
  mutate(error = estimate - true) |>
  group_by(estimator, parameter, bw_ratio) |>
  summarise(
    mean_bias = mean(error, na.rm = TRUE),
    ci_low    = quantile(error, 0.025, na.rm = TRUE),
    ci_high   = quantile(error, 0.975, na.rm = TRUE),
    .groups   = "drop"
  )

parameter_labels <- c(
  ar_x  = "AR(X)",
  ar_y  = "AR(Y)",
  cl_xy = "CL(X→Y)",
  cl_yx = "CL(Y→X)"
)

ggplot(sim1_bias, aes(x = bw_ratio, y = mean_bias)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_line(aes(group = 1), color = "grey60", linewidth = 0.3) +
  geom_linerange(aes(ymin = ci_low, ymax = ci_high),
                 color = "grey20", linewidth = 0.6) +
  geom_point(size = 2.2, alpha = 0.6, color = "black") +
  facet_grid(parameter ~ estimator, switch = "y",
             labeller = labeller(parameter = parameter_labels)) +
  scale_x_continuous(trans = "log10", breaks = bw_ratios) +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(x = "Between:within variance ratio (log scale)",
       y = "Bias (estimate − truth)",
       title = "Simulation 1: Bias by Estimator and Parameter") +
  theme_grand +
  theme(panel.spacing   = unit(0.3, "lines"),
        strip.placement = "outside",
        axis.text.x     = element_text(angle = 45, hjust = 1))
```

The RMSEA estimates vary wildly across models [@fig-sim1-rmse], and the CLPM generally fits poorly. Though most models collapse when there is effectively only trait variance. 

@fig-sim1-rmse

```{r}
#| label: fig-sim1-rmse
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Root mean squared error of AR/CL estimates under the RI-CLPM DGP, by between:within variance ratio. Lower is better. The RI-CLPM line should stay flat; misspecified estimators should climb as the trait variance grows."
#| fig-height: 7
#| fig-width: 13

sim1_rmse <- sim1 |>
  filter(converged, !is.na(estimate)) |>
  group_by(estimator, parameter, bw_ratio) |>
  summarise(rmse = sqrt(mean((estimate - true)^2, na.rm = TRUE)), .groups = "drop")

ggplot(sim1_rmse, aes(x = bw_ratio, y = rmse, color = estimator)) +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  facet_wrap(~ parameter, nrow = 1,
             labeller = labeller(parameter = parameter_labels)) +
  scale_x_continuous(trans = "log10", breaks = bw_ratios) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(9, "YlOrRd")[4:9],
    name   = NULL,
    guide  = guide_legend(override.aes = list(linewidth = 1.4, size = 3))) +
  labs(x = "Between : within variance ratio (log scale)",
       y = "RMSE",
       title    = "Simulation 1: RMSE by Estimator",
       subtitle = "Dashed line at RMSE = 0.05 (typical 'small bias' cutoff).") +
  theme_grand +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Convergence

@tbl-sim1-convergence displays the convergence statistics. This is an important indicator of each models ability to operate in extreme conditions, but conditions that are quite common in panel data in the social sciences. The convergence rate is simply the proportion of replications that produce parameter estimates. The Heywood rate is the proportion of *converged* replications that produce at least one negative variance estimate. These errors are quite common in over parameterized models. The ALT and LCM-SR have very high rates of Heywood cases, especially as the trait variance grows. 

```{r}
#| label: tbl-sim1-convergence
#| echo: false
#| message: false

sim1 |>
  distinct(estimator, bw_ratio, trial, converged, heywood, n_warnings) |>
  group_by(estimator, bw_ratio) |>
  summarise(
    n_trials  = n(),
    conv_rate = mean(converged),
    heywood_rate = mean(heywood, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(c(conv_rate, heywood_rate), ~ round(., 3))) |>
  pivot_wider(id_cols = bw_ratio,
              names_from = estimator,
              values_from = c(conv_rate, heywood_rate),
              names_glue = "{estimator}_{.value}") |>
  gt() |>
  tab_header(title = "Simulation 1: Convergence and Heywood Rates",
             subtitle = paste0(trials, " replications · n = ", n_obs))
```

## Bayesian RI-CLPM with Informative Priors

One of the issues with the RI-CLPM is that it can be difficult to fit when there is a lot of trait variance. The Bayesian approach allows us to put informative priors on the parameters.

```{r}
#| echo: false
#| message: false
#| warning: false
make_riclpm_pop(8) |> head()
```

```{r}
#| label: bayes-riclpm-demo
#| message: false
#| warning: false
#| cache: false
#| echo: false
#| eval: !expr requireNamespace("blavaan", quietly = TRUE) && Sys.getenv("CI") != "true"

library(blavaan)

riclpm_syn_bayes_base <- estimateRICLPM(
  waves = waves, constrain_beta = FALSE, constrain_omega = FALSE,
  constrain_residual_variances = FALSE, constrain_residual_covariances = FALSE
)

add_riclpm_priors <- function(syn) {
  syn <- gsub('(ar_x[0-9]*)\\*',  'prior("normal(0.3,0.15)")*\\1*', syn, perl = TRUE)
  syn <- gsub('(ar_y[0-9]*)\\*',  'prior("normal(0.3,0.15)")*\\1*', syn, perl = TRUE)
  syn <- gsub('(cl_xy[0-9]*)\\*', 'prior("normal(0,0.1)")*\\1*',    syn, perl = TRUE)
  syn <- gsub('(cl_yx[0-9]*)\\*', 'prior("normal(0,0.1)")*\\1*',    syn, perl = TRUE)
  syn <- gsub('(d_var_x[0-9]*)\\*', 'prior("gamma(2,8)")[sd]*\\1*', syn, perl = TRUE)
  syn <- gsub('(d_var_y[0-9]*)\\*', 'prior("gamma(2,8)")[sd]*\\1*', syn, perl = TRUE)
  syn
}

riclpm_syn_bayes <- add_riclpm_priors(riclpm_syn_bayes_base)

set.seed(2026)
n_reps_bayes <- 50
pop_extreme  <- make_riclpm_pop(9)
bayes_file   <- here::here("data", "bayes_summary.rds")

if (!file.exists(bayes_file)) {
  bayes_results <- vector("list", n_reps_bayes)
  for (i in seq_len(n_reps_bayes)) {
    dat <- lavaan::simulateData(pop_extreme, sample.nobs = n_obs)
    ml_warns <- character(0)
    ml_fit <- tryCatch(
      withCallingHandlers(
        lavaan::lavaan(estimateRICLPM(waves = waves), data = dat),
        warning = function(w) { ml_warns <<- c(ml_warns, conditionMessage(w)); invokeRestart("muffleWarning") }
      ),
      error = function(e) NULL
    )
    ml_ok <- !is.null(ml_fit) && lavInspect(ml_fit, "converged")
    ml_heywood <- if (ml_ok) {
      pe <- parameterEstimates(ml_fit)
      any(pe$est[pe$op == "~~" & pe$lhs == pe$rhs] < 0, na.rm = TRUE)
    } else NA

    bayes_error <- NA_character_
    bayes_fit <- tryCatch(
      suppressWarnings(suppressMessages(
        blavaan::bsem(riclpm_syn_bayes, data = dat,
                      n.chains = 2, burnin = 500, sample = 500,
                      bcontrol = list(cores = 2), seed = 1000 + i)
      )),
      error = function(e) { bayes_error <<- conditionMessage(e); NULL }
    )
    bayes_ok <- !is.null(bayes_fit) &&
      isTRUE(tryCatch(blavaan::blavInspect(bayes_fit, "converged"), error = function(e) FALSE))
    bayes_heywood <- if (bayes_ok) {
      pe <- parameterEstimates(bayes_fit)
      any(pe$est[pe$op == "~~" & pe$lhs == pe$rhs] < 0, na.rm = TRUE)
    } else NA

    bayes_results[[i]] <- tibble(
      rep = i, ml_converged = ml_ok, ml_heywood = ml_heywood,
      ml_warnings = length(ml_warns),
      bayes_finished = bayes_ok, bayes_heywood = bayes_heywood,
      bayes_error = bayes_error
    )
  }
  bayes_summary <- bind_rows(bayes_results)
  saveRDS(bayes_summary, bayes_file)
} else {
  bayes_summary <- readRDS(bayes_file)
}

bayes_summary |>
  summarise(
    Replications         = n(),
    `ML converged`       = mean(ml_converged),
    `ML Heywood (of conv.)` = mean(ml_heywood[ml_converged], na.rm = TRUE),
    `ML mean # warnings` = mean(ml_warnings),
    `Bayes finished`     = mean(bayes_finished, na.rm = TRUE),
    `Bayes Heywood (of finished)` = dplyr::if_else(
      sum(bayes_finished, na.rm = TRUE) > 0,
      mean(bayes_heywood[bayes_finished], na.rm = TRUE),
      NA_real_
    )
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |>
  gt() |>
  fmt_number(columns = Value, decimals = 2) |>
  tab_header(title = "RI-CLPM at 9:1 Between:Within Ratio",
             subtitle = paste0(nrow(bayes_summary), " replications, n = ", n_obs))
```

