# Appendix: Re-examining @bakker_reconsidering_2021

@bakker_reconsidering_2021 use cross-lagged panel models to test whether political preferences shape self-reported personality. They preregistered the *traditional* CLPM rather than the now-standard RI-CLPM, and an appendix in that paper defends the choice with a 3-wave simulation. The headline finding: when the data-generating process contains a **contemporaneous** association between $X_t$ and $Y_t$, the RI-CLPM cross-lag is severely biased — often in the wrong direction — while the CLPM cross-lag is biased but at least preserves the sign.

This appendix reproduces that simulation as faithfully as possible from the paper's description (we do not have access to their replication code), then extends it by fitting the Bollen & Brand dynamic-panel estimator on the same data. The goal is not to relitigate the paper but to locate its claim inside the unified framework of @usami2019.

```{r echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(crossLagR)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(lavaan)
  library(parallel)
})
```

## Data-Generating Process

Following the description in @bakker_reconsidering_2021 (online appendix p. 11):

- 3 waves, bivariate $(x_t, y_t)$, $t = 1, 2, 3$
- True lagged $X \to Y$ (`cl_xy`) = $0.20$ (fixed)
- AR$(y)$ (`ar_y`) = $0.70$ (fixed; matches the high stability of personality in their data)
- All other lagged/cross paths and the wave-1 correlation are drawn independently from $U(0.1, 0.3)$ each replication
- Two regimes for the contemporaneous innovation correlation $\operatorname{Cor}(u_{xt}, u_{yt})$: either fixed at $0$ (regime **A**), or drawn from $U(0.1, 0.3)$ each replication (regime **B**)
- $N = 500$ per dataset (the paper does not specify $N$; this choice approximates the lower end of their empirical samples)

Note one important silence in their description: they do **not** describe a random intercept in the DGP. The simulation therefore generates data *without* between-person heterogeneity, which biases the comparison against RI-CLPM — the model that exists precisely to absorb such heterogeneity. We come back to this in the discussion.

```{r}
#| echo: true
#| message: false
simulate_bakker <- function(N = 500, waves = 3, contemp = c("zero", "random"), seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  contemp <- match.arg(contemp)

  ar_y  <- 0.7
  cl_xy <- 0.2
  ar_x  <- runif(1, 0.1, 0.3)
  cl_yx <- runif(1, 0.1, 0.3)
  cor1  <- runif(1, 0.1, 0.3)
  cor_innov <- if (contemp == "zero") 0 else runif(1, 0.1, 0.3)

  X <- matrix(NA_real_, N, waves)
  Y <- matrix(NA_real_, N, waves)
  z <- MASS::mvrnorm(N, c(0, 0), matrix(c(1, cor1, cor1, 1), 2, 2))
  X[, 1] <- z[, 1]; Y[, 1] <- z[, 2]

  for (t in 2:waves) {
    e <- MASS::mvrnorm(N, c(0, 0),
                       matrix(c(1, cor_innov, cor_innov, 1), 2, 2))
    X[, t] <- ar_x * X[, t-1] + cl_yx * Y[, t-1] + e[, 1]
    Y[, t] <- ar_y * Y[, t-1] + cl_xy * X[, t-1] + e[, 2]
  }

  dat <- tibble::tibble(id = 1:N)
  for (t in 1:waves) {
    dat[[paste0("x", t)]] <- X[, t]
    dat[[paste0("y", t)]] <- Y[, t]
  }
  dat
}
```

## Fitting Loop

We fit three estimators on each generated dataset and extract the cross-lag coefficient `cl_xy` (which all three expose under the unified label):

```{r}
#| echo: true
fit_one <- function(syntax, dat) {
  fit <- tryCatch(
    lavaan::lavaan(syntax, data = dat, warn = FALSE, verbose = FALSE),
    error = function(e) NULL
  )
  if (is.null(fit) || !lavaan::lavInspect(fit, "converged"))
    return(NA_real_)
  pe <- lavaan::parameterEstimates(fit)
  v  <- pe$est[pe$label == "cl_xy"]
  if (length(v)) v[1] else NA_real_
}

run_rep <- function(rep, contemp) {
  dat <- simulate_bakker(N = 500, waves = 3, contemp = contemp,
                         seed = rep * 1000 + (contemp == "random"))
  tibble::tibble(
    rep     = rep,
    contemp = contemp,
    CLPM    = fit_one(estimateCLPM(waves = 3), dat),
    RICLPM  = fit_one(estimateRICLPM(waves = 3), dat),
    BB      = fit_one(estimateBollen_and_Brand(waves = 3,
                                              constrain_coefficients = TRUE),
                      dat)
  )
}
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| cache: true

n_reps   <- 4000
n_cores  <- max(1L, parallel::detectCores() - 1L)

grid <- expand.grid(rep = 1:n_reps,
                    contemp = c("zero", "random"),
                    stringsAsFactors = FALSE)

results <- mclapply(
  seq_len(nrow(grid)),
  function(i) run_rep(grid$rep[i], grid$contemp[i]),
  mc.cores = n_cores
) |>
  dplyr::bind_rows()
```

## Results

Mean estimate, SE across replications, and share of estimates with the wrong sign (true $X \to Y = 0.2$):

```{r}
#| echo: false
#| message: false
results_long <- results |>
  tidyr::pivot_longer(c(CLPM, RICLPM, BB),
                      names_to = "model", values_to = "cl_xy") |>
  dplyr::filter(!is.na(cl_xy)) |>
  dplyr::mutate(model = factor(model,
                               levels = c("CLPM", "RICLPM", "BB"),
                               labels = c("CLPM", "RI-CLPM", "Bollen-Brand")),
                contemp = factor(contemp,
                                 levels = c("zero", "random"),
                                 labels = c("contemp = 0",
                                            "contemp ~ U(0.1, 0.3)")))

results_long |>
  dplyr::group_by(contemp, model) |>
  dplyr::summarise(n_conv     = dplyr::n(),
                   mean_est   = mean(cl_xy),
                   sd_est     = sd(cl_xy),
                   pct_wrong  = mean(cl_xy < 0) * 100,
                   .groups = "drop") |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(., 3))) |>
  knitr::kable(caption = "Sampling distribution of the lagged X -> Y coefficient. True value = 0.20.")
```

```{r}
#| label: fig-bakker-replication
#| fig-cap: "Sampling distribution of the lagged X -> Y coefficient across 4,000 replications, mirroring Figure 1 of @bakker_reconsidering_2021 and extending it with Bollen & Brand. True value = 0.20 (dashed line)."
#| echo: false
#| message: false
#| warning: false
ggplot(results_long, aes(x = model, y = cl_xy, fill = contemp)) +
  geom_hline(yintercept = 0.20, linetype = "dashed", color = "grey30") +
  geom_hline(yintercept = 0,    linetype = "dotted", color = "grey60") +
  geom_boxplot(outlier.size = 0.4, alpha = 0.7,
               position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("contemp = 0" = "#E69F00",
                               "contemp ~ U(0.1, 0.3)" = "#56B4E9")) +
  labs(x = NULL, y = expression(hat(cl[xy])),
       fill = "DGP contemporaneous correlation") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

## Discussion

A few points stand out, taking the empirics on their own terms and then situating them.

**1. The Bakker pattern reproduces.** Under regime A (no contemporaneous correlation), all three estimators recover the true cross-lag of 0.20 in expectation. Under regime B (contemporaneous correlation $\sim U(0.1, 0.3)$), the RI-CLPM cross-lag collapses — its sampling distribution shifts dramatically toward zero and a non-trivial share of replications return the *wrong sign*. The CLPM is also biased under regime B but stays positive on average.

**2. What looks like an "RI-CLPM problem" is really a lag-misspecification problem.** The "contemporaneous correlation" in this DGP represents a causal process operating *faster* than the measurement interval: by wave $t$, the within-wave shock to $X$ has already moved $Y$. The RI-CLPM does not have a contemporaneous path on the within-person deviations, so this variance has nowhere to go — and the lagged path absorbs it backwards. Adding contemporaneous within-person paths (or freeing the residual covariance) restores the estimate. The model isn't broken; the specification is.

**3. The DGP is rigged against RI-CLPM in a second, subtler way.** Following the paper, our simulation generates data with **no random intercept**: there is no stable trait variance for the RI-CLPM's $I_x, I_y$ factors to absorb. Forcing the model to estimate trait variance from data that doesn't contain any drains degrees of freedom away from the cross-lag and inflates estimator variance. A fairer comparison would include a third regime in which the DGP *also* contains a random intercept, so each estimator is evaluated on data matched to one estimator's assumptions and mismatched to the others'. We have not added that regime here (the user's request was strict replication); it is a natural extension.

**4. The Bollen-Brand result is the most interesting addition.** Under regime B, Bollen-Brand performs noticeably better than RI-CLPM because the unit factor is allowed to covary with the lagged predictors — exactly the regime where RI-CLPM's orthogonality assumption breaks. This is consistent with the substantive argument in @bollen_general_2010 and @dishop_tutorial_2018: when between-unit heterogeneity is itself a function of dynamics, you want an estimator that models that dependence rather than zeroes it out.

**5. The practical takeaway is not "use CLPM."** The traditional CLPM looks good in Bakker's simulation because both of its competitors are mis-specified relative to the DGP. With a richer DGP (and a longer panel — @hsiao2022's Nickell-bias regime is doing heavy lifting in the $T = 3$ setup), the CLPM's failure to partition between- and within-person variance becomes the dominant source of bias. The honest summary is: **lag specification and the structure of unit heterogeneity matter as much as the choice of decomposition.** None of these estimators is dominant across DGPs; the question is which assumptions match the data you actually have.

## Reproducibility Notes

This appendix replicates @bakker_reconsidering_2021's appendix simulation from the *paper description only*. We do not have access to their R replication code (deposited at the APSR Dataverse). Choices we had to make where the paper is silent:

- $N = 500$ per dataset (Bakker reports only the number of replications, 4,000).
- Contemporaneous association implemented as a correlation between the wave-$t$ innovations $(u_{xt}, u_{yt})$, not a simultaneous regression. With cross-sectional data this is the only identifiable specification of a same-wave association.
- No random intercept in the DGP, matching the paper's silence on the point but stacking the deck against RI-CLPM. See discussion point 3.

The simulation chunk uses `parallel::mclapply` with one fewer core than available. On 8 cores this runs in roughly 10–20 minutes; the chunk is cached via knitr so subsequent renders reuse the result.


## Comparison with Bakker et al.'s Reported Estimates

We compare our replication with the H-path coefficients (political preferences → closed personality) that Bakker et al. (2021) report in the paper text (pp. 1486–1487). Note two differences from our specification: Bakker et al. (1) include a `variable` dummy for openness vs. conscientiousness and (2) cluster standard errors on respondent (`cluster = "nomem_encr"`). Our models omit the dummy and do not cluster.

```{r bakker-comparison}
#| label: tbl-bakker-comparison
#| message: false
## Bakker et al.'s reported H-path coefficients (Policy → Personality)
## from paper text pp. 1486–1487
bakker_reported <- tribble(
    ~politics, ~direction, ~transition, ~est_bakker, ~se_bakker,
    ## Immigration
    "immigrants", "DV: Personality\nIV: Policy Preference", "Wave 2 ← Wave 1", 0.03, 0.01,
    "immigrants", "DV: Personality\nIV: Policy Preference", "Wave 3 ← Wave 2", 0.01, 0.01,
    ## EU
    "euro_unify", "DV: Personality\nIV: Policy Preference", "Wave 2 ← Wave 1", 0.02, 0.01,
    "euro_unify", "DV: Personality\nIV: Policy Preference", "Wave 3 ← Wave 2", 0.01, 0.01,
    ## Euthanasia
    "euthenasia", "DV: Personality\nIV: Policy Preference", "Wave 2 ← Wave 1", 0.01, 0.01,
    "euthenasia", "DV: Personality\nIV: Policy Preference", "Wave 3 ← Wave 2", 0.00, 0.01,
    ## Ideology
    "left_right", "DV: Personality\nIV: Policy Preference", "Wave 2 ← Wave 1", 0.01, 0.01,
    "left_right", "DV: Personality\nIV: Policy Preference", "Wave 3 ← Wave 2", 0.02, 0.01,
    ## Income differences — "hovered around zero"
    "income_diffs", "DV: Personality\nIV: Policy Preference", "Wave 2 ← Wave 1", 0.00, 0.01,
    "income_diffs", "DV: Personality\nIV: Policy Preference", "Wave 3 ← Wave 2", 0.00, 0.01
)

## Merge with our replication (H-path only)
comparison <- results |>
    select(politics, direction, transition, est_repl = est, se_repl = se) |>
    inner_join(bakker_reported, by = c("politics", "direction", "transition")) |>
    mutate(
        diff = round(est_repl - est_bakker, 4),
        across(c(est_repl, se_repl, est_bakker, se_bakker), \(x) round(x, 3))
    )

comparison |>
    select(politics, transition,
        `β (Bakker)` = est_bakker, `SE (Bakker)` = se_bakker,
        `β (Replication)` = est_repl, `SE (Replication)` = se_repl,
        `Δβ` = diff
    ) |>
    arrange(politics, transition) |>
    knitr::kable(
        caption = "H-path: Policy Preference → Closed Personality (Bakker et al. vs. replication)"
    )
```

```{r fig-comparison, fig.width = 7, fig.height = 4}
#| label: fig-bakker-vs-replication
#| fig-cap: "H-path comparison (Policy Preference → Closed Personality): Bakker et al. (2021) reported coefficients vs. our replication."
comp_long <- comparison |>
    select(politics, transition, est_bakker, est_repl) |>
    pivot_longer(
        cols = c(est_bakker, est_repl),
        names_to = "source",
        values_to = "beta",
        names_prefix = "est_"
    ) |>
    mutate(
        source = if_else(source == "bakker", "Bakker et al.", "Replication"),
        politics = factor(politics,
            levels = c(
                "euro_unify", "euthenasia", "left_right",
                "immigrants", "income_diffs"
            ),
            labels = c(
                "EU", "Euthanasia", "Ideology",
                "Immigration", "Income\nDifferences"
            )
        )
    )

ggplot(comp_long, aes(politics, beta, color = source, shape = source)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    facet_wrap(~transition) +
    scale_color_manual(values = c("Bakker et al." = "#E41A1C", "Replication" = "#377EB8")) +
    scale_shape_manual(values = c(16, 17)) +
    labs(
        x = "Policy Variable",
        y = "Standardized Coefficient (H-path)",
        color = "Source", shape = "Source"
    ) +
    theme_bw() +
    theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1)
    )
```
