# Package Reference {#sec-package-reference}

This appendix is a compact tour of every user-facing function in `crossLagR`. Each estimator section shows the generated lavaan syntax, fits the model to simulated data, and prints a `crossLagRSummary()` so you can see what an actual call returns. The final section documents the four S3 output classes the package exposes.

```{r}
#| label: setup-appendix
#| message: false
#| warning: false
#| echo: false
suppressPackageStartupMessages({
  library(crossLagR)
  library(lavaan)
})
devtools::load_all(quiet = TRUE)
set.seed(1234)
```

## Quick start

The intended user workflow is three lines: build syntax, fit + diagnose, inspect.

```{r}
#| eval: false
syntax <- estimateCLPM(waves = 4)
out    <- lavaanDiagnose(syntax, data = my_data, icc_vars = c("x", "y"))
out                                         # full bundled report
```

`lavaanDiagnose()` returns a `crossLagR_fit` object that bundles the lavaan fit, a labelled parameter summary, a Heywood-case diagnosis, and (optionally) an ICC report.

## Estimators

Every estimator function returns a character string of lavaan syntax. Parameter labels follow the unified scheme from @usami2019: `ar_x, ar_y, cl_xy, cl_yx` on the dynamic equation, `d_var_x, d_var_y, d_cov_xy` on dynamic residuals, plus model-specific labels for random intercepts (`I_x`), growth factors (`S_x, Q_x`), trait/state decompositions (`T_var_x, s_var_x`), etc. Run `unifiedLabels()` for the full catalog with plain-language descriptions.

### CLPM --- `estimateCLPM()`

Standard cross-lagged panel model. All observed variance is treated as within-person; no between-person random intercept.

```{r}
#| label: ref-clpm
estimateCLPM(waves = 4) |> cat()
```

```{r}
#| label: ref-clpm-fit
dat_clpm <- simCLPM(waves = 4, sample_size = 500, seed = 1)$data
fit_clpm <- lavaan(estimateCLPM(waves = 4), data = dat_clpm,
                   meanstructure = TRUE)
crossLagRSummary(fit_clpm)
```

### RI-CLPM --- `estimateRICLPM()`

Random-Intercept CLPM [@hamaker_critique_2015]. Decomposes each observation into a stable trait (`I_x, I_y`) plus a within-person deviation, with the dynamic equation on the deviations.

```{r}
#| label: ref-riclpm
estimateRICLPM(waves = 4) |> cat()
```

```{r}
#| label: ref-riclpm-fit
#| warning: false
dat_riclpm <- simRICLPM(waves = 4, sample.nobs = 500)$data
fit_riclpm <- lavaan(estimateRICLPM(waves = 4), data = dat_riclpm,
                     meanstructure = TRUE)
crossLagRSummary(fit_riclpm)
```

### ALT --- `estimateALT()`

Autoregressive Latent Trajectory model. AR/CL paths run on the observed scores directly, on top of an intercept (`I_x`) and slope (`S_x`) growth factor.

```{r}
#| label: ref-alt
estimateALT(waves = 4) |> cat()
```

### LCM-SR --- `estimateLCMSR()`

Latent Curve Model with Structured Residuals. Linear growth (`I, S`) absorbs the trajectory, and AR/CL paths are estimated on the growth residuals.

```{r}
#| label: ref-lcmsr
estimateLCMSR(waves = 4) |> cat()
```

### LGM --- `estimateLGM()`

Linear (or quadratic) Latent Growth Model. Univariate by default; pass `variable_type = "bivariate"` for a parallel-process LGM.

```{r}
#| label: ref-lgm
estimateLGM(waves = 4, variable_type = "bivariate") |> cat()
```

### Latent Change --- `estimateLChange()`

Latent change score (McArdle) parameterization. Set `estimate_constant_change = TRUE` to add an accumulating-change factor (`A_x`).

```{r}
#| label: ref-lchange
estimateLChange(waves = 4, variable_type = "bivariate") |> cat()
```

### Bollen & Brand --- `estimateBollen_and_Brand()`

Dynamic-panel SEM with a latent unit-effect that loads on every wave of the *observed* CLPM equation, with the lagged predictor allowed to covary with that effect [@bollen_general_2010]. SEM analogue of Arellano-Bond.

```{r}
#| label: ref-bb
estimateBollen_and_Brand(waves = 4,
                        x_effect         = "lagged",
                        x_autoregression = TRUE,
                        y_effect_on_x    = TRUE) |> cat()
```

### TSO --- `estimateTSO()`

Bivariate Trait-State-Occasion variance decomposition (random-intercept-only special case of the RI-CLPM with no lagged dynamics). Use this when the question is "how much of the variance is between vs within."

```{r}
#| label: ref-tso
estimateTSO(waves = 4) |> cat()
```

## Simulation

### `simCLPM()` / `simRICLPM()` / `simCLPMu()`

Direct simulators for specific data-generating processes. Each returns a `list(model, data)`; `simCLPMu()` adds a time-varying unmeasured confounder for sensitivity studies.

```{r}
#| label: ref-sim-direct
sim <- simCLPM(waves = 4, beta_x = 0.4, beta_y = 0.4,
               omega_xy = 0.15, omega_yx = 0.10,
               sample_size = 200, seed = 1)
head(sim$data, 3)
```

### `simFromSyntax()` --- generic

The general path. Pass an estimator name and concrete numeric values for the unified labels; the function builds that estimator's lavaan syntax, substitutes the numeric values via `populate_unified_labels()`, and calls `lavaan::simulateData()`. Works for any of the unified-label estimators.

```{r}
#| label: ref-sim-syntax
dat <- simFromSyntax(estimator = "RICLPM", waves = 4, sample_size = 300,
                     ar_x = 0.3, ar_y = 0.3, cl_xy = 0.15, cl_yx = 0.10,
                     d_var_x = 0.5, d_var_y = 0.5, seed = 2)
dim(dat); head(names(dat))
```

## Monte Carlo

### `monteCarloLavaan()`

Generic Monte Carlo wrapper: pick an estimator, pick a DGP, supply a parameter grid, get one row per trial with unified-label estimates, SEs, fit indices, n_obs / n_miss (per @hamaker_critique_2015 reporting norms), Heywood diagnostics, and ICCs.

```{r}
#| label: ref-mc
#| warning: false
grid <- data.frame(
  stability_p = 0.3, stability_q = 0.3,
  cross_p = 0.10, cross_q = 0.15,
  variance_p = 0.5, variance_q = 0.5,
  variance_between_x = 0.5, variance_between_y = 0.5,
  cov_pq = 0
)
mc <- monteCarloLavaan(
  estimator = "RICLPM", dgp = "CLPM",
  trials = 3, waves = 4, sample_size = 300,
  param_grid = grid
)
mc[, c("trial", "ar_x", "cl_xy", "cl_yx", "cfi", "rmsea",
       "n_obs", "heywood_n_issues", "icc_x", "icc_y")]
```

`run_mc_sims()` is a convenience wrapper around `monteCarloLavaan()` that takes a `data_generation` argument in place of `dgp` for backwards compatibility with older scripts.

## Diagnostics

The diagnostic system answers three questions: *did the fit succeed*, *do the estimates make sense*, and *does the data warrant this model*.

### `lavaanDiagnose()` --- bundled fit

One-call wrapper. Fits the lavaan model, captures warnings that lavaan otherwise emits in places hard to intercept (NPD covariance, vcov singular), runs `crossLagRSummary()`, `diagnoseFit()`, and optionally `iccReport()`. Returns everything bundled in a `crossLagR_fit` object.

```{r}
#| label: ref-diagnose-bundle
#| warning: false
out <- lavaanDiagnose(
  estimateCLPM(waves = 4),
  data          = dat_clpm,
  meanstructure = TRUE,
  icc_vars      = c("x", "y")
)
names(out)
```

### `crossLagRSummary()` --- labelled estimates + fit

```{r}
#| label: ref-summary
out$summary
```

The parameter table joins each estimated unified label to a plain-language description, attaches CI / sig-stars, and reports standardized estimates. The fit block prints chi-square, CFI/TLI, RMSEA + 90% CI, SRMR, AIC, BIC.

### `diagnoseFit()` --- Heywood detection

Scans the lavaan fit for boundary conditions (negative variances, standardized loadings > 1, latent correlations > 1, NPD covariances, vcov singular, non-convergence, iteration-limit hits) and pairs each with an explanation and concrete remedies.

```{r}
#| label: ref-diagnose-fit
out$diagnosis
```

Call `heywoodHelp()` without a fit to browse the full catalog of detectable codes; `heywoodHelp("negative_variance")` prints just one entry.

### `iccReport()` --- between vs within

Computes the empirical ICC for each construct directly from data --- no model fit required. When any ICC exceeds the threshold (default 0.65), a single warning fires recommending sensitivity analysis with the bundled Shiny simulator (`launch_shiny_sim()`), because CLPM vs RI-CLPM estimates can diverge substantially in high-ICC regimes [@hamaker_critique_2015].

```{r}
#| label: ref-icc
out$icc
```

`withinBetween()` is the underlying decomposition; it accepts long-format data and supports both an empirical (`method = "dplyr"`) and a TSO-based (`method = "tso"`) split.

### `unifiedLabels()` --- label catalog

Browse every unified label and its description, grouped by `dynamics / dynamic_residuals / random_intercepts / growth / trait_state / residuals / change`.

```{r}
#| label: ref-unified
unifiedLabels(print = FALSE) |> head(10)
```

### Shiny simulator --- `launch_shiny_sim()`

Interactive simulator for visualising bias and recovery across estimator / DGP combinations. Useful when an ICC warning fires and you want to see how the choice of model bites on your data:

```{r}
#| eval: false
launch_shiny_sim()
```

## Output object reference {#sec-output-objects}

The package returns S3 objects so users get pretty-printed summaries by default while retaining access to the raw fields.

### `crossLagR_fit`

Returned by `lavaanDiagnose()`. A list with class `crossLagR_fit` and elements:

| Field | Class | Contents |
|---|---|---|
| `$fit` | `lavaan` (S4) | The raw lavaan fit. Use for anything not exposed in the summary. |
| `$summary` | `crossLagR_summary` | Labelled parameter table + fit indices. |
| `$diagnosis` | `crossLagR_diagnosis` | Heywood-case detection result. |
| `$icc` | `crossLagR_icc` or NULL | ICC report (only if `icc_vars` was supplied). |
| `$warnings` | character | All lavaan warning messages captured during the fit. |

`print()` and `summary()` walk all three diagnostic sub-objects in order.

### `crossLagR_summary`

Returned by `crossLagRSummary()` (and stored at `out$summary`).

| Field | Class | Contents |
|---|---|---|
| `$parameters` | data.frame | One row per recognized unified label: `label, description, est, se, z, p, ci.lower, ci.upper, std.all`. |
| `$other` | data.frame | Estimated parameters whose label was *not* in the unified catalog (`label, est, se, z, p`). |
| `$fit` | named numeric | `chisq, df, pvalue, cfi, tli, rmsea, rmsea.ci.lower, rmsea.ci.upper, srmr, aic, bic`. |
| `$n_obs`, `$n_miss` | integer | Sample-size bookkeeping (always reported alongside estimates). |
| `$converged` | logical | From `lavInspect(fit, "converged")`. |
| `$ci_level`, `$digits` | numeric / int | Stored for the print method. |

### `crossLagR_diagnosis`

Returned by `diagnoseFit()` (and stored at `out$diagnosis`).

| Field | Class | Contents |
|---|---|---|
| `$issues` | list | One entry per detected issue: `code, severity, parameters, message, explanation, remedies`. |
| `$n_issues` | integer | Number of detected issues. |
| `$codes` | character | Detected issue codes (also semicolon-joined in MC output's `heywood_codes` column). |
| `$summary` | character | One-line summary string. |

Detected codes:

| Code | Severity | Trigger |
|---|---|---|
| `non_convergence` | high | `lavInspect(fit, "converged")` is `FALSE` |
| `negative_variance` | high | any `~~` variance row with `est < tol_negative` |
| `std_loading_gt_one` | medium | `=~` row with `|std.all| > 1` |
| `std_correlation_gt_one` | high | latent `~~` row with `|std.all| > 1` |
| `npd_covariance` | high | warning text matches non-positive-definite |
| `vcov_not_pd` | medium | warning text matches vcov singular |
| `iter_limit` | high | warning text matches iteration limit reached |

### `crossLagR_icc`

Returned by `iccReport()` (and stored at `out$icc`). A data.frame with class `crossLagR_icc`, one row per construct, columns:

| Column | Meaning |
|---|---|
| `variable` | Construct name (stem or column). |
| `between_var` | Var of person means. |
| `within_var` | Var of within-person deviations. |
| `total_var` | Sum of the two. |
| `icc` | `between_var / total_var`. |
| `n_obs` | Complete cases used. |
| `n_waves` | Number of wave columns matched. |
| `exceeds_threshold` | TRUE if `icc > threshold` (default 0.65). |

The threshold is stored as `attr(x, "threshold")`. The print method repeats the Shiny-sensitivity recommendation for any flagged variable.

### Monte Carlo output

`monteCarloLavaan()` returns a plain `data.frame` (one row per trial) with the following column groups:

| Group | Columns |
|---|---|
| Bookkeeping | `trial, param_combo, estimator, dgp` |
| Estimates | `ar_x, ar_y, cl_xy, cl_yx` (+ `se_*`, `std_*`) |
| Truth | `true_ar_x, true_ar_y, true_cl_xy, true_cl_yx` |
| Fit | `chisq, df, pvalue, cfi, tli, rmsea, srmr, aic, bic` |
| Sample | `n_obs, n_miss, converged` |
| Warnings | `warning_count, warn_npd, warn_heywood, warn_iter` |
| Heywood diag. | `heywood_n_issues, heywood_codes` |
| ICC | `icc_x, icc_y, icc_exceeds` |
| Errors | `error_occurred, error_message` |
