# Simulation and Estimation

The $\texttt{crossLagR}$ package includes several functions to simulate data.

$\texttt{simRICLPM()}$ simulates data from a random intercept cross-lagged panel model.

$\texttt{simCLPM()}$ simulates data from a cross-lagged panel model.

$\texttt{reshape\_long\_sim\_cr()}$ Takes a wide data frame, pivots to wide format, includes lags.


```{r, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE }
# devtools::load_all()
library(dplyr)

df = simCLPM(waves = 5,
             variance_p = 0.5,      # underscore
             variance_q = 0.5,      # underscore
             stability_p = 0.9,      # underscore
             stability_q = 0.9,      # underscore
             cross_q = 0,
             cross_p = 0,
             cov_pq = 0.00,
             means_p = c(1,1.1,1.2,1.3,1.4),
             sample.nobs = 1000)$data
  df |>
  mutate(id = seq(1:nrow(df)))  |>
  reshape_long_sim_cr() -> dat_long
head(dat_long)
lm(y ~  ylag + xlag, data = dat_long)
```
### Trajectories

```{r}
library(dplyr)
library(ggplot2)
dat_long |> 
  ggplot(aes(x = wave, y = y, group = id)) +
  geom_line(alpha = 0.1, color = "black", position = position_jitter(width = 0.1, height = 0.2)) +
  # geom_point(alpha = 0.3, size = 0.5, position = position_jitter(width = 0.1, height = 0.2)) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "lightblue", size = 1, alpha = 0.8) +
  stat_summary(aes(group = 1), fun = mean, geom = "point", 
               color = "lightblue", size = 1) +
  labs(
    title = "Individual Trajectories: Simulated Data",
    subtitle = "Linear Trend",
    x = "Survey Wave",
    y = "Response"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(-5, 6) ) +
  theme(strip.text = element_text(color = "black")) -> plot
print(plot)
```

```{r}
dat_long <- dat_long |>
  mutate(
    # xlag = if_else(is.na(xlag), mean(x, na.rm = TRUE), xlag),
    # ylag = if_else(is.na(ylag), mean(y, na.rm = TRUE), ylag),
    predicted = predict(lm(y ~ ylag + xlag, data = dat_long), newdata = dat_long)
  )
dat_long |> 
  ggplot(aes(x = wave, y = y, group = id)) +
  geom_line(alpha = 0.1, color = "black", position = position_jitter(width = 0.1, height = 0.2)) +
  # Observed mean trend
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "lightblue", size = 1, alpha = 0.8) +
  stat_summary(aes(group = 1), fun = mean, geom = "point", 
               color = "lightblue", size = 1) +
  # Model predictions mean trend
  stat_summary(aes(y = predicted, group = 1), fun = mean, geom = "line", 
               color = "red", size = 1, alpha = 0.8, na.rm = TRUE) +
  stat_summary(aes(y = predicted, group = 1), fun = mean, geom = "point", 
               color = "red", size = 1, na.rm = TRUE) +
  labs(
    title = "Individual Trajectories: Simulated Data",
    subtitle = "Observed (blue) vs CLPM Predicted (red) trends",
    x = "Survey Wave",
    y = "Response"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(-5, 6)) +
  theme(strip.text = element_text(color = "black"))


```
### Cross-Lagged Panel Model

Let's fit several models under a random intercept cross lagged panel data generating process.

```{r}
data = simCLPM(waves = 5,
             variance_p = 0.5,      # underscore
             variance_q = 0.5,      # underscore
             stability_p = 0.9,      # underscore
             stability_q = 0.9,      # underscore
             cross_q = 0,
             cross_p = 0,
             cov_pq = 0.00,
             # means_p = c(1,1.1,1.2,1.3,1.4),
             sample.nobs = 1000)$data

lavaan::lavaan(estimateLChange(waves = 3,
                               constrain_omega = TRUE,
                               constrain_beta =  TRUE),
              data = data,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml") |>
  lavaan::summary()

```

```{r}
library(crossLagR)
library(lavaan)
# Simulate Data Under RI CLPM
df = simRICLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                variance.between.x = 0.5,
                variance.between.y = 0.5,
                cov.between = 0,
                sample.nobs = 10e2)
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long

fit1 <- lavaan(estimateCLPM(waves = 5),
              data = dat,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml")

summary(fit1)
```

### Random Intercept Cross-Lagged Panel Model

```{r}
devtools::load_all()
library(lavaan)
WAVES = 5

fit2 <- lavaan(estimateRICLPM(data= dat,
                              waves = WAVES, 
                              time_varying_x = paste0("x", 1:WAVES), 
                              time_varying_y = paste0("y", 1:WAVES)),
              data = dat,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml")

summary(fit2)
```

The CLPM clearly does not fit the data well, and provides rather inaccurate estimates.

### Hiearchical Linear Model, Random Intercepts and Slopes

```{r, echo = FALSE, message = FALSE, warning = FALSE}
model = estimateHLM(data = dat_long, residual_correlation = FALSE)
summary(model)
```

### Continuous Time Model

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dat_long %>%
  select(id, wave, x, y) -> ct_sem_data
# ctsem package Follows this convention:
names(ct_sem_data) = c("id", "time", "x", "y")
model = estimateCTSEM(data = ct_sem_data)
# [1] Returns a list; Extract cstem model
# [2] Return a discrete time CL and AR model
summary(model[[1]])$parmatrices
print(model[[2]] %>% t())
```
