--- 
title: "Cross Lagged Regression"
author: "Christopher Weber"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]  # ← First one
description: |
  Cross Lagged Regression
link-citations: yes
github-repo: crweber9874/advancedRegression
---
--- 
title: "Cross Lagged Regression"
author: "Christopher Weber"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]  # ← First one
description: |
  Cross Lagged Regression
link-citations: yes
github-repo: crweber9874/advancedRegression
---


# Preface {-}

The Cross-Lagged Panel Model (CLPM) is widely accepted as a standard method for longitudinal causal analysis in political science. Its application, however, is often susceptible to unaccounted time-invariant confounders and measurement error, leading to spurious estimates of autoregressive and cross-lagged relationships, as well as bias in statistical inferences and substantive conclusions. In these documentation files, I demonstrate the similarities and differences between different panel analysis models, showing in particular, that the standard CLPM is strongly related to a more flexible and robust alternatives, the change score regression model and the random-intercept cross-lagged regression model. Both the Latent Change Model (LCM) and the Random Intercept Cross-Lagged Panel Model (RI-CLPM) extend the standard CLPM by explicitly modeling "unit level" effects. In simulations and empirical examples, we show the myriad paths by which the CLPM produces biased parameter estimates due to unmodeled, stable-unit effects. Both approaches allow for "constant" differences between units over time: In the RI-CLPM, this is represented by a factor corresponding to unit-mean variation in the observed variables, a "random-intercept" while in the LCM, this is captured through a factor corresponding to unit-level change over time, a "random slope."

This documentation elaborates upon:

* The theoretical background for the CLPM, RI-CLPM, and LCM, highlighting their similarities and differences.
* An introduction to the `crossLagR` package, which includes various utilities to estimate and simulate effects from these models.
* A Monte Carlo simulation procedure that examines the degree of bias in parameter estimates from the CLPM, RI-CLPM, and LCM under different data-generating processes.



<!--chapter:end:index.Rmd-->


# Panel Data Models

Placeholder


## The Cross-Lagged Panel Model 
## The Change Score Regression
## Algebraic Equivalence
## Decomposing Change
## Bivariate Dual Change 
## The Random Intercept Cross Lagged Regression 
## Spurious and Causal Effects
## Confounding Variables

<!--chapter:end:01_Panel_Models.Rmd-->

# Simulation and Estimation

The $\texttt{crossLagR}$ package includes several functions to simulate data.

$\texttt{simRICLPM()}$ simulates data from a random intercept cross-lagged panel model.

$\texttt{simCLPM()}$ simulates data from a cross-lagged panel model.

$\texttt{reshape\_long\_sim\_cr()}$ Takes a wide data frame, pivots to wide format, includes lags.


```{r, echo = TRUE, message = FALSE, warning = FALSE}
devtools::load_all()
library(dplyr)


df = simRICLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                variance.between.x = 0.5,
                variance.between.y = 0.5,
                cov.between = 0,
                sample.nobs = 10e2,
                time_varying_y = paste0("y", 1:5),
                time_varying_x = paste0("x", 1:5))
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long
head(dat_long)
```

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(crossLagR)
df = simCLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                sample.nobs = 10e2)
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long
head(dat_long)
```

## Models

$\texttt{estCLPM()}$ Constructs the lavaan code to estimate a CLPM. $\texttt{estRICLPM()}$ Constructs the lavaan code to estimate a RI-CLPM. $\texttt{estCTSEM()}$ Estimates a continuous time structural equation model. $\texttt{estHLM()}$ Estimates the HLM.

### Cross-Lagged Panel Model

Let's fit several models under a random intercept cross lagged panel data generating process.

```{r}
library(crossLagR)
library(lavaan)
# Simulate Data Under RI CLPM
df = simRICLPM(waves = 5,
                variance.p = 0.5,
                variance.q = 0.5,
                stability.p = 0.20,
                stability.q = 0.20,
                cross.q = 0.30,
                cross.p = 0.30,
                cov.pq = 0.00,
                variance.between.x = 0.5,
                variance.between.y = 0.5,
                cov.between = 0,
                sample.nobs = 10e2)
df[[2]] %>%
  mutate(id = seq(1:nrow(df[[2]])))  -> dat
dat %>%
  reshape_long_sim_cr() -> dat_long

fit1 <- lavaan(estimateCLPM(waves = 5),
              data = dat,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml")

summary(fit1)
```

### Random Intercept Cross-Lagged Panel Model

```{r}
devtools::load_all()
library(lavaan)
WAVES = 5

fit2 <- lavaan(estimateRICLPM(data= dat,
                              waves = WAVES, 
                              time_varying_x = paste0("x", 1:WAVES), 
                              time_varying_y = paste0("y", 1:WAVES)),
              data = dat,
              int.ov.free = TRUE, # Fix intercepts to zero
              estimator="ml")

summary(fit2)
```

The CLPM clearly does not fit the data well, and provides rather inaccurate estimates.

### Hiearchical Linear Model, Random Intercepts and Slopes

```{r, echo = FALSE, message = FALSE, warning = FALSE}
model = estimateHLM(data = dat_long, residual_correlation = FALSE)
summary(model)
```

### Continuous Time Model

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dat_long %>%
  select(id, wave, x, y) -> ct_sem_data
# ctsem package Follows this convention:
names(ct_sem_data) = c("id", "time", "x", "y")
model = estimateCTSEM(data = ct_sem_data)
# [1] Returns a list; Extract cstem model
# [2] Return a discrete time CL and AR model
summary(model[[1]])$parmatrices
print(model[[2]] %>% t())
```

<!--chapter:end:02_CLPM_Estimation.Rmd-->

